---
title: "BST02: Using R for Statistics\\ in Medical Research"
subtitle: "Part B: Basic use of **R**"
author: "Eleni-Rosalina Andrinopoulou"
institute: "Department of Biostatistics, Erasmus Medical Center"
date: "24 - 28 February 2020"
email: "e.andrinopoulou@erasmusmc.nl"
output:
  beamer_presentation: 
    template: mytemplate.latex
    includes:
      in_header: SlideTemplate.tex
    keep_tex: yes
    incremental: false
classoption: aspectratio=169
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, comment=NA)

library(survival)
library(knitr)
library(kableExtra)
library(dplyr)
```

## In this Section

* Using R

* In practice examples

* Basics in R

* Common R objects

* Importing data and saving your work

*  Data transformation

* Data exploration

* Visualization of data

* Indexing

Lots of practising

## Using R

* R is a command-based procedural language
  - write and execute commands
  - use and define functions
  
* You may write the commands in the R console (Windows) or in a shell (Linux)

\vspace*{3ex}

\blue{You will become more familiar with the syntax as you use it}


## Using R (cont'd)

* Strongly advisable to use a suitable text editor - Some available options:
  - RWinEdt (for Windows; you also need WinEdt installed)
  - Tinn-R (for Windows; http://sciviews.org/Tinn-R/)
  - Rkward (for Linux)
  - Emacs (w. ESS, all platforms)
  - Visual Studio (for Windows)
  - Rstudio (all major platforms; http://www.rstudio.org/) 
  - for more check https://r-dir.com/blog/2013/01/list-of-r-editors.html
  
  
##  Using R (cont'd)

* For this course: Rstudio (http://www.rstudio.org/)
  - free
  - works fine in Windows, MacOS and Linux
  - helpful with errors
  
  
##  Using R (cont'd)

* Can I use R without Rstudio?

\vspace*{0.5cm}

* Can I use Rstudio without R?


## Practical Examples

* **Package \textcolor{blue}{survival} - pbc data set**

```{r example data, include=TRUE}
dt <- as.data.frame(pbc[1:5,c(1:6,11,12)])
kable(dt, "latex", booktabs = T, align = "c") %>%
 kable_styling(full_width = F) 
```


## Practical Examples (cont'd)

* \textbf{\textcolor{red}{id}}:	case number
* \textbf{\textcolor{red}{time}}:	number of days between registration and the earlier of death, transplantation, or study analysis in July, 1986
* \textbf{\textcolor{red}{status}}:	status at endpoint, 0/1/2 for censored, transplant, dead
* \textbf{\textcolor{red}{trt}}:	1/2/NA for D-penicillmain, placebo, not randomised
* \textbf{\textcolor{red}{age}}:	in years
* \textbf{\textcolor{red}{sex}}:	m/f
* \textbf{\textcolor{red}{bili}}:	serum bilirunbin (mg/dl)
* \textbf{\textcolor{red}{chol}}:	serum cholesterol (mg/dl)

More details: https://stat.ethz.ch/R-manual/R-devel/library/survival/html/pbc.html


## Practical Examples (cont'd)

* What is a \textcolor{red}{scalar}/vector/matrix

```{r example data b, include=TRUE}
dt <- as.data.frame(pbc[1:5,c(1:6,11,12)])
dt <- format(dt, digits = 7)
dt %>%
 mutate(
   dt = row.names(.),
   age = cell_spec(age, "latex", color = ifelse(age >= 70, "red", "black"), bold = ifelse(age >= 70, "TRUE", "FALSE"), align = c("c", rep("c", 4))),
   time = cell_spec(time, align = c("c", rep("c", 4))),
   id = cell_spec(id, align = c("c", rep("c", 4))),
   sex = cell_spec(sex, align = c("c", rep("c", 4))),
   status = cell_spec(status, align = c("c", rep("c", 4))),
   trt = cell_spec(trt, align = c("c", rep("c", 4))),
   bili = cell_spec(bili, align = c("c", rep("c", 4))),
   chol = cell_spec(chol, align = c("c", rep("c", 4))),
   )%>%
select(id, time, status, trt, age, sex, bili, chol) %>%
kable("latex", escape = F, booktabs = T, align = "c")  %>%
kable_styling(position = "center", full_width = F)
```


## Practical Examples (cont'd)

* What is a scalar/\textcolor{red}{vector}/matrix

```{r example data c, include=TRUE}
dt <- as.data.frame(pbc[1:5,c(1:6,11,12)])
kable(dt, "latex", booktabs = T, align = "c") %>%
 kable_styling(full_width = F) %>%
 column_spec(3, bold = TRUE, color = "red")
```


## Practical Examples (cont'd)

* What is a scalar/\textcolor{red}{vector}/matrix

```{r example data d, include=TRUE}
dt <- as.data.frame(pbc[1:5,c(1:6,11,12)])
kable(dt, "latex", booktabs = T, align = "c") %>%
 kable_styling(full_width = F) %>%
 column_spec(2, bold = TRUE, color = "red")
```


## Practical Examples (cont'd)

* What is a scalar/vector/\textcolor{red}{matrix}

```{r example data e, include=TRUE}
dt <- as.data.frame(pbc[1:5,c(1:6,11,12)])
kable(dt, "latex", booktabs = T, align = "c") %>%
 kable_styling(full_width = F) %>%
 column_spec(2:3, bold = TRUE, color = "red")
```


## Practical Examples (cont'd)

* What is a scalar/vector/\textcolor{red}{matrix}

```{r example data f, include=TRUE}
dt <- as.data.frame(pbc[1:5,c(1:6,11,12)])
kable(dt, "latex", booktabs = T, align = "c") %>%
 kable_styling(full_width = F) %>%
 column_spec(7:8, bold = TRUE, color = "red")
```


## Practical Examples (cont'd)

* Common questions
   - What is the average age?
   - What is the average serum bilirubin?
   - What is the average serum cholesterol?
   - What is the percentage of females?
   - How many missing values do we have for serum cholesterol?
   
   
## Basics in R

* Elementary commands: \textbf{\textcolor{red}{expressions}} and \textbf{\textcolor{blue}{assignments}}

* An \textbf{\textcolor{red}{expression}} given as command is evaluated printed and discarded

* An \textbf{\textcolor{blue}{assignment}} evaluates an expression and passes the value to a variable - the result is not automatically printed


## Basics in R (cont'd)

Expression is given as a command,
```{r, eval = TRUE, include = TRUE, echo = TRUE}
103473
```

\vspace*{3ex}

 - However, it cannot be viewed again unless the command is rerun.
 
\pause
 
\vspace*{3ex}

 In order to store information, the expression should assign the command
 
```{r, eval = TRUE, include = TRUE, echo = TRUE}
x <- 103473
x
```


## Basics in R (cont'd)

* Basic arithmetics
 ```{r, eval = FALSE, include = TRUE, echo = TRUE}
+, -, *, /, ^
```

* Complicated arithmetics


## Basics in R (cont'd)

* R is case sensitive, e.g.,
  - \textbf{\textcolor{blue}{"sex"}} is different than \textbf{\textcolor{blue}{"Sex"}}
* Commands are separated by a semi-colon or by a newline
* Comments can be put anywhere, starting with a hashmark \textbf{\textcolor{blue}{\#}}: everything to the end of the line is a comment
* Assign a value to an object by \textbf{\textcolor{blue}{<-}} or \textbf{\textcolor{blue}{=}}


## Basics in R (cont'd)

* Missing values
  - are coded as \textbf{\textcolor{blue}{NA}} (i.e., not available) \textbf{\textcolor{blue}{is.na()}}
* Infinity
  - is coded as \textbf{\textcolor{blue}{Inf}} (plus infinity) or \textbf{\textcolor{blue}{-Inf}} (minus infinity) \textbf{\textcolor{blue}{is.finite()}}
* Not a number
  - is coded as \textbf{\textcolor{blue}{NaN}} (Not a Number). Example: 
```{r, eval = TRUE, include = TRUE, echo = TRUE}
0/0
```


## Basics in R (cont'd)


\begin{columns}[T,onlytextwidth]
\begin{column}{0.47\linewidth}
\begin{block}{Demos}
\begin{itemize}
\item Basic R \button{01_Basic_R.R}{R} \button{01_Basic_R.html}{html}
\end{itemize}
\end{block}
\end{column}

\begin{column}{0.47\linewidth}
\end{column}
\end{columns}



## Common R Objects

* There are different kinds of variables

\vspace*{3ex}

```{r example data g, include=TRUE}
dt <- as.data.frame(pbc[1:5,c(1:6,11,12)])
kable(dt, "latex", booktabs = T, digits = 2) %>%
 kable_styling(full_width = F) 
```


## Common R Objects (cont'd)

* in R Everything (data, results, ...) is an object

* In order to list the created objects use the following functions
```{r, eval = FALSE, include = TRUE, echo = TRUE}
objects()
ls()
```

\pause

  - In order to remove objects
```{r, eval = FALSE, include = TRUE, echo = TRUE}
rm(...)
rm(list=ls(all=TRUE))
```


## Common R Objects (cont'd)

* To investigate a specific object (e.g. pbc)
```{r, eval = TRUE, include = TRUE, echo = TRUE}
str(pbc[,c("id", "time", "status", "trt", "age", "sex", "bili", "chol")])
```


## Common R Objects (cont'd)

The simplest data structures are:

* \textbf{\textcolor{red}{numeric}} : quantitative data
* \textbf{\textcolor{red}{character}} : qualitative data
* \textbf{\textcolor{red}{integer}} : whole numbers
* \textbf{\textcolor{red}{logical}} : TRUE or FALSE
* \textbf{\textcolor{red}{factors}} : qualitative data (ordered)

To find out what type of object you have, you can use the following function
```{r, eval = TRUE, include = TRUE, echo = TRUE}
mode(pbc$age)
```


## Common R Objects (cont'd)

* Every object in R has a class

  - \textbf{\textcolor{red}{Vectors}} have the same type of elements 
  - \textbf{\textcolor{red}{Matrices}} have the same type of elements 
  - \textbf{\textcolor{red}{Data.frames}} can have elements of different type
  - \textbf{\textcolor{red}{Lists}} can have elements of different length and type


## Common R Objects (cont'd)

* Differences between \textbf{\textcolor{red}{Vectors}}, Matrices, Data.frames and Lists
```{r, eval = TRUE, include = TRUE, echo = TRUE}
pbc[1:6, c("age")]
```

## Common R Objects (cont'd)

* Differences between Vectors, \textbf{\textcolor{red}{Matrices}}, Data.frames and Lists
```{r, eval = TRUE, include = TRUE, echo = TRUE}
pbc[1:6, c("age", "bili", "chol")]
```


## Common R Objects (cont'd)

* Differences between Vectors, Matrices, \textbf{\textcolor{red}{Data.frames}} and Lists
```{r, eval = TRUE, include = TRUE, echo = TRUE}
pbc[1:6, c("id", "sex", "bili", "chol")]
```


## Common R Objects (cont'd)

* Differences between Vectors, Matrices, Data.frames and \textbf{\textcolor{red}{Lists}}
```{r, eval = TRUE, include = TRUE, echo = TRUE}
list(pbc[1:6, c("sex")], pbc[1:2, c("sex", "bili")], pbc$age[1:4])
```


## Common R Objects (cont'd)
\blue{Create a vector}

```{r, eval = TRUE, include = TRUE, echo = TRUE}
vec <- c(1, 2, 3, 4, 5)
vec
vec <- c(1:5)
vec
```



## Common R Objects (cont'd)
\blue{Create a matrix}

\begincols[t]
\begincol{0.5\linewidth}
```{r, eval = TRUE, include = TRUE, echo = TRUE}
vec <- c(1, 2, 3, 4, 5, 6)
mat <- matrix(vec, 3, 3)
mat 
```

\pause

\endcol
\begincol{0.5\linewidth}
```{r, eval = TRUE, include = TRUE, echo = TRUE}
vec <- c(1, 2, 3, 4, 5, 6)
mat <- matrix(vec, 3, 3, 
              byrow = TRUE)
mat
```
\endcol
\endcols\vspace*{3ex}



## Common R Objects (cont'd)
\blue{Create a data frame}

\begincols[t]
\begincol{0.5\linewidth}
```{r, eval = TRUE, include = TRUE, echo = TRUE}
dtf <- data.frame(pbc$sex, pbc$age)
dtf[1:3,]
```

\pause

\endcol
\begincol{0.5\linewidth}
```{r, eval = TRUE, include = TRUE, echo = TRUE}
dtf <- data.frame(Gender = pbc$sex, 
                  Age = pbc$age)
dtf[1:3,]
```
\endcol
\endcols\vspace*{3ex}



## Common R Objects (cont'd)
\blue{Create a list}

```{r, eval = TRUE, include = TRUE, echo = TRUE}
list1 <- list(vec = c(1:5), mat = pbc[1:2, c("age", "sex")])
list1
```



## Common R Objects (cont'd)

\begin{columns}[T,onlytextwidth]
\begin{column}{0.47\linewidth}
\begin{block}{Demos}
\begin{itemize}
\item Common R Objects \button{02_Common_R_objects.R}{R} \button{02_Common_R_objects.html}{html}
\end{itemize}
\end{block}
\end{column}

\begin{column}{0.47\linewidth}
\end{column}
\end{columns}



## Importing Data and Saving your Work


* function: \textbf{\textcolor{red}{read.table()}}, \textbf{\textcolor{red}{read.csv()}} and its variants
  - note: use forward slashes or double backward slashes in the file names, e.g., \
    "`C:/Documents and Settings/User/Data/file.txt`" or \
    "`C:\\Documents and Settings\\User\\Data\\file.txt`"
    
\vspace*{3ex}    
    
* Specialized functions for importing data from other programs
  - package: \textbf{\textcolor{red}{foreign}}, function: \textbf{\textcolor{red}{read.spss()}}, \textbf{\textcolor{red}{read.dta()}}
  - package: \textbf{\textcolor{red}{Hmisc}}, function: \textbf{\textcolor{red}{sas.get()}}
  - package: \textbf{\textcolor{red}{readxl}}, function: \textbf{\textcolor{red}{read\_excel()}}
  - package: \textbf{\textcolor{red}{haven}}, function: \textbf{\textcolor{red}{read\_spss()}}
  - etc


## Importing Data and Saving your Work (cont'd)

\blue{Tips:}
* Short names are prefered over longer names
* Try to avoid using names that contain symbols
* Avoid spaces in names
* Remove any comments in your data set
* Make sure that any missing values in your data set are indicated with the same value (or no value)

## Importing Data and Saving your Work (cont'd)

* \textbf{\textcolor{red}{save()}}
  - can be used to save a list of R objects
  - a binary file with all the objects available in your last R session
  - platform independent
* You can load your saved R objects using \textbf{\textcolor{red}{load()}}
  - be careful about overwriting
* Using \textbf{\textcolor{red}{saveRDS}} and \textbf{\textcolor{red}{readRDS}} you can save and read a single R object
  - The result has to be assigned to a variable

\vspace*{3ex}

* Specialized functions for exporting data to other programs
  - function: \textbf{\textcolor{red}{write.table()}}, \textbf{\textcolor{red}{write.csv()}}
  - package: \textbf{\textcolor{red}{foreign}}, function: \textbf{\textcolor{red}{write.spss()}}, \textbf{\textcolor{red}{write.dta()}}
  - etc


## Importing Data and Saving your Work (cont'd)

\begin{columns}[T,onlytextwidth]
\begin{column}{0.47\linewidth}
\begin{block}{Demos}
\begin{itemize}
\item Importing and Saving \button{03_Importing_Saving.R}{R} \button{03_Importing_Saving.html}{html}
\end{itemize}
\end{block}
\end{column}

\begin{column}{0.47\linewidth}
\end{column}
\end{columns}

## Data Transformation

* \textbf{\textcolor{red}{Round}} continuous variables
* Convert \textbf{\textcolor{red}{numeric}} variables to \textbf{\textcolor{red}{factors}}
* Compute \textbf{\textcolor{red}{new variables}}
  - transform variables
* Data sets of \textbf{\textcolor{red}{wide}} $\Longleftrightarrow$ \textbf{\textcolor{red}{long}} format


## Data Transformation (cont'd)

* Round continuous variables
```{r, eval = TRUE, include = TRUE, echo = TRUE}
DF <- format(pbc[,c("id", "time", "status", "trt", "age", 
                    "sex", "bili", "chol")], digits = 2)
head(DF)
```


## Data Transformation (cont'd)

* Convert \textbf{\textcolor{red}{numeric}} variables to \textbf{\textcolor{red}{factors}}
```{r, eval = TRUE, include = TRUE, echo = TRUE}
DF <- pbc[,c("id", "time", "status", "trt", "age", 
                    "sex", "bili", "chol")]
DF$trt <- factor(DF$trt, levels = c(1, 2), 
                 labels = c("D-penicillmain", "placebo"))
head(DF)
```


## Data Transformation (cont'd)

* Compute \textbf{\textcolor{red}{new variables}}
  - transform variables
```{r, eval = TRUE, include = TRUE, echo = TRUE}
DF <- pbc[,c("id", "time", "status", "trt", "age", 
                    "sex", "bili", "chol")]
DF$time <- DF$time/30
DF$time_years <- DF$time/12
head(DF)
``` 
  
  
## Data Transformation (cont'd)

* Data sets of \textbf{\textcolor{red}{wide}} $\Longleftrightarrow$ long format  
```{r, eval = TRUE, include = TRUE, echo = TRUE}
head(pbc[,c("id", "time", "status", "trt", "age", 
                    "sex", "bili", "chol")])
```  
  
  
## Data Transformation (cont'd)

* Data sets of wide $\Longleftrightarrow$ \textbf{\textcolor{red}{long}} format  
```{r, eval = TRUE, include = TRUE, echo = TRUE}
head(pbcseq[, c("id", "futime", "status", "trt", "age", "day",
                    "sex", "bili", "chol")])
```  
  

## Data Transformation (cont'd)

\begin{columns}[T,onlytextwidth]
\begin{column}{0.47\linewidth}
\begin{block}{Demos}
\begin{itemize}
\item Data Transformation \button{04_Data_Transformation.R}{R} \button{04_Data_Transformation.html}{html}
\end{itemize}
\end{block}
\end{column}

\begin{column}{0.47\linewidth}
\end{column}
\end{columns}


## Data Exploration

* Common questions
  - What is the mean and sd for age?
  - What is the mean and sd for time?
  - What is the median and interquartile range for age?
  - What is the percentage of placebo patients?
  - What is the percentage of females?
  - What is the mean and sd for age in males?
  - What is the mean and sd for baseline serum bilirubin?
  - What is the percentage of missings in serum bilirubin?

## Data Exploration (cont'd)

* \blue{Hints}

\vspace*{3ex}  

  - Check functions: \textbf{\textcolor{blue}{mean(...)}}, \textbf{\textcolor{blue}{sd(...)}}, \textbf{\textcolor{blue}{percent(...)}}, \textbf{\textcolor{blue}{median(...)}}, \textbf{\textcolor{blue}{IQR(...)}}



## Data Exploration (cont'd)

\begin{columns}[T,onlytextwidth]
\begin{column}{0.47\linewidth}
\begin{block}{Demos}
\begin{itemize}
\item Data Exploration \button{05_Data_Exploration.R}{R} \button{05_Data_Exploration.html}{html}
\end{itemize}
\end{block}
\end{column}

\begin{column}{0.47\linewidth}
\end{column}
\end{columns}


## Data Visualization

* It is important to investigate each variable in our data set using plots
  - Descriptive statistics for continuous and categorical variables
  - Distributions of variables
  - Distributions of variables per group
  - Extreme values
  - Linear/nonlinear evolutions
  
  
## Data Visualization (cont'd)

* R has very powerful graphics capabilities
* Some good references are
  - Murrel, P. (2005) R Graphics. Boca Raton: Chapman & Hall/CRC.
  - Sarkar, D. (2008) Lattice Multivariate Data Visualization with R. New York: Springer-Verlag.


## Data Visualization (cont'd)

* Traditional graphics system
  - package \textbf{\textcolor{blue}{graphics}}
* Trellis graphics system
  - package \textbf{\textcolor{blue}{lattice}} (which is based on package grid)
* Grammar of Graphics implementation (i.e., Wilkinson, L. (1999) The Grammar of Graphics. New York: Springer-Verlag)
  - packages \textbf{\textcolor{blue}{ggplot}} & \textbf{\textcolor{blue}{ggplot2}}


## Data Visualization (cont'd)

Continuous variables
```{r, eval = TRUE, include = TRUE, echo = TRUE, fig.height = 6, fig.width = 15}
plot(pbc$age, pbc$bili)
```


## Data Visualization (cont'd)

Important plotting functions

* \textbf{\textcolor{blue}{plot()}}: scatter plot (and others)
* \textbf{\textcolor{blue}{barplot()}}: bar plots
* \textbf{\textcolor{blue}{boxplot()}}: box-and-whisker plots
* \textbf{\textcolor{blue}{dotchart()}}: dot plots
* \textbf{\textcolor{blue}{hist()}}: histograms
* \textbf{\textcolor{blue}{pie()}}: pie charts
* \textbf{\textcolor{blue}{qqnorm()}}, \textbf{\textcolor{blue}{qqline()}}, \textbf{\textcolor{blue}{qqplot()}}: distribution plots
* \textbf{\textcolor{blue}{pairs()}}: for multivariate data


## Data Visualization (cont'd)

Continuous variables
```{r, eval = TRUE, include = TRUE, echo = TRUE, fig.height = 6, fig.width = 15}
plot(pbc$age, pbc$bili, xlab = "age", ylab = "bili", 
     cex.lab = 1.9, cex.axis = 1.5)
```  


## Data Visualization (cont'd)

Continuous variables
```{r, eval = TRUE, include = TRUE, echo = TRUE, fig.height = 6, fig.width = 15}
plot(pbc$age, pbc$bili, xlab = "age", ylab = "bili", 
     cex.lab = 1.9, cex.axis = 1.5, col = "red")
```  

## Data Visualization (cont'd)

Continuous variables per group
```{r, eval = TRUE, include = TRUE, echo = TRUE, fig.height = 6, fig.width = 15}
boxplot(pbc$age ~ pbc$sex, xlab = "sex", ylab = "age", 
     cex.lab = 1.9, cex.axis = 1.5)
```  


## Data Visualization (cont'd)

Continuous variables per group
```{r, eval = FALSE, include = TRUE, echo = TRUE, fig.height = 3.5, fig.width = 15}
pbc_male_bili <- pbc$bili[pbc$sex == "m"]
pbc_female_bili <- pbc$bili[pbc$sex == "f"]
plot(density(pbc_male_bili), col = rgb(0,0,1,0.5), 
     main = "Density plots", xlab = "age", ylab = "")
polygon(density(pbc_male_bili), col = rgb(0,0,1,0.5), 
        border = "blue")
lines(density(pbc_female_bili), col = rgb(1,0,0,0.5))
polygon(density(pbc_female_bili), col = rgb(1,0,0,0.5), 
        border = "red")
legend(20,0.03, c("male", "female"), 
       col = c(rgb(0,0,1,0.5), rgb(1,0,0,0.5)), lty = 1)
```  


## Data Visualization (cont'd)

Continuous variables per group
```{r, eval = TRUE, include = TRUE, echo = FALSE, fig.height = 6, fig.width = 15}
pbc_male_bili <- pbc$bili[pbc$sex == "m"]
pbc_female_bili <- pbc$bili[pbc$sex == "f"]
plot(density(pbc_male_bili), col = rgb(0,0,1,0.5), 
     main = "Density plots", xlab = "age", ylab = "", ylim = c(0,0.4))
polygon(density(pbc_male_bili), col = rgb(0,0,1,0.5), 
        border = "blue")
lines(density(pbc_female_bili), col = rgb(1,0,0,0.5))
polygon(density(pbc_female_bili), col = rgb(1,0,0,0.5), 
        border = "red")
legend(20,0.03, c("male", "female"), 
       col = c(rgb(0,0,1,0.5), rgb(1,0,0,0.5)), lty = 1)
``` 


## Data Visualization (cont'd)

Categorical variables 
```{r, eval = TRUE, include = TRUE, echo = TRUE, fig.height = 5, fig.width = 15}
pbc$status <- factor(pbc$status, levels = c(0, 1, 2), 
                     labels = c("censored", "transplant", "dead"))
pie(prop.table(table(pbc$status)), 
    col = c("green", "blue", "red"), cex = 2)
``` 


## Data Visualization (cont'd)

\begin{columns}[T,onlytextwidth]
\begin{column}{0.47\linewidth}
\begin{block}{Demos}
\begin{itemize}
\item Data Visualization \button{06_Data_Visualization.R}{R} \button{06_Data_Visualization.html}{html}
\end{itemize}
\end{block}
\end{column}

\begin{column}{0.47\linewidth}
\end{column}
\end{columns}


## Useful Summary: Data manipulation

\begin{columns}[onlytextwidth,T]
\begin{column}{0.35\linewidth}
\textbf{\textcolor{blue}{Common R objects}}
\begin{itemize}\vspace*{-0.5ex}
\item \texttt{c()}
\item \texttt{matrix()}
\item \texttt{data.frame()}
\item \texttt{list()}
\item \texttt{cbind()}, \texttt{rbind()}
\end{itemize}

\bigskip

\textbf{\textcolor{blue}{Transformation}}
\begin{itemize}\vspace*{-0.5ex}
\item \texttt{factor()}
\item \texttt{reshape()}
\item \texttt{order()}
\item \texttt{complete.cases()}
\end{itemize}
\end{column}

\begin{column}{0.30\linewidth}
\textbf{\textcolor{blue}{Exploration}}
\begin{itemize}\vspace*{-0.5ex}
\item \texttt{is.na()}
\item \texttt{dim()}
\item \texttt{mean()}, \texttt{sd()}
\item \texttt{median()}, \texttt{IQR()}
\item \texttt{percent()}
\end{itemize}

\bigskip

\textbf{\textcolor{blue}{Visualization}}
\begin{itemize}
\item \texttt{plot()}, \texttt{legend()}
\item \texttt{hist()}
\item \texttt{barchart()}
\item \texttt{boxplot()}
\item \texttt{par()}
\item \texttt{xyplot()}
\item \texttt{ggplot()}
\end{itemize}
\end{column}

\begin{column}{0.35\linewidth}
\textbf{\textcolor{blue}{Import/Save/Present}}
\begin{itemize}
\item \texttt{head()}
\item \texttt{save()}, \texttt{saveRDS()}
\item \texttt{load()}, \texttt{readRDS()}
\item \texttt{read.csv()}, \texttt{write.csv()}, \texttt{read.xlsx()}, \texttt{write.xlsx()}, \texttt{read.table()}, \texttt{write.table()}
\end{itemize}

\bigskip

\textbf{\textcolor{blue}{Other}}
\begin{itemize}
\item \texttt{install.packages()}
\item \texttt{library()}, \texttt{getwd()}
\item \texttt{ls()}, \texttt{objects()}
\end{itemize}

\end{column}
\end{columns}


## Indexing/Subsetting

* When transforming and analyzing data we often need to select specific observations or variables.
  - Examples: Select ...
      * the 3rd element for vector age
      * the 3rd column of the pbc data set
      * the sex of the 10th patient
      * the baseline details of the 5th patient
      * the serum cholesterol for all males
      * the age for male patients or patients that have serum bilirubin more than 3
      * the first measurement per patient
    

## Indexing/Subsetting (cont'd)

* This can be done using square bracket (\textbf{\textcolor{blue}{[ ]}}) notation and indices.
* Three basic types
  - position indexing
  - logical indexing
  - name indexing
  
  
## Indexing/Subsetting (cont'd)

\textbf{\textcolor{blue}{Indexing with vectors}}

* For position indexing, use a \textbf{\textcolor{red}{positive}} value to select an element
```{r, eval = TRUE, include = TRUE, echo = TRUE}
x <- c(6:17)
x
x[2]
```

* Use multiple positive values to select multiple elements
```{r, eval = TRUE, include = TRUE, echo = TRUE}
x[c(2,3,4)]
```


## Indexing/Subsetting (cont'd)

\textbf{\textcolor{blue}{Indexing with vectors}}

* For position indexing, use dublicated \textbf{\textcolor{red}{positive}} values to select the same elements
```{r, eval = TRUE, include = TRUE, echo = TRUE}
x <- c(6:17)
x
x[c(2,2,2)]
```


## Indexing/Subsetting (cont'd)

\textbf{\textcolor{blue}{Indexing with vectors}}

* For position indexing, use a \textbf{\textcolor{red}{negative}} value to remove an element
```{r, eval = TRUE, include = TRUE, echo = TRUE}
x <- c(6:17)
x
x[-5]
```

\vspace*{3ex}  

* \textbf{Positive and negative indices cannot be combined}


## Indexing/Subsetting (cont'd)

\textbf{\textcolor{blue}{Indexing with vectors}}

* Use logical index of the same length to select elements where the value is \textbf{\textcolor{red}{TRUE}}
```{r, eval = TRUE, include = TRUE, echo = TRUE}
x <- c(6:10)
y <- c(TRUE, FALSE, FALSE, FALSE, FALSE)
x[y]
```


## Indexing/Subsetting (cont'd)

\textbf{\textcolor{blue}{Indexing with vectors}}

* Use logical indexing in combination with conditions
```{r, eval = TRUE, include = TRUE, echo = TRUE}
x <- c(6:10)
x[x > 7]
x[x > 7 & x > 9]
x[x > 7 | x > 9]
```


## Indexing/Subsetting (cont'd)

\textbf{\textcolor{blue}{Indexing with vectors}}

* For name/character indexing, use the name of the element names
```{r, eval = TRUE, include = TRUE, echo = TRUE}
x <- c(foo=5, bar=4, one=7, two=12, three=2)
x[c('foo', 'one')]
```

* Use the function `names` to obtain the names


## Indexing/Subsetting (cont'd)

\textbf{\textcolor{blue}{Indexing with Matrices}}

* Indexing matrices are similar to indexing vectors but with double index

  - The first position denotes the rows
  - The first position denotes the columns

\pause

* Use position indexing as:
```{r, eval = TRUE, include = TRUE, echo = TRUE}
mat <- matrix(1:6, 3, 3)
mat
mat[2, 2]
```  
  
  
## Indexing/Subsetting (cont'd)

\textbf{\textcolor{blue}{Indexing with Matrices}}

* When we leave a position blank all elements are selected
```{r, eval = TRUE, include = TRUE, echo = TRUE}
mat <- matrix(1:6, 3, 3)
mat
mat[2, ]
```  


## Indexing/Subsetting (cont'd)

\textbf{\textcolor{blue}{Indexing with Data Frames}}

* When using a single index the data frame acts like a list of variables

\begincols[t]
\begincol{0.5\linewidth}
```{r, eval = TRUE, include = TRUE, echo = TRUE}
DF <- data.frame(x = 1:4, 
          y = c(35, 23, 14, 45))
DF
```  

\pause

\endcol
\begincol{0.5\linewidth}
* Use position indexing as:
```{r, eval = TRUE, include = TRUE, echo = TRUE}
DF[2]
```  
\endcol
\endcols


## Indexing/Subsetting (cont'd)

\textbf{\textcolor{blue}{Indexing with Data Frames}}

* \textcolor{blue}{\$} provides a convinient notation to extract an element by name
```{r, eval = TRUE, include = TRUE, echo = TRUE}
head(pbc$time)
```  


## Indexing/Subsetting (cont'd)

\textbf{\textcolor{blue}{Indexing with Data Frames}}

\begincols[t]
\begincol{0.5\linewidth}
* When using a double index, indexing works like a matrix
```{r, eval = TRUE, include = TRUE, echo = TRUE}
DF <- data.frame(x = 1:3, 
          y = c(35, 23, 14))
DF
```  

\pause

\endcol
\begincol{0.5\linewidth}
* Use position indexing
```{r, eval = TRUE, include = TRUE, echo = TRUE}
DF[2, 2]
```  
* Use logical indexing
```{r, eval = TRUE, include = TRUE, echo = TRUE}
DF[DF$y < 30, 2]
```  
\endcol
\endcols


## Indexing/Subsetting (cont'd)

\textbf{\textcolor{blue}{Indexing with Data Frames}}

* Combine logical and position indexing in data frame
```{r, eval = TRUE, include = TRUE, echo = TRUE}
pbcDF <- data.frame(pbc)
head(pbc[pbc$sex == "m", 1:7])
```  


## Indexing/Subsetting (cont'd)

\textbf{\textcolor{blue}{Indexing with Data Frames}}

* Combine logical and position indexing in data frame
```{r, eval = TRUE, include = TRUE, echo = TRUE}
pbcDF <- data.frame(pbc)
head(pbc[pbc$age > 30 | pbc$sex == "f", 1:7])
```  


## Indexing/Subsetting (cont'd)

\textbf{\textcolor{blue}{Indexing with Data Frames}}

* Combine logical and position indexing in data frame
```{r, eval = TRUE, include = TRUE, echo = TRUE}
pbcDF <- data.frame(pbc)
head(pbc[pbc$age > 30 & pbc$sex == "f", 1:7])
``` 


## Indexing/Subsetting (cont'd)

\textbf{\textcolor{blue}{Indexing with Lists}}

* Lists can be subsetted in the same way as vectors using single brackets - Note that the output is a list

\vspace*{0.6cm}

* Use position indexing
```{r, eval = TRUE, include = TRUE, echo = TRUE}
mylist <- list(y = c(14, 45), z = c("m", "f", "f"))
mylist[2]
```  


## Indexing/Subsetting (cont'd)

\textbf{\textcolor{blue}{Indexing with Lists}}

* Double square brackets can be also used - Note that the output is a vector
* Use position indexing
```{r, eval = TRUE, include = TRUE, echo = TRUE}
mylist <- list(y = c(14, 45), z = c("m", "f", "f"))
mylist[[2]]
```  


## Indexing/Subsetting (cont'd)

\textbf{\textcolor{blue}{Indexing with Lists}}

* \textcolor{blue}{\$} provides a convinient notation to extract an element by name - Note that the output is a vector
```{r, eval = TRUE, include = TRUE, echo = TRUE}
mylist <- list(y = c(14, 45), z = c("m", "f", "f"))
mylist
mylist$y
```  


## Indexing/Subsetting (cont'd)

* Use the following webpage to further investigate subsetting
https://emcbiostatistics.shinyapps.io/the_apply_family/

* The **R** code for the shiny app is also available: 

\begin{columns}[T,onlytextwidth]
\begin{column}{0.52\linewidth}
\begin{block}{Demos}
\begin{itemize}
\item Indexing Subsetting shiny app \button{08_shiny_app_indexing.R}{R} \
\end{itemize}
\end{block}
\end{column}

\begin{column}{0.48\linewidth}
\end{column}
\end{columns}

In order to run the app you will need to install the package `shiny`


## Useful Summary: Indexing/Subsetting

\begin{columns}[onlytextwidth,T]
\begin{column}{0.5\linewidth}
\textbf{\textcolor{blue}{Vectors}}
\begin{itemize}\vspace*{-0.5ex}
\item \texttt{[]}
\item \texttt{[""]}
\end{itemize}

\bigskip

\textbf{\textcolor{blue}{Matrices}}
\begin{itemize}
\item \texttt{[,]}
\end{itemize}
\end{column}


\begin{column}{0.5\linewidth}
\textbf{\textcolor{blue}{Data frames}}
\begin{itemize}
\item \texttt{[,]}
\item \texttt{[[]]}
\item \texttt{\$}
\end{itemize}

\bigskip

\textbf{\textcolor{blue}{Lists}}
\begin{itemize}
\item \texttt{[]}
\item \texttt{[[]]}
\item \texttt{\$}
\end{itemize}


\end{column}

\end{columns}



## Useful Summary: Indexing/Subsetting (cont'd)

\begin{columns}[T,onlytextwidth]
\begin{column}{0.47\linewidth}
\begin{block}{Demos}
\begin{itemize}
\item Indexing Subsetting \button{07_Indexing_Subsetting.R}{R} \button{07_Indexing_Subsetting.html}{html}
\end{itemize}
\end{block}
\end{column}

\begin{column}{0.47\linewidth}
\begin{block}{Practicals}
\begin{itemize}
\item Basic Use R \button{https://nerler.github.io/BST02/practical/Functions and Programming/Basic_Use_R..html}{html}
\end{itemize}
\end{block}
\end{column}
\end{columns}


