---
title: "BST02: Using R for Statistics\\ in Medical Research"
subtitle: "Part B: Basic use of **R**"
author: "Eleni-Rosalina Andrinopoulou"
institute: "Department of Biostatistics, Erasmus Medical Center"
date: "24 - 28 February 2020"
email: "e.andrinopoulou@erasmusmc.nl"
output:
  beamer_presentation: 
    template: mytemplate.latex
    includes:
      in_header: SlideTemplate.tex
    keep_tex: yes
    incremental: false
classoption: aspectratio=169
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, comment=NA)

library(survival)
library(knitr)
library(kableExtra)
library(dplyr)
```


## Using R

* R is a command-based procedural language
  - write and execute commands
  - use and define functions
  
* You may write the commands in the R console (Windows) or in a shell (Linux)

\vspace*{3ex}

**You will become more familiar with the syntax as you use it** 


## Using R (cont'd)

* Strongly advisable to use a suitable text editor – Some available options:
  - RWinEdt (for Windows; you also need WinEdt installed)
  - Tinn-R (for Windows; http://sciviews.org/Tinn-R/)
  - Rkward (for Linux)
  - Emacs (w. ESS, all platforms)
  - Visual Studio (for Windows)
  - Rstudio (all major platforms; http://www.rstudio.org/) 
  - for more check http://www.sciviews.org/_rgui/projects/Editors.html
  
  
##  Using R (cont’d)

* For this course: Rstudio (http://www.rstudio.org/)
  - free
  - works fine in Windows, MacOS and Linux
  - helpful with errors
  
  
##  Using R (cont’d)

* Can I use R without Rstudio?
* Can I use Rstudio without R?


## Practical Examples

* **Package \textcolor{blue}{survival} - pbc data set**

```{r example data, include=TRUE}
dt <- as.data.frame(pbc[1:5,c(1:6,11,12)])
kable(dt) %>%
  kable_styling("striped", full_width = F)
```

## Practical Examples (cont’d)

* What is a \textcolor{red}{scalar}/vector/matrix

```{r example data b, include=TRUE}
dt <- as.data.frame(pbc[1:5,c(1:6,11,12)])
kable(dt) %>%
  kable_styling("striped", full_width = F) %>%
  column_spec(1, bold = T, color = "red")
```


