---
title: "BST02: Using R for Statistics\\ in Medical Research"
subtitle: "Part B: Basic use of **R**"
author: "Eleni-Rosalina Andrinopoulou"
institute: "Department of Biostatistics, Erasmus Medical Center"
email: "e.andrinopoulou@erasmusmc.nl"
output:
  beamer_presentation: 
    template: mytemplate.latex
    includes:
      in_header: SlideTemplate.tex
    keep_tex: yes
    incremental: false
classoption: aspectratio=169
---
  
```{r setup, include=FALSE}
def.chunk.hook  <- knitr::knit_hooks$get("chunk")
knitr::knit_hooks$set(chunk = function(x, options) {
  x <- def.chunk.hook(x, options)
  ifelse(options$size != "normalsize", paste0("\n \\", options$size,"\n\n", x, "\n\n \\normalsize"), x)
})

knitr::knit_hooks$set(
  nospace = function(before, options, envir) {
    if (before) {
      knitr::asis_output("\\vspace*{-1.5ex}")
    }
  }
)

knitr::opts_chunk$set(echo = FALSE, comment=NA)

library(survival)
library(knitr)
library(kableExtra)
library(dplyr)

pbc <- survival::pbc
pbcseq <- survival::pbcseq
```


##

\blue{\Huge{{Basic Use of R}}}

## In this Section

* Using R

* Examples with Data

* Getting Familiar with R

* Importing data and saving your work

* A lot of practice


## Using R

* R is a command-based procedural language
  - write and execute commands
  - use and define functions
  
* You may write the commands in the R console (Windows) or in a shell (Linux)

\vspace*{3ex}

\blue{You will become more familiar with the syntax as you use it}


## Using R  

* Strongly advisable to use a suitable text editor - Some available options:
  - RWinEdt (for Windows; you also need WinEdt installed)
  - Tinn-R (for Windows; http://sciviews.org/Tinn-R/)
  - Rkward (for Linux)
  - Emacs (w. ESS, all platforms)
  - Visual Studio (for Windows)
  - Rstudio (all major platforms; http://www.rstudio.org/) 
  - for more check https://r-dir.com/blog/2013/01/list-of-r-editors.html
  
  
##  Using R  

* For this course: Rstudio (http://www.rstudio.org/)
  - free
  - works fine in Windows, MacOS and Linux
  - helpful with errors
  - alternative output options
  
  
##  Using R  

* Can I use R without Rstudio?

\vspace*{0.5cm}

* Can I use Rstudio without R?


## Practical Examples

* **Package \texttt{\blue{survival}} - pbc data set**

```{r example data, include=TRUE}
dt <- as.data.frame(pbc[1:5,c(1:6,11,12)])
kable(dt, "latex", booktabs = T, align = "c") %>%
 kable_styling(full_width = F) 
```


## Practical Examples  

* \textbf{\textcolor{red}{id}}:	case number
* \textbf{\textcolor{red}{time}}:	number of days between registration and the earlier of death, transplantation, or study analysis in July, 1986
* \textbf{\textcolor{red}{status}}:	status at endpoint, 0/1/2 for censored, transplant, dead
* \textbf{\textcolor{red}{trt}}:	1/2/NA for D-penicillamine, placebo, not randomised
* \textbf{\textcolor{red}{age}}:	in years
* \textbf{\textcolor{red}{sex}}:	m/f
* \textbf{\textcolor{red}{bili}}:	serum bilirunbin (mg/dl)
* \textbf{\textcolor{red}{chol}}:	serum cholesterol (mg/dl)

More details: https://stat.ethz.ch/R-manual/R-devel/library/survival/html/pbc.html


## Practical Examples  

* What is a \blue{scalar}/vector/matrix

```{r example data b, include=TRUE}
dt <- as.data.frame(pbc[1:5, c(1:6,11,12)])
dt <- format(dt, digits = 7)
dt %>%
 mutate(
   dt = row.names(.),
   age = cell_spec(age, "latex", color = ifelse(age >= 70, "red", "black"), 
                   bold = ifelse(age >= 70, "TRUE", "FALSE"), align = c("c", rep("c", 4))),
   time = cell_spec(time, "latex", align = c("c", rep("c", 4))),
   id = cell_spec(id, "latex", align = c("c", rep("c", 4))),
   sex = cell_spec(sex, "latex", align = c("c", rep("c", 4))),
   status = cell_spec(status, "latex", align = c("c", rep("c", 4))),
   trt = cell_spec(trt, "latex", align = c("c", rep("c", 4))),
   bili = cell_spec(bili, "latex", align = c("c", rep("c", 4))),
   chol = cell_spec(chol, "latex", align = c("c", rep("c", 4))),
   )%>%
#select(c(id, time, status, trt, age, sex, bili, chol)) %>%
kable("latex", escape = F, booktabs = T, align = "c")  %>%
kable_styling(position = "center", full_width = F)
```


## Practical Examples  

* What is a scalar/\blue{vector}/matrix

```{r example data c, include=TRUE}
dt <- as.data.frame(pbc[1:5,c(1:6,11,12)])
kable(dt, "latex", booktabs = T, align = "c") %>%
 kable_styling(full_width = F) %>%
 column_spec(3, bold = TRUE, color = "red")
```


## Practical Examples  

* What is a scalar/\blue{vector}/matrix

```{r example data d, include=TRUE}
dt <- as.data.frame(pbc[1:5,c(1:6,11,12)])
kable(dt, "latex", booktabs = T, align = "c") %>%
 kable_styling(full_width = F) %>%
 column_spec(2, bold = TRUE, color = "red")
```


## Practical Examples  

* What is a scalar/vector/\blue{matrix}

```{r example data e, include=TRUE}
dt <- as.data.frame(pbc[1:5,c(1:6,11,12)])
kable(dt, "latex", booktabs = T, align = "c") %>%
 kable_styling(full_width = F) %>%
 column_spec(2:3, bold = TRUE, color = "red")
```


## Practical Examples  

* What is a scalar/vector/\blue{matrix}

```{r example data f, include=TRUE}
dt <- as.data.frame(pbc[1:5,c(1:6,11,12)])
kable(dt, "latex", booktabs = T, align = "c") %>%
 kable_styling(full_width = F) %>%
 column_spec(7:8, bold = TRUE, color = "red")
```


## Practical Examples  

* Common questions
   - What is the average age?
   - What is the average serum bilirubin?
   - What is the average serum cholesterol?
   - What is the percentage of females?
   - How many missing values do we have for serum cholesterol?

\vspace*{0.3cm}

\blue{All these questions can be answered using R!}


## Getting Familiar with R

* Elementary commands: \textbf{\textcolor{red}{expressions}} and \textbf{\blue{assignments}}

* An \textbf{\textcolor{red}{expression}} given as command is evaluated printed and discarded

* An \textbf{\blue{assignment}} evaluates an expression and passes the value to a variable - the result is not automatically printed


## Getting Familiar with R  

Expression is given as a command,
```{r, eval = TRUE, include = TRUE, echo = TRUE}
103473
```

\vspace*{3ex}

 - However, it cannot be viewed again unless the command is rerun.
 
\pause
 
\vspace*{3ex}

 In order to store information, the expression should assign the command
 
```{r, eval = TRUE, include = TRUE, echo = TRUE}
x <- 103473
x
```


## Getting Familiar with R  

\blue{You can use R as a calculator!}

* Basic arithmetics
 ```{r, eval = FALSE, include = TRUE, echo = TRUE}
+, -, *, /, ^
```

```{r, eval = TRUE, include = TRUE, echo = TRUE}
y <- 103473 + 100000
y
```

* Complicated arithmetics


## Getting Familiar with R  

\blue{Tips:}

* R is case sensitive, e.g.,
  - \textbf{\blue{"sex"}} is different than \textbf{\blue{"Sex"}}
* Commands are separated by a semi-colon or by a newline
* Comments can be put anywhere, starting with a hashmark \textbf{\blue{\#}}: everything to the end of the line is a comment
* Assign a value to an object by \textbf{\blue{<-}} or \textbf{\blue{=}}
* Working directory: get with \textbf{\blue{getwd()}} and set with \textbf{\blue{setwd()}} 


## Getting Familiar with R  

* Missing values
  - are coded as \textbf{\blue{NA}} (i.e., not available) \textbf{\blue{is.na()}}
* Infinity
  - is coded as \textbf{\blue{Inf}} (plus infinity) or \textbf{\blue{-Inf}} (minus infinity) \textbf{\blue{is.finite()}}
* The Null objects
  - are coded as \textbf{\blue{NULL}} (undefined) \textbf{\blue{is.null()}}
* Not a number
  - is coded as \textbf{\blue{NaN}} (Not a Number). Example: 
```{r, eval = TRUE, include = TRUE, echo = TRUE}
0/0
```


## Importing Data 


* function: \textbf{\blue{read.table()}}, \textbf{\blue{read.csv()}} and its variants
  - note: use forward slashes or double backward slashes in the file names, e.g., \
    "`C:/Documents and Settings/User/Data/file.txt`" or \
    "`C:\\Documents and Settings\\User\\Data\\file.txt`"
    
\vspace*{3ex}    
    
* Specialized functions for importing data from other programs
  - package: \textbf{\blue{foreign}}, function: \textbf{\blue{read.spss()}}, \textbf{\blue{read.dta()}}
  - package: \textbf{\blue{Hmisc}}, function: \textbf{\blue{sas.get()}}
  - package: \textbf{\blue{openxlsx}}, function: \textbf{\blue{read.xlsx()}}  
  - package: \textbf{\blue{readxl}}, function: \textbf{\blue{read\_excel()}}
  - package: \textbf{\blue{haven}}, function: \textbf{\blue{read\_spss()}}
  - \textit{etc}


## Exporting Data 

* Specialized functions for exporting data to other programs
  - function: \textbf{\blue{write.table()}}, \textbf{\blue{write.csv()}}
  - package: \textbf{\blue{foreign}}, function: \textbf{\blue{write.spss()}}, \textbf{\blue{write.dta()}}
  - package: \textbf{\blue{openxlsx}}, function: \textbf{\blue{write.xlsx()}}
  - \textit{etc}


## Saving and Loading your Work  

\textit{\textbf{\blue{Multiple objects:}}}

* You can save your R objects using \textbf{\blue{save()}}
  - be careful about overwriting 
* You can load your saved R objects using \textbf{\blue{load()}}

  
\textit{\textbf{\blue{Single object:}}}

* Using \textbf{\blue{saveRDS()}} you can save a single R object 
* Using \textbf{\blue{readRDS()}} you can load a single R object
  - we will need an assignment statement to store the results

\vspace*{0.7cm}

\textbf{\blue{Save your code by using the tab File in Rstudio!}}
  
  
## Saving and Loading your Work  

\textbf{\blue{Tips:}}

* Short names are preferred over longer names
* Try to avoid using names that contain symbols
* Avoid spaces in names
* Remove any comments in your data set
* Make sure that any missing values in your data set are indicated with the same value (or no value)





## Summary

\begin{columns}[onlytextwidth,T]
\begin{column}{0.30\linewidth}


\blue{Basic functions}
\begin{itemize}
\item \texttt{getwd()}, \texttt{setwd()}, 
\item \texttt{is.na()}, \texttt{is.finite()}, \texttt{is.null()}
\end{itemize}

\end{column}

\begin{column}{0.58\linewidth}


\blue{Import/Export}
\begin{itemize}
\item \texttt{read.csv()}, \texttt{write.csv()}
\item \texttt{read.xlsx()}, \texttt{write.xlsx()}
\item \texttt{read.table()}, \texttt{write.table()}
\end{itemize}

\bigskip

\blue{Save/Load}
\begin{itemize}
\item \texttt{save()}, \texttt{saveRDS()}
\item \texttt{load()}, \texttt{readRDS()}
\end{itemize}


\end{column}


\end{columns}



## Practice  

\begin{columns}[T,onlytextwidth]
\begin{column}{0.50\linewidth}
\begin{block}{Demos}
\begin{itemize}
\item Basic R 
\button{https://nerler.github.io/BST02/demo/basic-use-of-r/Basic_R.R}{R} \button{https://nerler.github.io/BST02/demo/basic-use-of-r/Basic_R.html}{html}
\item Importing and Saving 
\button{https://nerler.github.io/BST02/demo/basic-use-of-r/Importing_Saving.R}{R} 
\button{https://nerler.github.io/BST02/demo/basic-use-of-r/Importing_Saving.html}{html}
\end{itemize}
\end{block}
\end{column}

\begin{column}{0.45\linewidth}
\begin{block}{Practicals}
\begin{itemize}
\item Importing and Saving 
\button{https://nerler.github.io/BST02/practical/basic-use-of-r/Importing_and_Saving_data.html}{html}
\end{itemize}
\end{block}
\end{column}
\end{columns}


##

\blue{\Huge{{Common Objects in R}}}


## In this Section

* Objects in R

* Data types

* Data structures

* A lot of practice



## Objects in R

* In R Everything (data, results, ...) is an object

* In order to list the created objects use the following functions
```{r, eval = FALSE, include = TRUE, echo = TRUE}
objects()
ls()
```

\pause

  - In order to remove objects
```{r, eval = FALSE, include = TRUE, echo = TRUE}
rm()
rm(list=ls(all=TRUE))
```


## Objects in R  

* To investigate a specific object (e.g. pbc)
```{r, eval = TRUE, include = TRUE, echo = TRUE}
str(pbc[,c("id", "time", "status", "trt", "age", "sex", "bili", "chol")])
```


## Data Types

The simplest data types are:

* \textbf{\blue{numeric}} : quantitative data
* \textbf{\blue{character}} : qualitative data
* \textbf{\blue{integer}} : whole numbers
* \textbf{\blue{logical}} : TRUE or FALSE
* \textbf{\blue{factors}} : qualitative data (levels)


## Data Types in R

To find out what type of object you have, you can use the following function
```{r, eval = TRUE, include = TRUE, echo = TRUE}
mode(pbc$age)
```

```{r, eval = TRUE, include = TRUE, echo = TRUE}
str(pbc$age)
```

## Data Structures  

The most important data structures are:

  - \textbf{\blue{Scalar}} a single element
  - \textbf{\blue{Vectors}} have the same type of elements
  - \textbf{\blue{Matrices}} have the same type of elements with the same length 
  - \textbf{\blue{Arrays}} have the same type of elements with the same length but can store the data in more than two dimensions
  - \textbf{\blue{Data frames}} have elements of different type with the same length 
  - \textbf{\blue{Lists}} have elements of different type and length 


## Data Structures  

\textbf{\blue{How do these data structures look like?}}


## Data Structures  

* Differences between \blue{vector}, matrix, array, data.frame and list
```{r, eval = TRUE, include = TRUE, echo = TRUE}
pbc[1:6, c("age")]
```

## Data Structures  

* Differences between vector, \blue{matrix}, array, data.frame and list
```{r, eval = TRUE, include = TRUE, echo = TRUE}
pbc[1:6, c("age", "bili", "chol")]
```


## Data Structures  

* Differences between vector, matrix, \blue{array}, data.frame and list
```{r, eval = TRUE, include = TRUE, echo = TRUE}
pbc[1:3, c("age", "bili", "chol")]
pbc[4:6, c("age", "bili", "chol")]
```

## Data Structures  

* Differences between vector, matrix, \blue{array}, data.frame and list
```{r, eval = TRUE, include = TRUE, echo = TRUE}
pbc[1:2, c("protime", "time")]
pbc[3:4, c("protime", "time")]
```

## Data Structures  

* Differences between vector, matrix, array, \blue{data.frame} and list
```{r, eval = TRUE, include = TRUE, echo = TRUE}
pbc[1:6, c("id", "sex", "bili", "chol")]
```


## Data Structures  

* Differences between vector, matrix, array, data.frame and \blue{list}
```{r, eval = TRUE, include = TRUE, echo = TRUE}
pbc[1:6, c("sex")]
pbc[1:2, c("sex", "bili")]
pbc[1:4, c("age")]
```


## Data Structures in R 

\textbf{\blue{Let's now create different data structure in R!}}


## Data Structures in R  
\blue{Create a vector}

```{r, eval = TRUE, include = TRUE, echo = TRUE}
vec <- c(1, 2, 3, 4, 5)
vec
vec <- c(1:5)
vec
```



## Data Structures in R  
\blue{Create a matrix}

\begincols[t]
\begincol{0.5\linewidth}
```{r, eval = TRUE, include = TRUE, echo = TRUE}
vec <- c(1, 2, 3, 4, 5, 6, 7, 8, 9)
mat <- matrix(data = vec, 
              nrow = 3, ncol = 3)
mat 
```

\pause

\endcol
\begincol{0.5\linewidth}
```{r, eval = TRUE, include = TRUE, echo = TRUE}
vec <- c(1, 2, 3, 4, 5, 6, 7, 8, 9)
mat <- matrix(data = vec, 
              nrow = 3, ncol = 3, 
              byrow = TRUE)
mat
```
\endcol
\endcols\vspace*{3ex}


## Data Structures in R  
\blue{Create an array}
```{r, eval = TRUE, include = TRUE, echo = TRUE}
ar <- array(data = c(1, 2, 3, 4, 5, 6, 7, 8), dim = c(2, 2, 2))
ar
```


## Data Structures in R  
\blue{Create an array}
```{r, eval = TRUE, include = TRUE, echo = TRUE}
ar <- array(data = c(1, 2, 3, 4), dim = c(2, 2, 1))
ar
```


## Data Structures in R  
\blue{Create a data frame}

\begincols[t]
\begincol{0.45\linewidth}
```{r, eval = TRUE, include = TRUE, echo = TRUE}
dtf <- data.frame(pbc[, "sex"], 
                  pbc[, "age"])
dtf[1:3,]
```

\pause

\endcol
\begincol{0.58\linewidth}
```{r, eval = TRUE, include = TRUE, echo = TRUE}
dtf <- data.frame(Gender = pbc[, "sex"], 
            Age = pbc[, "age"])
dtf[1:3,]
```
\endcol
\endcols\vspace*{3ex}



## Data Structures in R  
\blue{Create a list}

```{r, eval = TRUE, include = TRUE, echo = TRUE}
list1 <- list(vec = c(1:5), mat = pbc[1:2, c("age", "sex")])
list1
```



## Summary

\begin{columns}[onlytextwidth,T]
\begin{column}{0.30\linewidth}

\bigskip

\blue{Data types}
\begin{itemize}\vspace*{-0.5ex}
\item \texttt{numeric}
\item \texttt{character}
\item \texttt{integer}
\item \texttt{logical}
\item \texttt{factors}
\item \texttt{str()}, \texttt{mode()}
\end{itemize}

\end{column}

\begin{column}{0.58\linewidth}

\blue{Data structures}
\begin{itemize}\vspace*{-0.5ex}
\item \texttt{c()}
\item \texttt{matrix()}
\item \texttt{array()}
\item \texttt{data.frame()}
\item \texttt{list()}
\end{itemize}

\bigskip

\blue{Other}
\begin{itemize}
\item \texttt{ls()}, \texttt{objects()}
\end{itemize}

\end{column}
\end{columns}



## Practice  

\begin{columns}[T,onlytextwidth]
\begin{column}{0.47\linewidth}
\begin{block}{Demos}
\begin{itemize}
\item Common R Objects 
\button{https://nerler.github.io/BST02/demo/basic-use-of-r/Common_R_Objects.R}{R} \button{https://nerler.github.io/BST02/demo/basic-use-of-r/Common_R_Objects.html}{html}
\end{itemize}
\end{block}
\end{column}

\begin{column}{0.47\linewidth}
\begin{block}{Practicals}
\begin{itemize}
\item Common R Objects 
\button{https://nerler.github.io/BST02/practical/basic-use-of-r/Common_R_Objects.html}{html}
\end{itemize}
\end{block}
\end{column}
\end{columns}


##

\blue{\Huge{{Indexing and Subsetting}}}


## In this Section

* Indexing

* Subsetting

* A lot of practice


## Indexing/Subsetting

* When transforming and analyzing data we often need to select specific observations or variables
  - Examples: Select ...
      * the 3rd element of vector age
      * the 3rd column of the pbc data set
      * the sex of the 10th patient
      * all information of the 5th patient
      * the serum cholesterol for all males
      * the age for male patients or patients that have serum bilirubin > 3
      * the first measurement per patient
    

## Indexing/Subsetting  

* This can be done using square bracket (\textbf{\blue{[ ]}}) notation and indices.
* Three basic types
  - position indexing
  - logical indexing
  - name indexing
  
  
## Vectors  

\blue{Indexing with \blue{\texttt{vector}}}

* For position indexing, use a \textbf{\blue{positive}} value to select an element
```{r, eval = TRUE, include = TRUE, echo = TRUE}
x <- c(6:17)
x
x[2]
```

* Use multiple positive values to select multiple elements
```{r, eval = TRUE, include = TRUE, echo = TRUE}
x[c(2,3,4)]
```


## Vectors  

\blue{Indexing with \blue{\texttt{vector}}}

* For position indexing, use duplicated \textbf{\blue{positive}} values to select the same elements
```{r, eval = TRUE, include = TRUE, echo = TRUE}
x <- c(6:17)
x
x[c(2,2,2)]
```


## Vectors  

\blue{Indexing with \blue{\texttt{vector}}}

* For position indexing, use a \textbf{\blue{negative}} value to remove an element
```{r, eval = TRUE, include = TRUE, echo = TRUE}
x <- c(6:17)
x
x[-5]
```

\vspace*{3ex}  

* \textbf{\blue{Positive and negative indices cannot be combined}}


## Vectors  

\blue{Indexing with \blue{\texttt{vector}}}

* Use logical index of the same length to select elements where the value is \textbf{\blue{TRUE}}
```{r, eval = TRUE, include = TRUE, echo = TRUE}
x <- c(6:10)
y <- c(TRUE, FALSE, FALSE, FALSE, FALSE)
x[y]
```


## Vectors  

\blue{Indexing with \blue{\texttt{vector}}}

* Use logical indexing in combination with conditions
```{r, eval = TRUE, include = TRUE, echo = TRUE}
x <- c(6:10)
x[x > 7]
x[x > 7 & x > 9]
x[x > 7 | x > 9]
```


## Vectors  

\blue{Indexing with \blue{\texttt{vector}}}

* For name/character indexing, use the name of the element
```{r, eval = TRUE, include = TRUE, echo = TRUE}
x <- c(foo=5, bar=4, one=7, two=12, three=2)
x[c('foo', 'one')]
```

* Use the function `names` to obtain the names


## Matrices  

\blue{Indexing with \blue{\texttt{matrix}}}

* Indexing matrices is similar to indexing vectors but with double index

  - The first position denotes the rows \blue{["index", ]}
  - The first position denotes the columns \blue{[, "index"]}

\pause

\begincols[t]
\begincol{0.5\linewidth}
```{r, eval = TRUE, include = TRUE, echo = TRUE}
mat <- matrix(data = 1:4, 
              nrow = 2, ncol = 2)
mat
```  
  
\endcol
\begincol{0.5\linewidth}
* Use position indexing as:
```{r, eval = TRUE, include = TRUE, echo = TRUE}
mat <- matrix(data = 1:4, 
              nrow = 2, ncol = 2)
mat[2, 2]
```  
\endcol
\endcols


## Matrices  

\blue{Indexing with \blue{\texttt{matrix}}}

* Be cautious, it also works with a single index. In this case, it selects the particular element of the vector that will be included in the matrix

\begincols[t]
\begincol{0.5\linewidth}
```{r, eval = TRUE, include = TRUE, echo = TRUE}
mat <- matrix(data = 1:4, 
              nrow = 2, ncol = 2)
mat[2]
mat[[2]]
```  
  
\endcol
\begincol{0.5\linewidth}
\endcol
\endcols
  
## Matrices  

\blue{Indexing with \blue{\texttt{matrix}}}

\begincols[t]
\begincol{0.5\linewidth}
* When we leave a position blank all elements are selected
```{r, eval = TRUE, include = TRUE, echo = TRUE}
mat <- matrix(data = 1:4, 
              nrow = 2, ncol = 2)
mat
mat[2, ]
```  

\endcol
\begincol{0.5\linewidth}
\endcol
\endcols



## Arrays 

\blue{Indexing with \blue{\texttt{array}}}

\begincols[t]
\begincol{0.5\linewidth}

```{r, eval = TRUE, include = TRUE, echo = TRUE}
ar <- array(data = 1:4, 
            dim = c(1,2,2))
ar
```  
\endcol
\begincol{0.5\linewidth}
```{r, eval = TRUE, include = TRUE, echo = TRUE, size = 0.5}
ar[1, 1, ]
```  

\endcol
\endcols


## Data Frames  

\blue{Indexing with \blue{\texttt{data.frame}}}
* Works with single and double index

\begincols[t]
\begincol{0.55\linewidth}
```{r, eval = TRUE, include = TRUE, echo = TRUE}
DF <- data.frame(x = 1:3, 
      y = c("male", "male", "female"))
DF
```  

\pause

\endcol
\begincol{0.45\linewidth}
* Use position single indexing
```{r, eval = TRUE, include = TRUE, echo = TRUE}
DF[2]
DF[[2]]
```  
\endcol
\endcols



## Data Frames  

\blue{Indexing with \blue{\texttt{data.frame}}}

\begincols[t]
\begincol{0.55\linewidth}
* When using double index, indexing works like a matrix
```{r, eval = TRUE, include = TRUE, echo = TRUE}
DF <- data.frame(x = 1:3, 
      y = c("male", "male", "female"))
DF
```  


\endcol
\begincol{0.45\linewidth}
* Use position indexing
```{r, eval = TRUE, include = TRUE, echo = TRUE}
DF[2, ]
```  
* Use logical indexing
```{r, eval = TRUE, include = TRUE, echo = TRUE}
DF[DF$x < 2, ]
```  
\endcol
\endcols

## Data Frames  

\blue{Indexing with \blue{\texttt{data.frame}}}

* \blue{\$} provides a convenient notation to extract an element by name
```{r, eval = TRUE, include = TRUE, echo = TRUE}
head(pbc$time)
head(pbc[ ,"time"])
```  




## Data Frames  

\blue{Indexing with \blue{\texttt{data.frame}}}

* Combine logical and position indexing in data frame
```{r, eval = TRUE, include = TRUE, echo = TRUE}
head(pbc[pbc$sex == "m", 1:7])
```  


## Data Frames  

\blue{Indexing with \blue{\texttt{data.frame}}}

* Combine logical and position indexing in data frame
```{r, eval = TRUE, include = TRUE, echo = TRUE}
head(pbc[pbc$age > 30 | pbc$sex == "f", 1:7])
```  


## Data Frames  

\blue{Indexing with \blue{\texttt{data.frame}}}

* Combine logical and position indexing in data frame
```{r, eval = TRUE, include = TRUE, echo = TRUE}
head(pbc[pbc$age > 30 & pbc$sex == "f", 1:7])
``` 


## Lists  

\blue{Indexing with \blue{\texttt{list}}}

* Lists can be subsetted in the same way as vectors using single brackets - Note that the output is a list

\vspace*{0.6cm}

* Use position indexing
```{r, eval = TRUE, include = TRUE, echo = TRUE}
mylist <- list(y = c(14, 45), z = c("m", "f", "f"))
mylist[2]
```  


## Lists  

\blue{Indexing with \blue{\texttt{list}}}

* Double square brackets can be also used - Note that the output is a vector

\vspace*{0.6cm}

* Use position indexing
```{r, eval = TRUE, include = TRUE, echo = TRUE}
mylist <- list(y = c(14, 45), z = c("m", "f", "f"))
mylist[[2]]
```  


## Lists  

\blue{Indexing with \blue{\texttt{list}}}

* \blue{\$} provides a convenient notation to extract an element by name - Note that the output is a vector
```{r, eval = TRUE, include = TRUE, echo = TRUE}
mylist <- list(y = c(14, 45), z = c("m", "f", "f"))
mylist
mylist$y
```  



## Summary

\begin{columns}[onlytextwidth,T]
\begin{column}{0.5\linewidth}
\blue{Vectors}
\begin{itemize}\vspace*{-0.5ex}
\item \texttt{[]}
\item \texttt{[""]} - for categorical variables
\end{itemize}

\bigskip

\blue{Matrices}
\begin{itemize}
\item \texttt{[,]}
\item \texttt{[[]]}, \texttt{[]}
\end{itemize}

\bigskip

\blue{Arrays}
\begin{itemize}
\item \texttt{[ , , ]}
\end{itemize}
\end{column}


\begin{column}{0.5\linewidth}
\blue{Data frames}
\begin{itemize}
\item \texttt{[,]}
\item \texttt{[[]]}, \texttt{[]}
\item \texttt{\$}
\end{itemize}

\bigskip

\blue{Lists}
\begin{itemize}
\item \texttt{[]}
\item \texttt{[[]]}
\item \texttt{\$}
\end{itemize}


\end{column}

\end{columns}


## Practice  

* Use the following webpage to further investigate indexing and subsetting \
https://emcbiostatistics.shinyapps.io/indexing/

\begin{center}
\begin{columns}
\begin{column}{0.52\linewidth}
\begin{block}{Demos}
\begin{itemize}
\item Shiny app indexing subsetting \button{https://nerler.github.io/BST02/demo/basic-use-of-r/shiny_app_indexing.R}{R} 
\end{itemize}
\end{block}
\end{column}
\end{columns}
\end{center}

In order to run the app you will need to install the packages:

   * `survival`
   * `shiny`


## Practice   

\begin{columns}[T,onlytextwidth]
\begin{column}{0.47\linewidth}
\begin{block}{Demos}
\begin{itemize}
\item Indexing/Subsetting 
\button{https://nerler.github.io/BST02/demo/basic-use-of-r/Indexing_Subsetting.R}{R} \button{https://nerler.github.io/BST02/demo/basic-use-of-r/Indexing_Subsetting.html}{html}
\end{itemize}
\end{block}
\end{column}

\begin{column}{0.47\linewidth}
\begin{block}{Practicals}
\begin{itemize}
\item Indexing/Subsetting  \button{https://nerler.github.io/BST02/practical/basic-use-of-r/Indexing_Subsetting.html}{html}
\end{itemize}
\end{block}
\end{column}
\end{columns}


##

\blue{\Huge{{Data Manipulation}}}


## In this Section

* Data transformation

* Data exploration

* Data visualization 

* A lot of practice


## Data Transformation

\blue{You will never receive the perfect data set!}

* \blue{Round} continuous variables
* Convert \blue{numeric} variables to \blue{factors}
* Compute \blue{new variables}
  - transform variables
* \blue{Sort} the data set
* Data sets of \blue{wide} $\Longleftrightarrow$ \blue{long} format


## Data Transformation  

* \blue{Round} continuous variables
```{r, eval = TRUE, include = TRUE, echo = TRUE}
pbc[1:3, c("time", "age", "bili", "chol")]
round(pbc[1:3, c("time", "age", "bili", "chol")], digits = 2)
```


## Data Transformation  

* Convert \blue{numeric} variables to \blue{factors}
```{r, eval = TRUE, include = TRUE, echo = TRUE}
DF <- pbc[,c("id", "time", "status", "trt", "age", 
                    "sex", "bili", "chol")]
head(DF)
```


## Data Transformation  

* Convert \blue{numeric} variables to \blue{factors}
```{r, eval = TRUE, include = TRUE, echo = TRUE}
DF <- pbc[,c("id", "time", "status", "trt", "age", 
                    "sex", "bili", "chol")]
DF$trt <- factor(x = DF$trt, levels = c(1, 2), 
                 labels = c("D-penicillmain", "placebo"))
head(DF)
```


## Data Transformation  

* Compute \blue{new variables}
  - transform variables
```{r, eval = TRUE, include = TRUE, echo = TRUE}
DF <- pbc[,c("id", "time", "status", "trt", "age", 
                    "sex", "bili", "chol")]
head(DF)
``` 


## Data Transformation  

* Compute \blue{new variables}
  - transform variables
```{r, eval = TRUE, include = TRUE, echo = TRUE}
DF <- pbc[,c("id", "time", "status", "trt", "age", 
                    "sex", "bili", "chol")]
DF$time <- DF$time/30
DF$time_years <- DF$time/12
head(DF)
``` 
  
  
## Data Transformation  

* \blue{Sort} the data set in either ascending or descending order
  * The variable by which we sort can be a numeric, string or factor 
```{r, eval = TRUE, include = TRUE, echo = TRUE}
head(sort(pbc$bili))
``` 


## Data Transformation  

* \blue{Sort} the data set in either ascending or descending order
  * The variable by which we sort can be a numeric, string or factor 
```{r, eval = TRUE, include = TRUE, echo = TRUE, size = "footnotesize"}
head(pbc[order(pbc$bili), ])
``` 


## Data Transformation  

* \blue{Sort} the data set in either ascending or descending order
  * The variable by which we sort can be a numeric, string or factor 
```{r, eval = TRUE, include = TRUE, echo = TRUE, size = "footnotesize"}
head(pbc[order(pbc$bili, pbc$age), ])
```


## Data Transformation  

* Data sets of \blue{wide} $\Longleftrightarrow$ long format  
```{r, eval = TRUE, include = TRUE, echo = TRUE}
head(pbc[,c("id", "time", "status", "trt", "age", 
                    "sex", "bili", "chol")])
```  
  
  
## Data Transformation  

* Data sets of wide $\Longleftrightarrow$ \blue{long} format  
```{r, eval = TRUE, include = TRUE, echo = TRUE}
head(pbcseq[, c("id", "futime", "status", "trt", "age", "day",
                    "sex", "bili", "chol")])
```  
  

## Data Transformation  

* Data sets of \blue{wide} $\Longleftrightarrow$ \blue{long} format  
```{r, eval = FALSE, include = TRUE, echo = TRUE}
?reshape
```



## Data Exploration

* Common questions for the pbc data set
  - What is the mean and standard deviation for age?
  - What is the mean and standard deviation for time?
  - What is the median and interquartile range for age?
  - What is the percentage of placebo patients?
  - What is the percentage of females?
  - What is the mean and standard deviation for age in males?
  - What is the mean and standard deviation for baseline serum bilirubin?
  - What is the percentage of missings in serum bilirubin?
  
\vspace*{0.3cm}

\blue{All these questions can be answered using R!}


## Data Exploration  

* \blue{Hints}

\vspace*{3ex}  
    - Check functions: \blue{mean(...)}, \blue{sd(...)}, \blue{percent(...)}, \blue{median(...)}, \blue{IQR(...)}, \blue{table(...)}

\pause
\vspace*{3ex} 

What is the mean value for age?
```{r, eval = TRUE, include = TRUE, echo = TRUE}
mean(pbc$age)
```  
 


## Data Visualization

* It is important to investigate each variable in our data set using plots
  - Descriptive statistics for continuous and categorical variables
  - Distributions of variables
  - Distributions of variables per group
  - Extreme values
  - Linear/nonlinear evolutions

<!-- https://genomicsclass.github.io/book/pages/plots_to_avoid.html -->
  
## Data Visualization
\blue{Take care!}
```{r, eval = TRUE, include = TRUE, echo = FALSE, fig.height = 10, fig.width = 15}
pie(table(pbc$bili), cex = 0.8, main = "Serum bilirubin")
```


## Data Visualization
\blue{Take care!}
```{r, eval = TRUE, include = TRUE, echo = FALSE, fig.height = 7, fig.width = 15}
pbcseq.id <- pbcseq[tapply(row.names(pbcseq), pbcseq$id, tail, 1), ]
pbcseq.idBase <- pbcseq[tapply(row.names(pbcseq), pbcseq$id, head, 1), ]

pbcseq.id$BA <- "Last"
pbcseq.idBase$BA <- "First"

pbcFL <- rbind(pbcseq.idBase, pbcseq.id)
pbcFL <- pbcFL[order(pbcFL$id, pbcFL$day),]
library(lattice)
p1 <- xyplot(bili ~ as.factor(BA), group = id, data = pbcFL[pbcFL$id %in% c(1, 24, 25, 6, 34, 17, 45, 23), ], 
             type = c("p", "l"), pch = 16,
             xlab = "", ylab = list("Serum bilirubin", cex = 1.3), col = "#C443AF",
       scales = list(x = list(limits = c(0.9,2.1), at = c(1,2), cex = 1.2, 
                          labels = c("First", "Last")), y = list(cex = 1.2) ))
p2 <- bwplot(bili ~ as.factor(BA), data = pbcFL[pbcFL$id %in% c(1, 24, 25, 6, 34, 17, 45, 23), ],
            ylab = list("Serum bilirubin", cex = 1.3), xlab = "", 
            scales = list(x = list(cex = 1.2), y = list(cex = 1.2)),
            par.settings = list(box.umbrella = list(col= c("#C443AF")), 
                                    box.dot = list(col= c("#C443AF")), 
                                    box.rectangle = list(col= c("#C443AF"))))

print(p1, split = c(1,1,2,1), more = TRUE)
print(p2, split = c(2,1,2,1))
```


## Data Visualization

\blue{Take care!}
```{r, eval = TRUE, include = TRUE, echo = FALSE, fig.width = 4, fig.height = 3}
library(lattice)
xyplot(bili ~ as.factor(BA), group = id, data = pbcFL[pbcFL$id %in% c(45, 6), ], 
       type = c("p", "l"), pch = 16, 
       xlab = "", ylab = list("Serum bilirubin", cex = 0.9), col = "#C443AF",
       scales = list(x = list(limits = c(0.9,2.1), at = c(1,2), cex = 0.8, 
                          labels = c("First", "Last")), y = list(cex = 0.8, draw = FALSE) ))
```

## Data Visualization

\blue{Take care!}
```{r, eval = TRUE, include = TRUE, echo = FALSE, fig.width = 4, fig.height = 3}
library(lattice)
xyplot(bili ~ as.factor(BA), group = id, data = pbcFL[pbcFL$id %in% c(45, 6), ], 
       type = c("p", "l"), pch = 16, 
       xlab = "", ylab = list("Serum bilirubin", cex = 0.9), col = "#C443AF",
       scales = list(x = list(limits = c(0.9,2.1), at = c(1,2), cex = 0.8, 
                          labels = c("First", "Last")), y = list(cex = 0.8) ))
```


## Data Visualization

\blue{Take care!}
```{r, eval = TRUE, include = TRUE, echo = FALSE, fig.width = 4, fig.height = 3}
library(lattice)
xyplot(bili ~ as.factor(BA), group = id, data = pbcFL[pbcFL$id %in% c(45, 6), ], 
       type = c("p", "l"), pch = 16, 
       xlab = "", ylab = list("Serum bilirubin", cex = 0.9), col = "#C443AF",
       ylim = c(0, 28), 
       scales = list(x = list(limits = c(0.9,2.1), at = c(1,2), cex = 0.8, 
                          labels = c("First", "Last")), y = list(cex = 0.8, limits = c(0, 28)) ))
```


## Data Visualization  

* R has very powerful graphics capabilities
* Some good references are
  - Murrel, P. (2005) R Graphics. Boca Raton: Chapman & Hall/CRC.
  - Sarkar, D. (2008) Lattice Multivariate Data Visualization with R. New York: Springer-Verlag.


## Data Visualization  

* Traditional graphics system
  - package \blue{graphics}
* Trellis graphics system
  - package \blue{lattice} (which is based on package grid)
* Grammar of Graphics implementation (i.e., Wilkinson, L. (1999) The Grammar of Graphics. New York: Springer-Verlag)
  - packages \blue{ggplot} & \blue{ggplot2}


## Data Visualization  

Important plotting basic functions

* \blue{plot()}: scatter plot (and others)
* \blue{barplot()}: bar plots
* \blue{boxplot()}: box-and-whisker plots
* \blue{hist()}: histograms
* \blue{dotchart()}: dot plots
* \blue{pie()}: pie charts
* \blue{qqnorm()}, \blue{qqline()}, \blue{qqplot()}: distribution plots
* \blue{pairs()}: for multivariate data


## Data Visualization  

Continuous variables
```{r, eval = TRUE, include = TRUE, echo = TRUE, fig.height = 6, fig.width = 15}
plot(x = pbc$age, y = pbc$bili)
```


## Data Visualization  

Continuous variables
```{r, eval = TRUE, include = TRUE, echo = TRUE, fig.height = 6, fig.width = 15}
plot(x = pbc$age, y = pbc$bili, xlab = "age", ylab = "bili", 
     cex.lab = 1.9, cex.axis = 1.5)
```  


## Data Visualization  

Continuous variables
```{r, eval = TRUE, include = TRUE, echo = TRUE, fig.height = 6, fig.width = 15}
plot(x = pbc$age, y = pbc$bili, xlab = "age", ylab = "bili", 
     cex.lab = 1.9, cex.axis = 1.5, col = "red")
```  


## Data Visualization  

* For more options check 
```{r, eval = FALSE, include = TRUE, echo = TRUE, fig.height = 6, fig.width = 15}
?plot
```  


## Data Visualization  

Continuous variables per group
```{r, eval = TRUE, include = TRUE, echo = TRUE, fig.height = 6, fig.width = 15}
plot(x = pbc$age, y = pbc$bili, xlab = "age", ylab = "bili", 
     cex.lab = 1.9, cex.axis = 1.5, col = pbc$sex)
```  


## Data Visualization  

Continuous variables per group
```{r, eval = TRUE, include = TRUE, echo = TRUE, fig.height = 6, fig.width = 15}
boxplot(formula = pbc$age ~ pbc$sex, xlab = "sex", ylab = "age", 
     cex.lab = 1.9, cex.axis = 1.5)
```  

## Data Visualization  

Continuous variables per group
```{r, eval = FALSE, include = TRUE, echo = TRUE, fig.height = 3.5, fig.width = 15}
pbc_male_bili <- pbc$bili[pbc$sex == "m"]
pbc_female_bili <- pbc$bili[pbc$sex == "f"]
plot(density(x = pbc_male_bili), col = rgb(0,0,1,0.5), 
     main = "Density plots", xlab = "bili", ylab = "")
lines(density(x = pbc_female_bili), col = rgb(1,0,0,0.5))
legend(x = 8, y = 0.2, legend = c("male", "female"), 
       col = c(rgb(0,0,1,0.5), rgb(1,0,0,0.5)), lty = 1)
```  


## Data Visualization  

Continuous variables per group
```{r, eval = TRUE, include = TRUE, echo = FALSE, fig.height = 6, fig.width = 15}
pbc_male_bili <- pbc$bili[pbc$sex == "m"]
pbc_female_bili <- pbc$bili[pbc$sex == "f"]
plot(density(x = pbc_male_bili), col = rgb(0,0,1,0.5), 
     main = "Density plots", xlab = "bili", ylab = "")
lines(density(x = pbc_female_bili), col = rgb(1,0,0,0.5))
legend(x = 8, y = 0.2, legend = c("male", "female"), 
       col = c(rgb(0,0,1,0.5), rgb(1,0,0,0.5)), lty = 1)
```

## Data Visualization  

Continuous variables per group
```{r, eval = FALSE, include = TRUE, echo = TRUE, fig.height = 3.5, fig.width = 15}
pbc_male_bili <- pbc$bili[pbc$sex == "m"]
pbc_female_bili <- pbc$bili[pbc$sex == "f"]
plot(density(x = pbc_male_bili), col = rgb(0,0,1,0.5), 
     main = "Density plots", xlab = "bili", ylab = "")
polygon(density(x = pbc_male_bili), col = rgb(0,0,1,0.5), 
        border = "blue")
lines(density(x = pbc_female_bili), col = rgb(1,0,0,0.5))
polygon(density(x = pbc_female_bili), col = rgb(1,0,0,0.5), 
        border = "red")
legend(x = 8, y = 0.2, legend = c("male", "female"), 
       col = c(rgb(0,0,1,0.5), rgb(1,0,0,0.5)), lty = 1)
```  


## Data Visualization  

Continuous variables per group
```{r, eval = TRUE, include = TRUE, echo = FALSE, fig.height = 6, fig.width = 15}
pbc_male_bili <- pbc$bili[pbc$sex == "m"]
pbc_female_bili <- pbc$bili[pbc$sex == "f"]
plot(density(x = pbc_male_bili), col = rgb(0,0,1,0.5), 
     main = "Density plots", xlab = "bili", ylab = "")
polygon(density(x = pbc_male_bili), col = rgb(0,0,1,0.5), 
        border = "blue")
lines(density(x = pbc_female_bili), col = rgb(1,0,0,0.5))
polygon(density(x = pbc_female_bili), col = rgb(1,0,0,0.5), 
        border = "red")
legend(x = 8, y = 0.2, legend = c("male", "female"), 
       col = c(rgb(0,0,1,0.5), rgb(1,0,0,0.5)), lty = 1)
``` 


## Data Visualization  

Categorical variables 
```{r, eval = TRUE, include = TRUE, echo = TRUE, fig.height = 5, fig.width = 15}
pbc$status <- factor(x = pbc$status, levels = c(0, 1, 2), 
                     labels = c("censored", "transplant", "dead"))
pie(table(pbc$status), col = c("green", "blue", "red"), cex = 2)
``` 




## Summary

\begin{columns}[onlytextwidth,T]
\begin{column}{0.30\linewidth}


\blue{Transformation}
\begin{itemize}\vspace*{-0.5ex}
\item \texttt{round()}
\item \texttt{factor()}
\item \texttt{order()}
\item \texttt{reshape()}
\end{itemize}
\end{column}

\begin{column}{0.30\linewidth}
\blue{Exploration}
\begin{itemize}\vspace*{-0.5ex}
\item \texttt{mean()}, \texttt{sd()}
\item \texttt{median()}, \texttt{IQR()}
\item \texttt{table()}
\end{itemize}
\end{column}

\begin{column}{0.30\linewidth}
\blue{Visualization}
\begin{itemize}
\item \texttt{plot()}, \texttt{legend()}
\item \texttt{hist()}
\item \texttt{barchart()}
\item \texttt{boxplot()}
\item \texttt{xyplot()}, \texttt{ggplot()}
\item \texttt{par()}
\end{itemize}
\end{column}

\end{columns}



## Practice   

\begin{columns}[T,onlytextwidth]
\begin{column}{0.47\linewidth}
\begin{block}{Demos}
\begin{itemize}
\item Data Transformation 
\button{https://nerler.github.io/BST02/demo/basic-use-of-r/Data_Transformation.R}{R} \button{https://nerler.github.io/BST02/demo/basic-use-of-r/Data_Transformation.html}{html}
\item Data Exploration 
\button{https://nerler.github.io/BST02/demo/basic-use-of-r/Data_Exploration.R}{R} 
\button{https://nerler.github.io/BST02/demo/basic-use-of-r/Data_Exploration.html}{html}
\item Data Visualization 
\button{https://nerler.github.io/BST02/demo/basic-use-of-r/Data_Visualization.R}{R} \button{https://nerler.github.io/BST02/demo/basic-use-of-r/Data_Visualization.html}{html}
\end{itemize}
\end{block}
\end{column}


\begin{column}{0.47\linewidth}
\begin{block}{Practicals}
\begin{itemize}
\item Data Manipulation 
\button{https://nerler.github.io/BST02/practical/basic-use-of-r/Data_Transformation_Exploration_Visualization.html}{html}
\end{itemize}
\end{block}
\end{column}

\end{columns}










