---
title: "BST02: Using R for Statistics\\ in Medical Research"
subtitle: "Part C: Functions and Programming"
author: ["Nicole Erler", "Eleni-Rosalina Andrinopoulou"]
institute: "Department of Biostatistics, Erasmus Medical Center"
date: "24 - 28 February 2020"
email: ["n.erler@erasmusmc.nl", "e.andrinopoulou@erasmusmc.nl"]
output:
  beamer_presentation: 
    template: mytemplate.latex
    includes:
      in_header: SlideTemplate.tex
    incremental: false
classoption: aspectratio=169
---

```{r setup, include=FALSE}
knitr::knit_hooks$set(
  nospace = function(before, options, envir) {
    if (before) {
      knitr::asis_output("\\vspace*{-1.5ex}")
    }
  }
)

knitr::opts_chunk$set(echo = TRUE, nospace = TRUE, nospaceafter = TRUE)
```

## Recap Part B
\begin{columns}[onlytextwidth,T]
\begin{column}{0.2\linewidth}

\blue{Objects}
\begin{itemize}
\item \texttt{vector}
\item \texttt{matrix}
\item \texttt{data.frame}
\item \texttt{list}
\end{itemize}

\bigskip

\onslide<2->{
\blue{Operators}
\begin{itemize}
\item \texttt{+}, \texttt{-}, \texttt{*}, \texttt{/}
\item \texttt{<-}, \texttt{=}
\item \texttt{<}, \texttt{>}, \texttt{==}
\end{itemize}
}
\end{column}

\begin{column}{0.25\linewidth}
\blue{Data Structures}
\begin{itemize}
\item \texttt{numeric}
\item \texttt{character}
\item \texttt{integer}
\item \texttt{logical}
\item \texttt{factor}
\end{itemize}

\bigskip

\onslide<2->{
\blue{Special Values}
\begin{itemize}
\item \texttt{NA}
\item \texttt{NaN}
\item \texttt{Inf}, \texttt{-Inf}
\end{itemize}
}
\end{column}

\begin{column}{0.5\linewidth}
\onslide<3->{
\blue{Data Transformations}
\begin{itemize}
\item rounding (\texttt{format()})
\item convert to factor (\texttt{factor()})
\end{itemize}
}

\bigskip

\onslide<4->{
\blue{Data Exploration}
\begin{itemize}
\item \texttt{mean()}, \texttt{median()}, \texttt{sd()}, \texttt{IQR()}, \ldots
\end{itemize}

\bigskip


\blue{Data Visualizations}
\begin{itemize}
\item plotting packages
\item plot types (\texttt{plot()}, \texttt{barplot()}, ...)
\end{itemize}
}

\bigskip

\onslide<5->{
\blue{Subsetting}
\begin{itemize}
\item \texttt{[[...]]}, \texttt{[...]}, \texttt{\$}, \ldots
\end{itemize}
}
\end{column}
\end{columns}


## In this Section

* What are functions?

* Useful functions for data exploration

* Useful functions for data manipulations

* Writing functions

* Control-flow constructs

* The `apply` family

* Lots of practising


## Functions

Sometimes we want to perform the same action / manipulation on several objects.

* Option 1: copy & paste
  - a lot of work
  - susceptible to mistakes

\pause

* Option 2: \blue{functions}

\pause

\blue{What are functions?}
* a group of (organized) R commands
* a (small) program with flexible (= not pre-specified) input

\vspace*{3ex}

**Almost all commands in R are functions!**

## Functions
\blue{Some examples:}
\begincols[t]
\begincol{0.3\linewidth}

* `mean()`
* `sum()`
* `plot()`
* ...

\endcol
\begincol{0.65\linewidth}
```{r, collapse=TRUE, nospace = TRUE}
class(mean)
class(sum)
class(plot)
```
\endcol
\endcols\vspace*{3ex}

\pause

Even `class()` is a function:
```{r}
class(class)
```

## Useful Functions for Data Exploration

\begin{columns}[T,onlytextwidth]
\begin{column}{0.47\linewidth}
\begin{block}{Demos}
\begin{itemize}
\item Functions for Data Exploration \button{https://nerler.github.io/BST02/demo/functions-and-programming/Functions_for_Data_Exploration.R}{R} \button{https://nerler.github.io/BST02/demo/functions-and-programming/Functions_for_Data_Exploration.html}{html}
\end{itemize}
\end{block}
\end{column}

\begin{column}{0.47\linewidth}
\begin{block}{Practicals}
\begin{itemize}
\item Exploring and Summarizing Data \button{https://nerler.github.io/BST02/practical/functions-and-programming/Exploring_and_Summarizing_Data.html}{html}
\end{itemize}
\end{block}
\end{column}
\end{columns}



## Useful Functions for Data Exploration
\begin{columns}[onlytextwidth,T]
\begin{column}{0.28\linewidth}
\blue{Dimension}
\begin{itemize}\vspace*{-0.5ex}
\item \texttt{dim()}
\item \texttt{nrow()}, \texttt{ncol()}
\item \texttt{length()}
\end{itemize}

\bigskip

\blue{Data Structure}
\begin{itemize}\vspace*{-0.5ex}
\item \texttt{str()}
\item \texttt{names()}, 
\item \texttt{head()}, \texttt{tail()}
\item \texttt{is.data.frame()}, \texttt{is.list()}, \texttt{is.matrix()}
      \texttt{is.numeric()}, \texttt{is.ordered()}, \ldots
\end{itemize}

\end{column}

\begin{column}{0.32\linewidth}
\blue{Descriptives for Continuous Variables}
\begin{itemize}\vspace*{-1ex}
\item \texttt{summary()}
\item \texttt{min()}, \texttt{max()}, \texttt{range()}
\item \texttt{mean()}, \texttt{median()}, \texttt{quantile()}, \texttt{IQR()}
\item \texttt{sd()}, \texttt{var()}
\item \texttt{ave()}
\end{itemize}

\bigskip

\blue{Tables}
\begin{itemize}\vspace*{-0.5ex}
\item \texttt{table()}, \texttt{prop.table()}
\item \texttt{addmargins()}, \texttt{ftable()}
\end{itemize}
\end{column}


\begin{column}{0.36\linewidth}
\blue{for \texttt{matrix} \& \texttt{data.frame}}
\begin{itemize}\vspace*{-0.5ex}
\item \texttt{summary()}
\item \texttt{var()}, \texttt{cor()}, \texttt{cov2cor()}
\item \texttt{colSums()}, \texttt{colMeans()}, \texttt{rowSums()}, \texttt{rowMeans()}
\end{itemize}

\bigskip

\blue{Duplicates \& Comparison}
\begin{itemize}\vspace*{-0.5ex}
\item \texttt{duplicated()}
\item \texttt{unique()}
\end{itemize}

\end{column}
\end{columns}

## Useful functions for Data Manipulation

\begin{columns}[T,onlytextwidth]
\begin{column}{0.47\linewidth}
\begin{block}{Demo}
\begin{itemize}
\item Functions for Data Manipulation \button{https://nerler.github.io/BST02/demo/functions-and-programming/Functions_for_Data_Manipulation.R}{R} \button{https://nerler.github.io/BST02/demo/functions-and-programming/Functions_for_Data_Manipulation.html}{html} \end{itemize}
\end{block}
\end{column}

\begin{column}{0.47\linewidth}
\begin{block}{Practicals}
\begin{itemize}
\item Merging Datasets \button{https://nerler.github.io/BST02/practical/functions-and-programming/Merging_Datasets.html}{html}
\end{itemize}
\end{block}
\end{column}
\end{columns}


## Useful functions for Data Manipulation
\begin{columns}[onlytextwidth,T]
\begin{column}{0.4\linewidth}
\blue{Transformations}
\begin{itemize}\vspace*{-0.5ex}
\item \texttt{log()}, \texttt{log2()}, \texttt{log10()}
\item \texttt{exp()}, \texttt{sqrt()}, \texttt{plogis()}
\end{itemize}

\bigskip

\blue{Splitting \& Combining}
\begin{itemize}\vspace*{-0.5ex}
\item \texttt{split()}, \texttt{cut()}
\item \texttt{cbind()}, \texttt{rbind()}
\item \texttt{merge()}
\item \texttt{subset()}
\item \texttt{c()}
\item \texttt{paste()}
\end{itemize}
\end{column}

\begin{column}{0.45\linewidth}
\blue{Sorting}
\begin{itemize}\vspace*{-0.5ex}
\item \texttt{sort()}, \texttt{order()}, \texttt{rev()}, \texttt{rank()}
\end{itemize}

\bigskip

\blue{Repetition \& Sequence}
\begin{itemize}\vspace*{-0.5ex}
\item \texttt{rep()}, \texttt{seq()}
\item \texttt{expand.grid()}
\end{itemize}

\bigskip

\blue{Converting Objects}
\begin{itemize}
\item \texttt{t()}
\item \texttt{unlist()}, \texttt{unname()}
\item \texttt{as.numeric()}, \texttt{as.matrix()}, \texttt{as.data.frame()}
\end{itemize}
\end{column}
\end{columns}

## Writing Functions
To write your own function:
```{r, eval = FALSE}
myfun <- function(arguments) {
  syntax
}
```

\pause

For example:
```{r}
square <- function(x) {
  x^2
}
```

```{r}
square(3)
```

## Writing Functions
Functions do not always need an argument:
```{r}
random <- function() {
  rnorm(n = 1)
}
```
```{r, collapse = TRUE}
random()
random()
random()
```

## Writing Functions
Functions can use \blue{multiple arguments}:
```{r}
subtract <- function(x, y) {
  x - y
}
```
```{r}
subtract(x = 5.2, y = 3.3)
```


## Writing Functions
Multiple arguments are interpreted in the \blue{pre-defined order}, unless they are named:
```{r}
subtract(5.2, 1.2)
```
is equivalent to
```{r}
subtract(x = 5.2, y = 1.2)
```

\pause
But this is different:
```{r}
subtract(y = 5.2, x = 1.2)
```

## Writing Functions
We can also define \blue{default values} for arguments.
```{r}
multiply <- function(x, y = 2) {
  x * y
}
```

The default value is used when the user does not specify a value for that argument:
```{r}
multiply(x = 3, y = 3)
multiply(x = 3)
```


## Writing Functions
\begin{center}
\begin{columns}
\begin{column}{0.47\linewidth}
\begin{block}{Practical}
\begin{itemize}
\item Rolling the Dice \button{https://nerler.github.io/BST02/practical/functions-and-programming/Rolling_the_Dice.html}{html}
\end{itemize}
\end{block}
\end{column}
\end{columns}
\end{center}

## Control-flow Constructs: `if()`
Sometimes, we may want to execute code only \blue{if a certain condition is fulfilled}.

To do this, we can use an \blue{\texttt{if}} statement
```{r eval = FALSE}
if (condition) {expression}
```

\pause

For example:
```{r}
x <- c(0.3, -1.2, 0.8,  1.7,  0.7, -0.1, -0.4, -0.1, -0.2,  0.6)
if (length(x) > 5) {mean(x)}
```
\pause
```{r}
x <- c(0.7, -0.1, -0.4, -0.1)
if (length(x) > 5) {mean(x)}
```

If the condition is not fulfilled, `NULL` is returned.

## Control-flow Constructs: `if()` and `else`
We can also specify an expression that is evaluated \blue{if the condition is not
fulfilled}:

```{r, eval = FALSE}
if (condition) {expression} else {alternative expression}
```

\pause

For example:
```{r}
if (length(x) > 5) {
  mean(x)
} else {
  x
}
```



## Conditional Element Selection: `ifelse()`
A similar function is \blue{\texttt{ifelse()}}, which performs \blue{conditional element selection}:
```{r, eval = FALSE}
ifelse(test, yes, no)
```

\pause

For example:
```{r}
x <- c(0.3, -1.2, 0.8,  1.7,  0.7, -0.1, -0.4, -0.1, -0.2,  0.6)
ifelse(x > 0, ">0", "<0")
```

\pause

\blue{Note:}

* `if()` expects \blue{one} condition
* `ifelse()` expects a \blue{vector of conditions}


## Control-flow Constructs: `for()`-loop
To perform an operation multiple times, we can use a \blue{\texttt{for}-loop}
```{r, eval = FALSE}
for (variable in sequence) {expression}
```

For example:
```{r}
for (i in 1:5) {
  print(2 * i)
}
```

## Control-flow Constructs: `for()`-loop
In a \blue{\texttt{for}-loop}, the `variable` does not need to be used in the `expression`:

```{r}
for (i in 1:5) {
  print('test')
}
```

Note that when using `for()`, always the full `sequence` is used, i.e., we cannot
skip iterations.


## Control-flow Constructs: `while()`-loop

The function \blue{\texttt{while()}} repeatedly evaluates an expression as 
long as a condition is fulfilled:
```{r, eval = FALSE}
while (condition) {expression}
```


\bigskip

\parbox[t]{\linewidth}{
\blue{Careful:}\\
If your condition is never \texttt{FALSE} this will run forever!!!\newline
(or until you stop it manually)
}

\bigskip

\parbox[t]{\linewidth}{
\blue{Note:}\\
\texttt{for()} and \texttt{while()} loops will not print output, unless we specifically use
the function \texttt{print()}.
}


## Control-flow Constructs: `while()`-loop

```{r, echo = FALSE}
set.seed(1234)
```

For example:
```{r}
s <- 1
while (s < 5) {
  s <- s + s/2
  print(s)
}
```


## Control-flow Constructs

\begin{columns}[T,onlytextwidth]
\begin{column}{0.47\linewidth}
\begin{block}{Demo}
\begin{itemize}
\item Control Flow \button{https://nerler.github.io/BST02/demo/functions-and-programming/Control_Flow.R}{R} \button{https://nerler.github.io/BST02/demo/functions-and-programming/Control_Flow.html}{html}
\end{itemize}
\end{block}
\end{column}

\begin{column}{0.47\linewidth}
\begin{block}{Practicals}
\begin{itemize}
\item Control Flow and Functions \button{https://nerler.github.io/BST02/practical/functions-and-programming/Control_Flow_and_Functions.html}{html}
\item Custom Subset Function
\button{https://nerler.github.io/BST02/practical/functions-and-programming/Custom_Subset_Function.html}{html}
\end{itemize}
\end{block}
\end{column}
\end{columns}



## Summary: Writing Functions
```{r, eval = FALSE}
function_name <- function(arguments) {
  "function body"
}
```

* can have 0, 1, 2, ...arguments
* arguments are interpreted in the \blue{pre-specified order}, unless the \blue{names are used}
* we can specify \blue{default values}


## Summary: Control-flow Constructs

\begin{itemize}\itemsep3mm
\item \blue{\texttt{if (condition) expression}}:\newline
  evaluates the \texttt{expression} only if the \texttt{condition} is \texttt{TRUE}
\item \blue{\texttt{if (condition) expression1 else expression2}}:\newline
   evaluates \texttt{expression1} if the 
  \texttt{condition} is \texttt{TRUE} and \texttt{expression2} if the \texttt{condition} is \texttt{FALSE}
\item \blue{\texttt{ifelse(test, yes, no)}}:\newline
   expects a vector of \texttt{test}s
\item \blue{\texttt{for()}} and \blue{\texttt{while()}} loops:\newline
  can be used to repeatedly perform the same action
\item to print output from within \texttt{for()} and \texttt{while()} 
      we need to use \texttt{print()}
\end{itemize}




```{r setup2, include=FALSE}
# This is the setup chunk for Elrozy's part of these slides
knitr::opts_chunk$set(echo = FALSE, comment = NA)

library(survival)
```


##  What is the apply Family

Manipulate \blue{vectors} or slices of data from \blue{matrices}, \blue{data frames} and \blue{lists} in a repetitive way avoiding explicit use of loop-constructs

\vspace*{0.7cm}

  - An aggregating function, like for example the mean, or the sum
  - Other transforming or subsetting functions
  - Other vectorized functions, which return more complex structures like lists, vectors and matrices
  
  
## What is the apply Family  

\blue{apply(), lapply() , sapply(), tapply(), mapply()}

\vspace*{3ex} 

\textbf{But how and when should we use these?}


## How To Use apply() in R

* Operates on \blue{\texttt{matrix}} and \blue{\texttt{data.frame}}

\begincols[t, onlytextwidth]
\begincol{0.47\linewidth}
* By column
```{r, eval = TRUE, include = TRUE, echo = TRUE}
mat <- matrix(1:6, 3, 3)
mat
apply(mat, 2, sum)
```

\pause

\endcol
\begincol{0.47\linewidth}
* By row
```{r, eval = TRUE, include = TRUE, echo = TRUE}
apply(mat, 1, sum)
```

\endcol
\endcols


## How To Use apply() in R  

* Operates on \blue{\texttt{matrix}} and \blue{\texttt{data.frame}}

\begincols[t, onlytextwidth]
\begincol{0.47\linewidth}
* By column
```{r, eval = TRUE, include = TRUE, echo = TRUE}
mat <- matrix(1:6, 3, 3)
mat
apply(mat, 2, mean)
```

\pause

\endcol
\begincol{0.47\linewidth}
* By row
```{r, eval = TRUE, include = TRUE, echo = TRUE}
apply(mat, 1, mean)
```

\endcol
\endcols


## How To Use apply() in R  

* You can also apply your own functions

\begincols[t, onlytextwidth]
\begincol{0.47\linewidth}
* By column
```{r, eval = TRUE, include = TRUE, echo = TRUE}
mat <- matrix(1:6, 3, 3)
mat
apply(mat, 2, function(x) 
          sum(x)/(length(x)-1))
```

\pause

\endcol
\begincol{0.47\linewidth}
* By row
```{r, eval = TRUE, include = TRUE, echo = TRUE}
apply(mat, 1, function(x) 
          sum(x)/(length(x)-1))
```

\endcol
\endcols


## How To Use lapply() in R 

* Apply a given function to every element of a \blue{\texttt{list}} and return a \blue{\texttt{list}}

\vspace*{0.6cm}

* The difference with \blue{\texttt{apply()}}:
  - It can be used for other objects like \blue{\texttt{vector}}, \blue{\texttt{data.frame}} or \blue{\texttt{list}}
  - The output returned is a list


## How To Use lapply() in R  

\begincols[t, onlytextwidth]
\begincol{0.47\linewidth}

```{r, eval = TRUE, include = TRUE, echo = TRUE}
myList <- list(x = c(1:6), 
               y = c("m", "f"), 
               z = c(30, 4, 23))
myList
```

\pause

\endcol
\begincol{0.47\linewidth}
* Use pre-specified functions
```{r, eval = TRUE, include = TRUE, echo = TRUE}
lapply(myList, length)
```

\endcol
\endcols


## How To Use lapply() in R  

\begincols[t, onlytextwidth]
\begincol{0.47\linewidth}

```{r, eval = TRUE, include = TRUE, echo = TRUE}
myList <- list(x = c(1:6), 
               y = c("m", "f"), 
               z = c(30, 4, 23))
myList
```


\endcol
\begincol{0.47\linewidth}
* Use pre-specified functions
```{r, eval = TRUE, include = TRUE, echo = TRUE, warning = FALSE}
lapply(myList, median)
```
* You can also apply your own functions!

\endcol
\endcols


## How To Use sapply() in R 

* \blue{\texttt{sapply()}} is similar to \blue{\texttt{lapply()}}, but it tries to simplify the output

\begincols[t, onlytextwidth]
\begincol{0.47\linewidth}

```{r, eval = TRUE, include = TRUE, echo = TRUE}
myList <- list(x = c(1:6), 
               y = c("m", "f"), 
               z = c(30, 4, 23))
myList
```

\pause

\endcol
\begincol{0.47\linewidth}
* Use pre-specified functions
```{r, eval = TRUE, include = TRUE, echo = TRUE, warning = FALSE}
sapply(myList, length)
sapply(myList, median)
```
* You can also apply your own functions!

\endcol
\endcols


## How To Use tapply() in R

* Apply a function to subsets of a \blue{\texttt{vector}} - The subsets are defined by some other \blue{\texttt{vector}}, usually a factor

```{r, eval = TRUE, include = TRUE, echo = TRUE}
tapply(pbc$bili, pbc$sex, mean)
tapply(pbc$age, pbc$sex, median)
```


## How To Use tapply() in R  

* You can also apply your own functions

```{r, eval = TRUE, include = TRUE, echo = TRUE}
tapply(pbc$bili, pbc$sex, function(x) sum(x)/(length(x)-1))
```


## How To Use mapply() in R

* Multivariate apply
* Its purpose is to be able to vectorize arguments to a function that is not usually accepting \blue{\texttt{vectors}} as arguments
* \blue{\texttt{mapply()}} applies a function to multiple \blue{\texttt{list}} or multiple \blue{\texttt{vector}} arguments

\vspace*{2ex}
```{r, echo = FALSE}
options(width = 70)
```
```{r, eval = FALSE, echo = TRUE}
mapply(length, pbc)
```
\vspace*{-2ex}
```{r, eval = TRUE, echo = FALSE}
# this is necessary to adjust the spacing between echo and output on the slide
mapply(length, pbc)
```
\vspace*{-3ex}

## How To Use mapply() in R  

* Overlapping between functions

\begincols[t, onlytextwidth]
\begincol{0.47\linewidth}
\vspace*{-3ex}
```{r, eval = TRUE, include = TRUE, echo = TRUE}
myList <- list(x = c(1:6), 
               y = c("m", "f"), 
               z = c(30, 4, 23))
mapply(length, myList, 
       SIMPLIFY = FALSE)
```

\pause

\endcol
\begincol{0.47\linewidth}
```{r, eval = TRUE, include = TRUE, echo = TRUE, warning = FALSE}
lapply(myList, length)
```
* You can also apply your own functions!

\endcol
\endcols

## Useful Summary: Apply Family

\begin{columns}[onlytextwidth,T]
\begin{column}{0.47\linewidth}
\blue{Vectors}
\begin{itemize}\vspace*{-0.5ex}
\item \texttt{tapply()}
\item \texttt{mapply()}
\end{itemize}

\bigskip

\blue{Matrices}
\begin{itemize}
\item \texttt{apply()}
\item \texttt{tapply()}
\item \texttt{lapply()}
\item \texttt{sapply()}
\item \texttt{mapply()}
\end{itemize}
\end{column}


\begin{column}{0.47\linewidth}
\blue{Data frames}
\begin{itemize}
\item \texttt{apply()}
\item \texttt{tapply()}
\item \texttt{lapply()}
\item \texttt{sapply()}
\item \texttt{mapply()}
\end{itemize}

\bigskip

\blue{Lists}
\begin{itemize}
\item \texttt{lapply()}
\item \texttt{sapply()}
\item \texttt{mapply()}
\end{itemize}


\end{column}

\end{columns}

## Useful Summary: Apply Family  

* Use the following webpage to further investigate the apply family \
https://emcbiostatistics.shinyapps.io/the_apply_family/

* The **R** code for the shiny app is also available: 

\begin{center}
\begin{columns}
\begin{column}{0.52\linewidth}
\begin{block}{Demos}
\begin{itemize}
\item Shiny app apply family \button{https://nerler.github.io/BST02/demo/functions-and-programming/shiny_app_apply_family.R}{R} 
\end{itemize}
\end{block}
\end{column}
\end{columns}
\end{center}

In order to run the app you will need to install the packages:

   * `survival`
   * `shiny`


## Useful Summary: Apply Family  


\begin{columns}[T,onlytextwidth]
\begin{column}{0.47\linewidth}
\begin{block}{Demos}
\begin{itemize}
\item The Apply Family 
\button{https://nerler.github.io/BST02/demo/functions-and-programming/The_Apply_Family.R}{R} \button{https://nerler.github.io/BST02/demo/functions-and-programming/The_Apply_Family.html}{html}
\end{itemize}
\end{block}
\end{column}

\begin{column}{0.47\linewidth}
\begin{block}{Practicals}
\begin{itemize}
\item The Apply Family \button{https://nerler.github.io/BST02/practical/functions-and-programming/The_Apply_Family.html}{html}
\end{itemize}
\end{block}
\end{column}
\end{columns}

