---
title: "BST02: Using R for Statistics\\ in Medical Research"
subtitle: "Part D: Statistics with R"
author: "Nicole Erler"
institute: "Department of Biostatistics, Erasmus Medical Center"
date: "24 - 28 February 2020"
email: "n.erler@erasmusmc.nl"
output:
  beamer_presentation: 
    template: mytemplate.latex
    includes:
      in_header: SlideTemplate.tex
    incremental: false
classoption: aspectratio=169
---

```{r setup, include=FALSE}
knitr::knit_hooks$set(
  nospace = function(before, options, envir) {
    if (before) {
      knitr::asis_output("\\vspace*{-1.5ex}")
    }
  }
)

knitr::opts_chunk$set(echo = TRUE, nospace = TRUE, nospaceafter = TRUE,
                      fig.align = 'center', out.width = "100%")


library(ggplot2)
EMCdark <- "#0c2074"
# \definecolor{EMCdark}{HTML}{0c2074}
# \definecolor{EMC80}{HTML}{3d4d90}
# \definecolor{EMC60}{HTML}{6d79ac}
# \definecolor{EMC40}{HTML}{9ea6c7}
# \definecolor{EMC20}{HTML}{ced2e3}
# \definecolor{darkred}{RGB}{204,0,0}
drkred <- rgb(204, 0, 0, maxColorValue = 255)
```

## In this Section


\begin{itemize}\itemsep3mm
\item Common statistical tests
  \begin{itemize}
  \item for continuous data
  \item for categorical data
  \end{itemize}
\item (Generalized) linear regression
\item Useful functions for regression models
\item Modelling non-linear effects
\end{itemize}




## t-test: `t.test()`
\blue{One-sample t-test}

* compares the \blue{mean of one sample} with a fixed value $\mu$

\pause
\bigskip

\blue{Two sample / independent samples t-test}

* compares the \blue{difference between the means} of two samples with a fixed value $\mu$

\pause
\bigskip

\blue{Related samples t-test}

* compares the \blue{mean of the difference} between related observations with a fixed value $\mu$
  (same as one-sample t-test)

## Wilcoxon Test: `wilcox.test()`
\blue{Wilcoxon Signed Rank Test}

* tests if \blue{one sample} (or the difference between two paired samples) is 
  \blue{symmetric about $\mu$}

\pause\bigskip

\blue{Wilcoxon Rank Sum Test / Mann-Whitney test}

* test for a \blue{location shift between the distributions} of two independent samples

\bigskip
\vfill
See also [BBR Sections 7.2 & 7.3 (http://hbiostat.org/doc/bbr.pdf)](http://hbiostat.org/doc/bbr.pdf)


## Kruskal-Wallis Rank Sum Test: `kruskal.test()`

\begin{itemize}\itemsep4mm
\item \blue{extension} of the Wilcoxon rank sum test for \blue{more than two groups}
\item test for a \blue{difference in location} of a continuous variable between multiple groups
\item the \blue{Wilcoxon rank sum test is a special case} of the Kruskal-Wallis rank sum test
\end{itemize}

## Other tests for continuous data

\begin{itemize}\itemsep5mm
\item \blue{Kolmogorov-Smirnov Test:} \texttt{ks.test()}\newline
  tests if two samples are drawn from the same continuous distribution
  
\item \blue{Shapiro-Wilk Normality Test:} \texttt{shapiro.test()}

\item \blue{Friedman Rank Sum Test:} \texttt{friedman.test()}\newline
  non-parametric test for two or more related samples
  
\item \ldots
\end{itemize}


## Tests for Continuous Data

\begin{center}
\begin{columns}
\begin{column}{0.47\linewidth}
\begin{block}{Demo}
\begin{itemize}
\item Tests for Continuous Data\newline \button{https://nerler.github.io/BST02/demo/statistics-with-r/Tests_for_Continuous_Data.R}{R}
\button{https://nerler.github.io/BST02/demo/statistics-with-r/Tests_for_Continuous_Data.html}{html}
\end{itemize}
\end{block}
\end{column}
\end{columns}
\end{center}


## Tests for Categorical Data / Proportions

\blue{One-sample Proportion Test}
* tests if the \blue{proportion in one sample} is equal to a fixed value $p$
* `prop.test()` and `binom.test()`

\pause\bigskip

\blue{Tests for Proportions in Multiple (independent) Groups}
* tests if the \blue{proportions in several samples} are equal
* `chisq.test()` and `fisher.test()` (when there are cells with 0)

\bigskip\vfill

See also [BBR Sections 5.7 & 6 (http://hbiostat.org/doc/bbr.pdf)](http://hbiostat.org/doc/bbr.pdf)


## Tests for Categorical Data / Proportions
\blue{Related Samples: McNemar Test}
 * Tests for \blue{symmetry} in a $2\times 2$ table
 * `mcnemar.test()`
 
 
\pause\bigskip


\blue{3-Dimensional Contingency Table}
* Cochrane-Mantel-Haenszel Test
* $\chi^2$ test for \blue{independence} of two nominal variables \blue{within each stratum}
* `mantelhaen.test()`


## Tests for Categorical Data

\begin{columns}
\begin{column}{0.47\linewidth}
\begin{block}{Demo}
\begin{itemize}
\item Tests for Categorical Data\newline \button{https://nerler.github.io/BST02/demo/statistics-with-r/Tests_for_Categorical_Data.R}{R}
\button{https://nerler.github.io/BST02/demo/statistics-with-r/Tests_for_Categorical_Data.html}{html}
\end{itemize}
\end{block}
\end{column}
\begin{column}{0.47\linewidth}
\begin{block}{Practical}
\begin{itemize}
\item Statistical Tests\newline
\button{https://nerler.github.io/BST02/practical/statistics-with-r/Statistical_Tests.html}{html}
\end{itemize}
\end{block}
\end{column}
\end{columns}


## Useful Functions: Statistical Tests
\begin{columns}[onlytextwidth,T]
\begin{column}{0.3\linewidth}

\blue{Continuous Outcomes}
\begin{itemize}\vspace*{-0.5ex}
\item \texttt{t.test()}
\item \texttt{wilcox.test()}
\item \texttt{kruskal.test()}
\item \texttt{ks.test()}
\item \texttt{friedman.test()}
\item \texttt{shapiro.test()}
\end{itemize}
\end{column}

\begin{column}{0.4\linewidth}
\blue{Categorical Outcomes}
\begin{itemize}\vspace*{-0.5ex}
\item \texttt{prop.test()}
\item \texttt{binom.test()}
\item \texttt{chisq.test()}
\item \texttt{fisher.test()}
\item \texttt{mcnemar.test()}
\item \texttt{mantelhaen.test()}
\end{itemize}

\bigskip

\blue{Pairwise tests}
\begin{itemize}
\item \texttt{pairwise.prop.test()}
\item \texttt{pairwise.t.test()}
\item \texttt{pairwise.wilcox.test()}
\end{itemize}
\end{column}

\begin{column}{0.3\linewidth}
\blue{Variance and Correlation}
\begin{itemize}\vspace*{-0.5ex}
\item \texttt{cor.test()}
\item \texttt{bartlett.test()}
\item \texttt{var.test()}
\end{itemize}

\bigskip

\blue{Multiple Testing Adjustment}
\begin{itemize}
\item \texttt{p.adjust()}
\end{itemize}
\end{column}
\end{columns}



## Linear Regression

A standard linear regression model has the form
$$ y = \beta_0 + \beta_1 x_1 + \beta_2 x_2 + \ldots + \beta_p x_p + \varepsilon 
\qquad \text{with}\quad \varepsilon \sim N(0, \sigma^2)$$

where

\begin{itemize}\itemsep2.5mm
\item \blue{$y$} is the \blue{outcome} variable ("dependent variable")
\item \blue{$x_1, \ldots, x_p$} are the \blue{covariates} ("independent variables")
\item \blue{$\beta_0, \ldots, \beta_p$} are the \blue{regression coefficients}
  \begin{itemize}
  \item $\beta_0$ is the intercept
  \item $\beta_1, \ldots, \beta_p$ estimate the effects of the covariates
  \end{itemize}
\item \blue{$\varepsilon$} is a vector of \blue{residuals}, which we assume to be
  (approximately) normally distributed.
\end{itemize}


## Linear Regression 

To fit a \blue{linear regression} in R we use the function \blue{\texttt{lm()}}.

\bigskip

The most important arguments are

\begin{itemize}\itemsep2mm
\item \blue{\texttt{formula}}:\newline a \texttt{formula} object
\item \blue{\texttt{data}}: \newline a \texttt{data.frame} (optional, but usually needed)
\item \blue{\texttt{subset}}:\newline a \texttt{vector} specifying which observations should be used (optional)\newline
      (works like the \texttt{subset} argument of the function \texttt{subset()})

\end{itemize}

## Model Formula
A \blue{\texttt{formula}} object has the form
```{r, eval = FALSE}
outcome ~ linear predictor
```
for example
```{r, eval = FALSE}
y ~ x1 + x2 + x3
```

\bigskip\pause 

\begin{itemize}\itemsep2mm
\item Variables are separated by "\texttt{\blue{+}}" signs.
\item An intercept is automatically included.
\item One-sided formulas (omitting the outcome) are possible\newline
  (used for random effects specification).
\end{itemize}


## Model Formula: Interactions
Interaction terms are written using "\blue{\texttt{:}}" or "\blue{\texttt{*}}".

"\blue{\texttt{*}}" includes the main effects and interaction terms, i.e., 
```{r, eval = FALSE}
y ~ x1 * x2
```
\vspace*{-2ex}is equivalent to
```{r, eval = FALSE}
y ~ x1 + x2 + x1:x2
```

\pause
Interactions between multiple variables can be written using "\blue{\texttt{()}}", i.e.,
```{r, eval = FALSE}
y ~ x1 * (x2 + x3)
```
\vspace*{-2ex}is equivalent to
```{r, eval = FALSE}
y ~ x1 * x2 + x1 * x3
```

## Model Formula: Interactions
To specify a \blue{higher level interaction} "\blue{\texttt{\^}}" is used.

For example:
```{r, eval = FALSE}
y ~ (x1 + x2 + x3)^3
```
\vspace*{-2ex}will create all interactions up to 3-way and is equivalent to
```{r, eval = FALSE}
y ~ x1 * x2 * x3
```
\vspace*{-2ex} and equivalent to
```{r, eval = FALSE}
y ~ x1 + x2 + x3 + x1:x2 + x1:x3 + x2:x3 + x1:x2:x3
```
\pause
and
```{r, eval = FALSE}
y ~ (x1 + x2 + x3)^2
```
\vspace*{-2ex}will create all two-way interactions and is equivalent to
```{r, eval = FALSE}
y ~ x1 + x2 + x3 + x1:x2 + x1:x3 + x2:x3
```

## Model Formula: Removing terms
The "\blue{\texttt{-}}" sign can be used to remove terms from a model formula.

For example
```{r, eval = FALSE}
y ~ x1 * x2 * x3 - x2 - x1:x3
```
\vspace*{-2ex}is equivalent to 
```{r, eval = FALSE}
y ~ x1 + x3 + x1:x2 + x2:x3 + x1:x2:x3
```

\pause
The \blue{intercept} can be removed from a formula by using "\blue{\texttt{-1}}"
or "\blue{\texttt{+0}}", i.e.
```{r, eval = FALSE}
y ~ x1 + x2 - 1
y ~ x1 + x2 + 0
```

## Generalized Linear Regression (GLM)
A \blue{generalized linear regression} model has the form
$$g(\mathbb{E}(y)) = \beta_0 + \beta_1 x_1 + \beta_2 x_2 + \ldots + \beta_p x_p$$
where $g(\ )$ is a link function and $y$ is from the exponential family.

\vfill\pause

For example \blue{logistic regression} for binary $y$:
$$\log\left(\frac{P(y = 1)}{1-P(y = 1)}\right) = \beta_0 + \beta_1 x_1 + \beta_2 x_2 + \ldots + \beta_p x_p$$

$\log\left(\frac{p}{1-p}\right)$ is the \blue{logit} link.

## Generalized Linear Regression (GLM)
To fit a \blue{GLM} in R we use the function \blue{\texttt{glm()}}.

\bigskip

The most important arguments are

\begin{itemize}\itemsep2mm
\item \blue{\texttt{formula}}:\newline a \texttt{formula} object
\item \blue{\texttt{family}}: \newline a \texttt{family} object or name of the family function,
   describing the error distribution and link function
\item \blue{\texttt{data}}: \newline a \texttt{data.frame} (optional, but usually needed)
\item \blue{\texttt{subset}}:\newline a \texttt{vector} specifying which observations should be used (optional)
\end{itemize}


## Families and Link Functions
Common families \& available links in R: \hfill (see also `?family`)
\begin{center}
\begin{tabular}{ll}
\toprule
family & link\\\midrule
\texttt{binomial} & \texttt{logit}, \texttt{probit}, \texttt{cauchit}, \texttt{log}, \texttt{cloglog}\\
\texttt{gaussian} & \texttt{identity}, \texttt{log}, \texttt{inverse}\\
\texttt{Gamma} & \texttt{inverse}, \texttt{identity}, \texttt{log}\\
\texttt{poisson} & \texttt{log}, \texttt{identity}, \texttt{sqrt}\\
\bottomrule
\end{tabular}
\end{center}

\pause
The `family` argument in `glm()` can be specified in the following ways:
\begin{columns}[onlytextwidth]
\begin{column}{0.5\linewidth}
\begin{itemize}
\item \texttt{binomial(link = "logit")}
\item \texttt{binomial()}
\item \texttt{binomial}
\item \texttt{"binomial"}
\end{itemize}
\end{column}
\pause
\begin{column}{0.5\linewidth}
\blue{Note:}\\
When the link is not explicitly specified (i.e. options 2-4),
the default link is used.
\end{column}
\end{columns}


## Regression

\begin{columns}[T,onlytextwidth]
\begin{column}{0.47\linewidth}
\begin{block}{Demo}
\begin{itemize}
\item Regression Basics\newline \button{https://nerler.github.io/BST02/demo/statistics-with-r/Regression_Basics.R}{R}
\button{https://nerler.github.io/BST02/demo/statistics-with-r/Regression_Basics.html}{html}
\end{itemize}
\end{block}
\end{column}

\begin{column}{0.47\linewidth}
\begin{block}{Practical}
\begin{itemize}
\item Linear Regression \button{https://nerler.github.io/BST02/practical/statistics-with-r/Linear_Regression.html}{html}
\end{itemize}
\end{block}
\end{column}
\end{columns}


## Model Evaluation

\blue{Linear model:}

Evaluate the \blue{assumptions} of a linear regression model visually, for example:

* Histogram of residuals
* Normal QQ-plot of residuals
* Scatterplot residuals vs fitted values

```{r, fig.width = 8, fig.height = 2, echo = FALSE, out.width = "100%"}
set.seed(2020)
N <- 2500
invS <- rWishart(1, df = 5, Sigma = diag(4))[, , 1]

x <- MASS::mvrnorm(N, rep(0, 4), solve(invS))
beta <- runif(ncol(x) + 1, -2, 2)

y <- cbind(1, x) %*% beta + rnorm(N, 0, 0.5)

lm_mod <- lm(y ~ x)

plotDF <- data.frame(resid = lm_mod$residuals,
                     fitted = lm_mod$fitted.values)


temp <- list(theme(panel.grid = element_blank(),
                   panel.background = element_rect(fill = grey(0.95)))
)

ggpubr::ggarrange(
  ggplot(plotDF, aes(x = resid)) +
    geom_histogram(bins = 50, fill = EMCdark, alpha = 0.5, color = EMCdark, size = 0.1) +
    xlab('Residuals') +
    temp,
  
  ggplot(data.frame(r = quantile(rstandard(lm_mod), seq(0, 1, length = 200)),
                    n = quantile(rnorm(N), seq(0, 1, length = 200))),
         aes(x = n, y = r)) +
    geom_abline(intercept = 0, slope = 1, color = drkred) +
    geom_point(color = EMCdark, alpha = 0.3, size = 1) +
    temp +
    xlab('theoretical quantiles') +
    ylab('observed quantiles') +
    scale_x_continuous(breaks = seq(-5, 5, 1)) +
    scale_y_continuous(breaks = seq(-5, 5, 1)),
  
  
    ggplot(plotDF, aes(x = fitted, y = resid)) +
    geom_point(color = EMCdark, alpha = 0.3, size = 1) +
    geom_smooth(method = 'loess', se = FALSE, color = drkred) +
    temp +
    xlab('fitted values') +
    ylab('residuals'),
ncol = 3, nrow = 1)
```



## Model Comparison

\blue{Nested models:}
* model is a \blue{special case} of the other, i.e.,
* model B is a special case of model A when B can be obtained by setting some 
  regression coefficients in A to zero
  
Comparison using a \blue{likelihood ratio (LR) test}, for example:
```{r, eval = FALSE}
anova(modelA, modelB)
anova(modelA, modelB, test = "LRT") # for a glm
```

\pause

\blue{Non-nested models:}
Comparison using information criteria, e.g.
```{r, eval = FALSE}
AIC(modelA, modelB)
BIC(modelA, modelB)
```
The model with the \blue{smaller} AIC (or BIC) has the \blue{better} fit.


## Model Evaluation

\begin{center}
\begin{columns}
\begin{column}{0.47\linewidth}
\begin{block}{Demo}
\begin{itemize}
\item Model Evaluation \button{https://nerler.github.io/BST02/demo/statistics-with-r/Model_Evaluation.R}{R} \button{https://nerler.github.io/BST02/demo/statistics-with-r/Model_Evaluation.html}{html}
\end{itemize}
\end{block}
\end{column}
\end{columns}
\end{center}



```{r, echo = FALSE}
set.seed(2020)
dat <- data.frame(x = runif(250, 0, 7))
dat$y <- with(dat, (0.5*x - 1)^2) + rnorm(250, 0, 0.8)
dat$z <- with(dat, x^(1/4)) + rnorm(250, 0, 0.1)

m1 <- lm(y ~ x, data = dat)
m2 <- lm(z ~ x, data = dat)

m1b <- lm(y ~ x + I(x^2), data = dat)

m2b <- lm(z ~ log(x), data = dat)

library(ggplot2)
ggtemp <- list(theme_bw(),
               theme(axis.title = element_text(size = 14),
                     axis.text = element_blank(),
                     axis.ticks = element_blank(),
                     panel.grid = element_blank(),
                     panel.border = element_rect(color = grey(0.8)))
)

p0list <- list(
  ggplot(dat, aes(x = x, y = y)) +
    ggtemp +
    xlab(expression(x)) +
    geom_point(alpha = 0.2) +
    theme(plot.title = element_text(margin = margin(t = 6, b = -20), hjust = 0.05)),
  
  ggplot(dat, aes(x = x, y = z)) +
    ggtemp +
    xlab(expression(x)) +
    geom_point(alpha = 0.2) +
    ylab('y') +
    theme(plot.title = element_text(margin = margin(t = 6, b = -20), hjust = 0.05))
)

p3list <- list(
    p0list[[1]] +
    geom_line(data = cbind(dat, fit = m1b$fit), aes(x = x, y = fit), size = 1, lty = 1, color = EMCdark) +
    geom_segment(aes(x = 0, xend = 1, y = coef(m1b) %*% c(1, 0, 0), yend = coef(m1b) %*% c(1, 0, 0)),
                 lty = 2) +
    geom_segment(aes(x = 1, xend = 1,
                     y = coef(m1b) %*% c(1, 0, 0), yend = coef(m1b) %*% c(1, 1, 1)),
                 arrow = arrow(ends = 'both', type = 'closed', length = unit(0.02, "npc"))) +
    geom_segment(aes(x = 5, xend = 6, y = coef(m1b) %*% c(1, 5, 5^2), yend = coef(m1b) %*% c(1, 5, 5^2)),
                 lty = 2) +
    geom_segment(aes(x = 6, xend = 6,
                     y = coef(m1b) %*% c(1, 5, 5^2), yend = coef(m1b) %*% c(1, 6, 6^2)),
                 arrow = arrow(ends = 'both', type = 'closed', length = unit(0.02, "npc"))),
  
  p0list[[2]] +
    geom_line(data = cbind(dat, fit = m2b$fit), aes(x = x, y = fit), size = 1, lty = 1, color = EMCdark) +
    geom_segment(aes(x = 0.05, xend = 1.05, y = coef(m2b) %*% c(1, log(0.05)), 
                     yend = coef(m2b) %*% c(1, log(0.05))),
                 lty = 2) +
    geom_segment(aes(x = 1.05, xend = 1.05,
                     y = coef(m2b) %*% c(1, log(0.05)), yend = coef(m2b) %*% c(1, log(1.05))),
                 arrow = arrow(ends = 'both', type = 'closed', length = unit(0.02, "npc"))) +
    geom_segment(aes(x = 5, xend = 6, y = coef(m2b) %*% c(1, log(5)), yend = coef(m2b) %*% c(1, log(5))),
                 lty = 2) +
    geom_segment(aes(x = 6, xend = 6,
                     y = coef(m2b) %*% c(1, log(5)), yend = coef(m2b) %*% c(1, log(6))),
                 arrow = arrow(ends = 'both', type = 'closed', length = unit(0.02, "npc")))
)
```

## Non-linear Effects
Default assumption: \blue{linear effect}, i.e.,\quad
x $\rightarrow$ y \quad \blue{\ding{225}} \quad x + 1 $\rightarrow$ y + $\beta$, \quad $\forall x$
\vfill
```{r, echo = FALSE, fig.width = 4, fig.height = 3.5, out.width = "60%", warning = FALSE}
set.seed(2020)
dat0 <- data.frame(x = runif(250, 0, 5))
dat0$y <- 0.3*dat0$x + rnorm(250, 0, 0.3)

m <- lm(y ~ x, data = dat0)

ggplot(dat0, aes(x = x, y = y)) +
  ggtemp +
  geom_point(alpha = 0.1) +
  geom_line(data = cbind(dat0, fit = m$fit), aes(y = fit), color = EMCdark, size = 1) +
  geom_segment(aes(x = 0, xend = 1, y = coef(m) %*% c(1, 0), yend = coef(m) %*% c(1, 0)),
               lty = 2) +
  geom_segment(aes(x = 1, xend = 1,
                   y = coef(m) %*% c(1, 0), yend = coef(m) %*% c(1, 1)),
               arrow = arrow(ends = 'both', type = 'closed', length = unit(0.02, "npc"))) +
  geom_segment(aes(x = 3, xend = 4, y = coef(m) %*% c(1, 3), yend = coef(m) %*% c(1, 3)),
               lty = 2) +
  geom_segment(aes(x = 4, xend = 4,
                   y = coef(m) %*% c(1, 3), yend = coef(m) %*% c(1, 4)),
               arrow = arrow(ends = 'both', type = 'closed', length = unit(0.02, "npc"))) +
  annotate('text', x = 1, y = coef(m) %*% c(1, 0.5), label = "beta", parse = TRUE,
           color = EMCdark, hjust = -0.5, size = 6) +
  annotate('text', x = 4, y = coef(m) %*% c(1, 3.5), label = "beta",
           color = EMCdark, hjust = -0.5, size = 6, parse = TRUE) +
  annotate('text', x = 0.5, y = coef(m) %*% c(1, 0), label = "1 unit",
           color = EMCdark, vjust = 1.1, size = 4) +
  annotate('text', x = 3.5, y = coef(m) %*% c(1, 3), label = "1 unit",
           color = EMCdark, vjust = 1.1, size = 4)
```


## Non-linear Effects
Default assumption: \blue{linear effect}, i.e.,\quad
x $\rightarrow$ y \quad \blue{\ding{225}} \quad x + 1 $\rightarrow$ y + $\beta$, \quad $\forall x$

\vfill

This may not always be the case:
```{r, echo = FALSE, fig.width = 8, fig.height = 3.5, out.width = "90%"}
cowplot::plot_grid(
  p0list[[1]] +
    geom_line(data = cbind(dat, fit = m1$fit), aes(x = x, y = fit), size = 1, color = EMCdark),
  p0list[[2]] +
    geom_line(data = cbind(dat, fit = m2$fit), aes(x = x, y = fit), size = 1, color = EMCdark)
)
```

## Non-linear Effects
Here, we would like to allow the \blue{effect} of a one-unit increase \blue{of x} to
\blue{change with the value of x}:

\vfill

```{r p3list, echo = FALSE, fig.width = 8, fig.height = 3.5, out.width = "90%"}
cowplot::plot_grid(plotlist = p3list)
```

## Non-linear Effects
\begincols[t, onlytextwidth]
\begincol{0.5\linewidth}
Sometimes, we can use
\begin{itemize}
\item a \blue{transformation of x}, or
\item x as well as \blue{a polynomial} of x (or a transformation).
\end{itemize}
\bigskip

For example:
```{r, eval = FALSE}
y ~ x + I(x^2)
```
or
```{r, eval = FALSE}
y ~ log(x)
```
\endcol
\begincol{0.5\linewidth}
```{r, echo = FALSE, fig.width = 4, fig.height = 5, out.width = "80%"}
cowplot::plot_grid(
  p0list[[1]] +
    geom_line(data = cbind(dat, fit = m1b$fit), aes(x = x, y = fit), size = 1, lty = 1, color = EMCdark),
  
  p0list[[2]] +
    geom_line(data = cbind(dat, fit = m2b$fit), aes(x = x, y = fit), size = 1, lty = 1, color = EMCdark),
  ncol = 1)
```
\endcol
\endcols

## Non-linear Effects: \texttt{I()}

The function \blue{\texttt{I()}} is needed to distinguish between operators
that need to be interpreted as 

* arithmetic operators and 
* formula operators

\bigskip\pause

Example:
```{r, eval = FALSE}
y ~ I(a + b)
```
\vspace*{-2ex}would be the same as
```{r, eval = FALSE}
z <- a + b
y ~ z
```
\vspace*{-2ex}but not the same as
```{r, eval = FALSE}
y ~ a + b
```



```{r prepsplines, echo = FALSE}
tpower <- function(x, t, p)
  # Function for truncated p-th power function
  (x - t) ^ p * (x > t)

bbase <- function(x, ndx, deg){
  xl <- min(x, na.rm = TRUE)
  xr <- max(x, na.rm = TRUE)
  
  # Function for B-spline basis
  dx <- (xr - xl) / ndx
  # knots <- seq(xl - deg * dx, xr + deg * dx, by = dx)
  knots <- knots(x, ndx, deg)
  P <- outer(x, knots, tpower, deg)
  n <- dim(P)[2]
  D <- diff(diag(n), diff = deg + 1) / (gamma(deg + 1) * dx ^ deg)
  B <- (-1) ^ (deg + 1) * P %*% t(D)
  B 
}

knots <- function(x, ndx, deg){
  xl <- min(x, na.rm = TRUE)
  xr <- max(x, na.rm = TRUE)
  dx <- (xr - xl) / ndx
  seq(xl - deg * dx, xr + deg * dx, by = dx)
}

set.seed(2019)
dat <- data.frame(x = sort(runif(250, min = 0, max = 2)))
dat$y <- 4 + sin(5*dat$x) + 1*dat$x^3 - 2 * dat$x^2 + rnorm(length(dat$x), 0, 0.5)

kn <- knots(dat$x, 6, deg = 3)
B <- bbase(dat$x, 6, deg = 3)

mod <- lm(dat$y ~ B - 1)
dat$fit <- mod$fitted.values

dat2 <- reshape2::melt(cbind(dat, B*2), id.vars = c('x', 'y', 'fit'))
dat3 <- reshape2::melt(cbind(dat, B * matrix(nrow = nrow(B), ncol = ncol(B),
                                data = rep(coef(mod), nrow(B)), byrow = TRUE)),
                       id.vars = c('x', 'y', 'fit'))

p0 <- ggplot(dat, aes(x = x, y = y)) +
  ggtemp +
  xlab("x") +
  geom_point(alpha = 0.3)
```


## Complex Non-linear Effects
\begincols[t, onlytextwidth]
\begincol{0.35\linewidth}
Non-linear effects may be \blue{more complex} than can be modelled with a simple
transformation or polynomial.
\endcol
\begincol{0.65\linewidth}
```{r p0, echo = FALSE, fig.width = 7, fig.height = 4}
p0
```
\endcol
\endcols
\blue{Also:} the shape may depend on other covariates in the model\newline
\blue{\ding{225}} we do not always know the shape in advance\newline
\blue{\ding{225} Regression Splines} / \blue{B-Splines}



## B-Splines
A \blue{B-Spline} is a linear combination of a set of \blue{basis functions}.

These basis functions are defined so that they are \vspace*{-2ex}

* a \blue{polynomial functions} inside a given interval, and
* zero outside that interval.

The intervals are defined by a set of \blue{knots}.

The polynomial function have a certain \blue{degree}
(i.e., constant, linear, quadratic, ...)

\bigskip\pause

\blue{B-Splines in R}
The R package \blue{splines} provides the functions

* \blue{\texttt{bs()}}: B-splines
* \blue{\texttt{ns()}}: natural cubic (B-)splines

## B-Splines
Instead of \quad $\displaystyle y \sim \beta_0  + \beta_1 x + \ldots$ \quad we assume \quad
$\displaystyle y \sim \beta_0 + \sum_{\ell = 1}^d \beta_\ell {\color{darkred}B_\ell (x)} + \ldots$
```{r p1, echo = FALSE, fig.width = 8, fig.height = 4, out.width = "90%"}
p0 + 
  geom_line(data = dat2, aes(y = value, group = variable), color = drkred)
```

## B-Splines
Instead of \quad $y \sim \beta_0  + \beta_1 x + \ldots$ \quad we assume \quad
$\displaystyle y \sim \beta_0 + \sum_{\ell = 1}^d {\color{darkred}\beta_\ell} B_\ell (x) + \ldots$
```{r p2, echo = FALSE, fig.width = 8, fig.height = 4, out.width = "90%"}
p0 + 
  geom_line(data = dat3, aes(y = value, group = variable), color = drkred)
```

## B-Splines
Instead of \quad $y \sim \beta_0  + \beta_1 x + \ldots$ \quad we assume \quad
$\displaystyle y \sim \beta_0 + {\color{darkred}\sum_{\ell = 1}^d} \beta_\ell B_\ell (x) + \ldots$
```{r, echo = FALSE, fig.width = 8, fig.height = 4, out.width = "90%"}
p0 + 
  geom_line(data = dat3, aes(y = value, group = variable)) +
  geom_line(aes(y = fit), size = 1, col = drkred)
```


## B-Splines: \texttt{degree}
```{r, echo = FALSE, fig.width = 8, fig.height = 4.5}
plist <- list()

for (dg in 0:3) {
  set.seed(2019)
  dat <- data.frame(x = sort(runif(250, min = 0, max = 2)))
  dat$y <- 4 + sin(5*dat$x) + 1*dat$x^3 - 2 * dat$x^2 + rnorm(length(dat$x), 0, 0.5)

  kn <- knots(dat$x, 6, deg = dg)
  B <- bbase(dat$x, 6, deg = dg)
  
  mod <- lm(dat$y ~ B - 1)
  dat$fit <- mod$fitted.values
  dat$fit[1] <- NA

  dat2 <- reshape2::melt(cbind(dat, B * 1), id.vars = c('x', 'y', 'fit'))

  plist[[dg + 1]] <- ggplot(dat, aes(x = x, y = y)) +
    ggtemp +
    xlab("x") +
    geom_point(alpha = 0.3) +
    geom_line(data = dat2, aes(y = value, group = variable), color = EMCdark) +
    geom_line(aes(y = fit), size = 1, color = EMCdark, na.rm = TRUE) +
    ggtitle(paste0("degree = ", dg)) +
    theme(plot.title = element_text(hjust = 0.5, margin = margin(t = 10, b = -20),
                                    family = "mono"),
          plot.margin = margin(l = 5, t = 5, b = 5, r = 5, unit = "pt"),
          axis.title = element_blank())
}

cowplot::plot_grid(plotlist = plist)
```


```{r, echo = FALSE}
plist_NS <- plist_BS <- list()

dfs <- c(1, 2, 3, 5, 10, 30)
for (k in seq_along(dfs)) {
  df <- dfs[k]
  
  set.seed(2019)
  dat <- data.frame(x = sort(runif(250, min = 0, max = 2)))
  dat$y <- 4 + sin(5*dat$x) + 1*dat$x^3 - 2 * dat$x^2 + rnorm(length(dat$x), 0, 0.5)
  
  mod_ns <- lm(y ~ splines::ns(x, df = df), data = dat)
  dat$fit_ns <- mod_ns$fitted.values
  NS <- splines::ns(dat$x, df = df)
  if (df >= 3) {
    mod_bs <- lm(y ~ splines::bs(x, df = df), data = dat)
    dat$fit_bs <- mod_bs$fitted.values
    BS <- splines::bs(dat$x, df = df)
    dat_BS <- reshape2::melt(cbind(dat, BS), id.vars = c('x', 'y', 'fit_bs', 'fit_ns'))
  } else {
    dat$fit_bs <- NA
  }
  
  dat_NS <- reshape2::melt(cbind(dat, NS), id.vars = c('x', 'y', 'fit_bs', 'fit_ns'))

  p <- ggplot(dat, aes(x = x, y = y)) +
    ggtemp +
    geom_point(alpha = 0.3) +
    theme(plot.title = element_text(hjust = 0.5, margin = margin(t = 10, b = -15),
                                    family = "mono"),
          plot.margin = margin(l = 5, t = 5, b = 5, r = 5, unit = "pt"),
          axis.title = element_blank())

  
  plist_NS[[k]] <- p +
    geom_vline(xintercept = attr(NS, 'knots'), lty = 3, color = grey(0.7)) +
    geom_vline(xintercept = attr(NS, 'Boundary.knots'), lty = 2, color = grey(0.7)) +
    geom_line(data = dat_NS, aes(y = value, group = variable), color = EMCdark) +
    geom_line(aes(y = fit_ns), size = 1, color = EMCdark) +
    ggtitle(paste0("ns(x, df = ", df, ")"))
  
  if (df >= 3) {
    plist_BS[[k]] <- p +
    geom_vline(xintercept = attr(BS, 'knots'), lty = 3, color = grey(0.7)) +
    geom_vline(xintercept = attr(BS, 'Boundary.knots'), lty = 2, color = grey(0.7)) +
      geom_line(data = dat_BS, aes(y = value, group = variable), color = EMCdark) +
      geom_line(aes(y = fit_bs), size = 1, color = EMCdark) +
      ggtitle(paste0("bs(x, df = ", df, ")"))
  }
}
```

## B-Splines: \texttt{df}
```{r, echo = FALSE, fig.width = 8, fig.height = 4.3, cache = TRUE}
cowplot::plot_grid(plotlist = plist_NS)
```
\vspace*{-3ex}

## B-Splines: \texttt{df}
```{r, echo = FALSE, fig.width = 8, fig.height = 4.3, cache = TRUE}
cowplot::plot_grid(plotlist = plist_BS)
```
\vspace*{-3ex}


## B-Splines in R: \texttt{bs()} & \texttt{ns()}
Important arguments of `ns()` and `bs()` are:

\begincols[t, onlytextwidth]
\begincol{0.5\linewidth}

\blue{\texttt{degree}}

* degree of the polynomial in each of the basis functions
* in \texttt{bs()}: default is 3
* in \texttt{ns()}: always 3 ("cubic")

\bigskip\pause

\blue{\texttt{df}} 

* degrees of freedom, i.e., "number of regression coefficients" used
* for \texttt{bs()}: has to be $\geq$ \texttt{degree}

\endcol\pause
\begincol{0.5\linewidth}

\blue{\texttt{knots}}

* position of (inner) knots
* if unspecified:
  * \texttt{df-degree} knots are used
  * positioned at equally spaced quantiles

\bigskip\pause

\blue{\texttt{Boundary.knots}}

* by default: \texttt{range(x)}
* outside the \texttt{Boundary.knots} the fit is extrapolated

\endcol
\endcols

```{r, include = FALSE, eval= FALSE}
boot::urine
MASS::birthwt
MASS::Boston # Housing Values
MASS::Cars93
MASS::crabs
MASS::Pima.te
MASS::survey???
MASS::UScereal
MASS::UScrime

rpart::stagec
rpart::car90

survival::pbc
```



## Non-linear Effects

\begin{columns}[T,onlytextwidth]
\begin{column}{0.37\linewidth}
\begin{block}{Demo}
\begin{itemize}
\item Splines \button{https://nerler.github.io/BST02/demo/statistics-with-r/}{R}
\button{https://nerler.github.io/BST02/demo/statistics-with-r/}{html}
\end{itemize}
\end{block}
\end{column}

\begin{column}{0.57\linewidth}
\begin{block}{Practicals}
\begin{itemize}
\item Logistic Regression \& More \button{https://nerler.github.io/BST02/practical/statistics-with-r/Logistic_Regression_and_More.html}{html}
\item Logistic Regression II \button{https://nerler.github.io/BST02/practical/statistics-with-r/Logistic_Regression_II.html}{html}
\item Custom Model Summary Function \button{https://nerler.github.io/BST02/practical/statistics-with-r/Model_Summary_Function.html}{html}
\end{itemize}
\end{block}
\end{column}
\end{columns}

## Regression
\begin{columns}[onlytextwidth,T]
\begin{column}{0.5\linewidth}

\blue{Regression Models}
\begin{itemize}
\item \texttt{lm()}
\item \texttt{glm()}
\end{itemize}

\bigskip

\blue{Regression Results}
\begin{itemize}\vspace*{-0.5ex}
\item \texttt{summary()}
\item \texttt{coef()}, \texttt{confint()}
\item \texttt{fitted()}, \texttt{residuals()}, \texttt{rstandard()}
\item \texttt{AIC()}, \texttt{BIC()}
\item \texttt{anova()}
\end{itemize}
\end{column}

\begin{column}{0.5\linewidth}
\blue{Plots}
\begin{itemize}\vspace*{-0.5ex}
\item \texttt{plot()}
\item \texttt{qqnorm()}, \texttt{qqline()}, \texttt{qqplot()}
\end{itemize}

\bigskip

\blue{Formulas}
\begin{itemize}\vspace*{-0.5ex}
\item Formula operators: \texttt{+}, \texttt{-}, \texttt{*}, \texttt{:}, \text{\texttt{\^}}
\item \texttt{ns()}, \texttt{bs()}, \texttt{I()}
\item \texttt{all.vars()}
\item \texttt{update()}
\item \texttt{as.formula()}
\end{itemize}

\end{column}
\end{columns}
