---
title: 'Control Flow and Functions'
output: 
  html_document:
    code_folding: show
    df_print: paged
    toc: true
    toc_float:
      collapsed: false
    number_sections: false
    theme: spacelab
    highlight: tango
    includes:
      after_body: ../footerNE.html
    css: ../style.css
---


```{r setup, include = FALSE}
library(knitr)
library(kableExtra)

opts_chunk$set(echo = TRUE)

hint <- function(text) {
  id <- paste0('hint_', sample.int(1e10, size = 1))
    cat(
    paste0('\n',
      '<button type="button" class="btn btn-info btn-sm" data-toggle="collapse" ',
      'data-target="#', id, '">Hint</button>',
      '<div id = "', id, '" class="collapse" style="border:1px; ',
      'border-style:solid; padding: 1em; border-color:#1F78B4">',
      text, '</div>')
  )
}
```

## Data
For this practical, we simulate data using the following syntax:
```{r}
set.seed(2020)
nrj <- data.frame(sex = factor(sample(c('male', 'female'), size = 100, replace = TRUE)),
                  kcal = round(rnorm(100, mean = 2200, sd = 250)),
                  weight = runif(100, min = 45, max = 150),
                  height = rnorm(100, mean = 170, sd = 10),
                  age = rnorm(100, mean = 50, sd = 10),
                  sports = factor(sample(c('never', 'sometimes', 'regularly'),
                                         size = 100, replace = TRUE), ordered = TRUE)
)
```

The first six rows of the resulting `data.frame` are:
```{r, echo = FALSE}
head(nrj) %>% 
  kable(digits = 2) %>% 
  kable_styling(full_width = FALSE)

```


## Calculating the BMR {.tabset .tabset-pills .tabset-fade}
We want to calculate the Basal Metabolic Rate (BMR) for the individuals in the
**nrj** data.

The formula differs for men and women:

* men: $(13.75 \times \text{weight}) + (5 \times \text{height}) - (6.76 \times \text{age}) + 66$
* women: $(9.56 \times \text{weight}) + (1.85 \times \text{height}) - (4.68 \times \text{age}) + 655$

### Task 1
<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">
Which function would you choose to calculate the BMR? `ifelse()` or `if() ... else ...`?
</div>

### Solution 1
Both are possible, but `ifelse()` is more straightforward in this setting (in my opinion).

### Task 2
<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">
Calculate the BMR using `ifelse()` and add it to the dataset as a new variable `BMR1`.

```{r, results = 'asis', echo = FALSE}
hint('The function `ifelse()` has arguments `test`, `yes` and `no`.')
hint('You could use `nrj$sex == "male"` as the `test`.')
```

</div>

### Solution 2
```{r}
nrj$BMR1 <- ifelse(nrj$sex == 'male',
                  13.75 * nrj$weight + 5 * nrj$height − 6.76 * nrj$age + 66,
                  9.56 * nrj$weight + 1.85 * nrj$height − 4.68 * nrj$age + 655
)

head(nrj)
```

### Task 3
We now want to calculate the BMR using `if() ... else ...`.
<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">
How is the `condition` used in `if()` different from the `test` used in `ifelse()`?
</div>

### Solution 3
The argument `test` in `ifelse()` expects a vector of "tests" (a vector of `TRUE` and `FALSE`)
while the `condition` argument in `if()` expects a single "test" (a single `TRUE` or `FALSE`).

### Task 4
<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">
How can we check row by row if a subject is male or female?
</div>

### Solution 4
We could use a `for()`-loop that runs through all rows in **nrj**.

### Task  5
<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">
Calculate the BMR using `if() ... else ...` and add it to the **nrj** data as
a new variable `BMR2`.
```{r, echo = FALSE, results = 'asis'}
hint('You need a sequence from 1 to the number of rows of **nrj**.')
hint('You need to pre-specify (an empty version of) the new variable `BMR2`.')
hint("Don't forget that you now need to use indices.")
```
</div>

### Solution 5
```{r}
nrj$BMR2 <- NA
for (i in 1:nrow(nrj)) {
  nrj$BMR2[i] <- if (nrj$sex[i] == 'male') {
    13.75 * nrj$weight[i] + 5 * nrj$height[i] − 6.76 * nrj$age[i] + 66
  } else {
    9.56 * nrj$weight[i] + 1.85 * nrj$height[i] − 4.68 * nrj$age[i] + 655
  }
}
```

### Task 6
<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">
Check that `BMR1` and `BMR2` are the same.
</div>

### Solution 6
There are multiple possible ways to check this:
```{r}
all.equal(nrj$BMR1, nrj$BMR2)
identical(nrj$BMR1, nrj$BMR2)
table(nrj$BMR1 == nrj$BMR2)
summary(nrj$BMR1 - nrj$BMR2)
```


## Data Summary {.tabset .tabset-pills .tabset-fade}
We would now like to get descriptive statistics of the **nrj** data.
For continuous variables, we want the mean and standard deviation, for 
categorical data we want to report the proportion of subjects in each category.

### Task 1
Remember how to get summary measures?

<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">

* Get the mean and standard deviation of the variable `kcal`.
* Get the proportion of observations in each category for the variable `sports`.
```{r, echo = FALSE, results = 'asis'}
hint('You might use `mean()`, `sd()`, `table()` and `prop.table()`.')
```

</div>

### Solution 1
```{r}
mean(nrj$kcal)
sd(nrj$kcal)

prop.table(table(nrj$sports))
```

### Task 2
<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">
Use `paste()` to create some nice-looking output for these summaries:

* "kcal: mean (sd)" for `kcal`
* "sports: percentage% category 1, percentage% category2, ..."

```{r, echo = FALSE, results = 'asis'}
hint("Remember: The function `paste()` has arguments `sep` and `collapse`, but you could also use `paste0()`.")
hint("You might also want to use the function `round()` to reduce the number of
     digits in the output.")
```

</div>

### Solution 2
```{r}
# for "kcal":
paste0("kcal: ", round(mean(nrj$kcal), 1), " (", round(sd(nrj$kcal), 1), ")")


# for "sports":
# table of proportions:
tab <- prop.table(table(nrj$sports))
# combine proportions with category names:
props <- paste0(tab * 100, "% ", names(tab), collapse = ", ")
# combine variable name with proportions string:
paste('sports:', props)

```

### Task 3
<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">
Write a loop that creates the summary strings we created in the previous Solution
for each variable in the **nrj** data.

Use `print()` to print each of the summaries.

```{r, echo = FALSE, results = 'asis'}
hint("You need a loop that iterates through the columns of **nrj**.")
hint("You need to autmatically check for each variable if it is a `factor()` or not
     to choose the correct summary type.")
```
</div>

### Solution 3
```{r}
for (i in 1:ncol(nrj)) {
  if (is.factor(nrj[, i])) {
    tab <- prop.table(table(nrj[, i]))
    props <- paste0(tab * 100, "% ", names(tab), collapse = ", ")
    paste0(names(nrj)[i], ": ", props)
  } else {
    print(paste0(names(nrj)[i], ": ", 
                 round(mean(nrj[, i]), 1), " (", round(sd(nrj[, i]), 1), ")"))
  }
}
```

## Writing your own function {.tabset .tabset-pills .tabset-fade}
We also want to practice writing our own functions.

### Task 1
<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">
Write two functions:

* one that creates the summary string for a continuous variable,
* one that creates the output for a categorical variable.

Assume that the input will be a vector, such as `nrj$kcal` or `nrj$sports`.
Note that when we work with these vectors, they do not contain the variable
name any more, so we cannot use the variable name in the two functions.

```{r, echo = FALSE, results = 'asis'}
hint("The general form of a function is `function(){...}`.")
```
</div>

### Solution 1
We can just re-use the syntax we used before, and only remove the part that 
adds the variable name.

Our functions have one argument, which we call `x` here:
```{r}
summary_continuous <- function(x) {
  paste0(round(mean(x), 1), " (", round(sd(x), 1), ")")
}

summary_continuous(nrj$kcal)

summary_categorical <- function(x) {
  tab <- prop.table(table(x))
  paste0(tab * 100, "% ", names(tab), collapse = ", ")
}

summary_categorical(nrj$sports)
```

### Task 2
<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">
Write another function that has a `data.frame` as input and prints the 
summary string for each variable, using the two functions from the previous solution.
</div>

### Solution 2
This function again has one argument which we call `dat`. We can again re-use
syntax from above:
```{r}
summary_df <- function(dat) {
  for (i in 1:ncol(dat)) {
    summary_string <- if (is.factor(dat[, i])) {
      summary_categorical(dat[, i])
    } else {
      summary_continuous(dat[, i])
    }
    print(summary_string)
  }
}

summary_df(dat = nrj)
```

Note that it would also be possible to use `print()` directly around the
functions `summary_categorical()` and `summary_continuous()`, but for the 
next Task it is more convenient to first collect the summary string in an
object (`summary_string`), and do the further steps with that object.


### Task 3
The function `summary_df()` that we created in the previous solution does not
contain any variable names.
<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">
Modify the function so that the output strings look like in the previous exercise 
(with "variable name: ...").
</div>

### Solution 3
To adjust the function, the only row that needs changing is the one in which
we print the summary:
```{r}
summary_df <- function(dat) {
  for (i in 1:ncol(dat)) {
    summary_string <- if (is.factor(dat[, i])) {
      summary_categorical(dat[, i])
    } else {
      summary_continuous(dat[, i])
    }
    print(paste0(names(dat)[i], ": ", summary_string))
  }
}

summary_df(dat = nrj)
```

