---
title: 'A Custom Subset Function'
output: 
  html_document:
    df_print: paged
    toc: true
    toc_float:
      collapsed: false
    number_sections: false
    theme: spacelab
    highlight: tango
    includes:
      after_body: ../footerNE.html
    css: ../style.css
---


```{r setup, include = FALSE}
library(knitr)
library(kableExtra)
options(knitr.kable.NA = '')


opts_chunk$set(echo = TRUE)

hint <- function(text) {
  id <- paste0('hint_', sample.int(1e10, size = 1))
    cat(
    paste0('\n',
      '<button type="button" class="btn btn-info btn-sm" data-toggle="collapse" ',
      'data-target="#', id, '">Hint</button>',
      '<div id = "', id, '" class="collapse" style="border:1px; ',
      'border-style:solid; padding: 1em; border-color:#1F78B4">',
      text, '</div>')
  )
}

set.seed(2020)
```


## Selection based on an interval {.tabset .tabset-pills .tabset-fade}

Aim of this practical is to write a function that selects a subset of 
a dataset based on whether a particular variable is inside an interval.

<div style="border:2px; border-style:solid; padding: 1em; border-color:#009B8F">
<p style = "color:#009B8F; font-weight:bold;">Note:</p>
When writing functions, there are usually many ways that lead to the same solution.
The solutions given here are suggestions, but there are many other ways the
functions could be written.
</div>
<br>


### Task 1
<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">
* How many arguments does our function need? 
* What type of input do you need for these arguments?
</div>


### Solution 1
The function needs three arguments. We will call them

* `dat`: `a data.frame`
* `variable`: the name of the variable based on which the subset is created (a character string)
* `interval`: a numeric vector of length two, with the lower and upper limit of the interval


### Task 2
<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">

Now, write the function.

* It should create a filter variable that tells us for each row of the
  dataset `dat` if the value of `variable` is inside the `interval` or not.
* Using this filter variable, create the subset and return it.
* Create your own example `data.frame` to test the function.

</div>


### Solution 2

```{r}
fun1 <- function(dat, variable, interval) {
  filter <- dat[, variable] > interval[1] & dat[, variable] < interval[2]
  
  subset(dat, subset = filter)
}
```

To test the function, we create a dataset that allows us to check easily
if what the function returns is correct:
```{r}
exdat <- data.frame(a = 1:10,
                    b = sample(c('A', 'B'), size = 10, replace = TRUE)
)
```
```{r, eval = FALSE}
exdat
```
```{r, echo = FALSE}
cat(paste0(capture.output(exdat), collapse = "\n"))
```
For `interval = c(3, 7)` the output should be those rows with `a` equal to 4, 5, or 6:
```{r, eval = FALSE}
fun1(exdat, variable = 'a', interval = c(3, 7))
```
```{r, echo = FALSE}
cat(paste0(capture.output(
  fun1(exdat, variable = 'a', interval = c(3, 7))
), collapse = "\n"))
```


### Task 3
<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">
Extend the function with an argument `incl_boundaries` that allows the user to 
specify whether the boundaries of the `interval` should be included in the 
subset that is returned.

* Specify this argument so that by default the boundaries are always included.
* Check your function using the example data.
</div>

### Solution 3
```{r}
fun2 <- function(dat, variable, interval, incl_boundaries = TRUE) {
  
  # check if boundaries should be included
  filter <- if (incl_boundaries) {
    
    # syntax for subset including boundaries
    dat[, variable] >= interval[1] & dat[, variable] <= interval[2]
    
  } else {
    
    # syntax for subset not including boundaries
    dat[, variable] > interval[1] & dat[, variable] < interval[2]
  }
  
  subset(dat, subset = filter)
}
```

With the default setting to include the boundaries, `fun2()` should return the
rows of `exdat` where `a` is 3, ..., 6:
```{r, eval = FALSE}
fun2(exdat, variable = 'a', interval = c(3, 6))
```
```{r, echo = FALSE}
cat(paste0(capture.output(fun2(exdat, variable = 'a', interval = c(3, 6))), collapse = "\n"))
```

When we set `incl_boundaries = FALSE` the rows where `a = 3` and `a = 6`
should be excluded:
```{r, eval = FALSE}
fun2(exdat, variable = 'a', interval = c(3, 6), incl_boundaries = FALSE)
```
```{r, echo = FALSE}
cat(paste0(capture.output(
  fun2(exdat, variable = 'a', interval = c(3, 6), incl_boundaries = FALSE)
), collapse = "\n"))

```



## Inside or outside? {.tabset .tabset-pills .tabset-fade}

### Task 1
We want to extend the function further with an additional argument `outside`
that allows the user to choose whether cases with values of the `variable` inside
or outside the specified interval should be selected.

<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">
* Write this extended version of the function using another `if() ... else` statement.
* Set the argument so that by default values inside the interval are selected.
</div>

### Solution 1
```{r}
fun3 <- function(dat, variable, interval, incl_boundaries = TRUE, outside = FALSE) {
  
  # check if boundaries should be included
  filter <- if (incl_boundaries) {
    
    # check if values outside (or inside) the interval should be selected
    if (outside) {
      # syntax for subset including boundaries
      dat[, variable] <= interval[1] | dat[, variable] >= interval[2]
    } else {
      # syntax for subset including boundaries
      dat[, variable] >= interval[1] & dat[, variable] <= interval[2]
    }
    
  } else {
    
    # check if values outside (or inside) the interval should be selected
    if (outside) {
      # syntax for subset not including boundaries
      dat[, variable] < interval[1] | dat[, variable] > interval[2]
    } else {
      # syntax for subset not including boundaries
      dat[, variable] > interval[1] & dat[, variable] < interval[2]
    }
  }
  
  subset(dat, subset = filter)
}
```

We test all four scenarios (different combinations of `incl_boundaries` and `outside`)
with `interval = c(3, 7)`:

`incl_boundaries = TRUE, outside = FALSE` should include values 3, ..., 7:
```{r, eval = FALSE}
fun3(exdat, variable = 'a', interval = c(3, 7), incl_boundaries = TRUE, outside = FALSE)
```
```{r, echo = FALSE}
cat(paste0(capture.output(
  fun3(exdat, variable = 'a', interval = c(3, 7), incl_boundaries = TRUE, outside = FALSE)
), collapse = "\n"))
```


`incl_boundaries = TRUE, outside = TRUE` should include values 1, 2, 3, and 7, ..., 10:
```{r, eval = FALSE}
fun3(exdat, variable = 'a', interval = c(3, 7), incl_boundaries = TRUE, outside = TRUE)
```
```{r, echo = FALSE}
cat(paste0(capture.output(
fun3(exdat, variable = 'a', interval = c(3, 7), incl_boundaries = TRUE, outside = TRUE)
), collapse = "\n"))
```


`incl_boundaries = FALSE, outside = FALSE` should include values 4, 5, 6:
```{r, eval = FALSE}
fun3(exdat, variable = 'a', interval = c(3, 7), incl_boundaries = FALSE, outside = FALSE)
```
```{r, echo = FALSE}
cat(paste0(capture.output(
  fun3(exdat, variable = 'a', interval = c(3, 7), incl_boundaries = FALSE, outside = FALSE)
), collapse = "\n"))
```


`incl_boundaries = FALSE, outside = TRUE` should include values 1, 2 and 8, 9, 10:
```{r, eval = FALSE}
fun3(exdat, variable = 'a', interval = c(3, 7), incl_boundaries = FALSE, outside = TRUE)
```
```{r, echo = FALSE}
cat(paste0(capture.output(
  fun3(exdat, variable = 'a', interval = c(3, 7), incl_boundaries = FALSE, outside = TRUE)
), collapse = "\n"))
```




### Task 2
In `fun3()`, many of the lines of syntax are almost identical.

<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">
Re-write the function to make better use of the fact that a `logical` value can be
reversed (i.e., you can use the `filter` to specify which cases should be selected
or which cases should be excluded.)

```{r, echo = FALSE, results = 'asis'}
hint('The idea is that we could use `filter` when we want to select cases with
values inside the interval and `!filter` when we want to select cases outside
the interval.')
hint('It is helpful to make a table of the combinations of options for `incl_boundaries`
and `outside` and then think about which version of the syntax is needed:

* version 1) `dat[, variable] >= interval[1] & dat[, variable] <= interval[2]`
* version 2) `dat[, variable] > interval[1] & dat[, variable] < interval[2]`')
```

</div>

### Solution 2

Our first idea might be to try this function:
```{r, eval = FALSE}
fun4_false <- function(dat, variable, interval, incl_boundaries = TRUE, outside = FALSE) {
  
  filter <- if (incl_boundaries) {
    dat[, variable] >= interval[1] & dat[, variable] <= interval[2]  # version 1
  } else {
    dat[, variable] > interval[1] & dat[, variable] < interval[2]    # version 2
  }
  
  if (outside) 
    filter <- !filter
  
  subset(dat, filter)
}
```

* Depending on the `incl_boundary` argument, the `filter` selects either the inside
  of the interval, or the inside AND the boundaries. 
* Depending on the argument `outside`, we then use `filter` directly (when we want
  to return the inside of the interval) or negate it (i.e., turn `TRUE` into `FALSE`
  and vice versa) when `outside = TRUE`.

The problem is that by negating the filter, we also "negate" the in- or 
exclusion of the boundary values:

* when `filter` includes the boundary values, they are excluded in `!filter`
* when `filter` excludes the boundary values, they are included in `!filter`

We somehow need to consider the two arguments `incl_boundary` and `outside`
together.
This is where the table comes in handy:
```{r, echo = FALSE}
tab <- expand.grid(incl_boundary = c(TRUE, FALSE),
                   outside = c(TRUE, FALSE),
                   version = NA)
kable(tab) %>%
  kable_styling(full_width = FALSE)
```

We can then fill in the table by looking at each scenario:

* If we want the outside of the intervals including the boundaries, we need to
  exclude the inside of the interval without boundaries => version 2
* If we want the outside of the intervals excluding the boundaries, we need to
  exclude the inside of the interval inclusive the boundaries => version 2
* If we want the inside of the interval including the boundaries, we select 
  exactly that => version 1
* If we want the inside of the interval excluding the boundaries, we select 
  exactly that => version 2

The filled-in version of the table then is:
```{r, echo = FALSE}
tab$version <- c(2, 1, 1, 2)

kable(tab) %>%
  kable_styling(full_width = FALSE)
```

We note that we need to choose version 1 when `incl_boundary != outside` and
version 2 when `incl_boundary == outside`!

With this we can fix our function, which now is a lot shorter than `fun3`:
```{r}
fun4_correct <- function(dat, variable, interval, incl_boundaries = TRUE, outside = FALSE) {
  
  filter <- if (incl_boundaries != outside) {
    dat[, variable] >= interval[1] & dat[, variable] <= interval[2]
  } else {
    dat[, variable] > interval[1] & dat[, variable] < interval[2]
  }
  
  if (outside) 
    filter <- !filter
  
  subset(dat, filter)
}
```


We repeat the same set of tests as before, with `interval = c(3, 7)`:

`incl_boundaries = TRUE, outside = FALSE` should include values 3, ..., 7:
```{r, eval = FALSE}
fun4_correct(exdat, variable = 'a', interval = c(3, 7), incl_boundaries = TRUE, outside = FALSE)
```
```{r, echo = FALSE}
cat(paste0(capture.output(
  fun4_correct(exdat, variable = 'a', interval = c(3, 7), incl_boundaries = TRUE, outside = FALSE)
), collapse = "\n"))
```


`incl_boundaries = TRUE, outside = TRUE` should include values 1, 2, 3, and 7, ..., 10:
```{r, eval = FALSE}
fun4_correct(exdat, variable = 'a', interval = c(3, 7), incl_boundaries = TRUE, outside = TRUE)
```
```{r, echo = FALSE}
cat(paste0(capture.output(
fun4_correct(exdat, variable = 'a', interval = c(3, 7), incl_boundaries = TRUE, outside = TRUE)
), collapse = "\n"))
```


`incl_boundaries = FALSE, outside = FALSE` should include values 4, 5, 6:
```{r, eval = FALSE}
fun4_correct(exdat, variable = 'a', interval = c(3, 7), incl_boundaries = FALSE, outside = FALSE)
```
```{r, echo = FALSE}
cat(paste0(capture.output(
  fun4_correct(exdat, variable = 'a', interval = c(3, 7), incl_boundaries = FALSE, outside = FALSE)
), collapse = "\n"))
```


`incl_boundaries = FALSE, outside = TRUE` should include values 1, 2 and 8, 9, 10:
```{r, eval = FALSE}
fun4_correct(exdat, variable = 'a', interval = c(3, 7), incl_boundaries = FALSE, outside = TRUE)
```
```{r, echo = FALSE}
cat(paste0(capture.output(
  fun4_correct(exdat, variable = 'a', interval = c(3, 7), incl_boundaries = FALSE, outside = TRUE)
), collapse = "\n"))
```


<div style="border:2px; border-style:solid; padding: 1em; border-color:#009B8F">
<p style = "color:#009B8F; font-weight:bold;">Note:</p>
If you didn't figure this one out by yourself, don't worry. This really was a though one!
</div>


## Check for categorical variables {.tabset .tabset-pills .tabset-fade}

### Task 1
<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">
What happens when we call `fun4_correct()` but specify `variable` to be a factor?
</div>


### Solution 1
We can easily check that:
```{r}
fun4_correct(exdat, variable = 'b', interval = c(3, 7))
```

### Task 2
<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">
Add a check to `fun4_correct` that first checks if `variable` is of type `numeric`.
When a non-numeric variable is selected, the function should not try to produce
a subset but instead print a message.
```{r, echo = FALSE, results = 'asis'}
hint('You can use the function `is.numeric()` to check this.')
```

</div>

### Solution 2
With the functions that we have seen so far, one solution would be:
```{r}
fun5 <- function(dat, variable, interval, incl_boundaries = TRUE, outside = FALSE) {
  
  if (!is.numeric(dat[, variable])) {
    print('The variable you selected is not numeric!')
    
  } else { 
  
  filter <- if (incl_boundaries != outside) {
    dat[, variable] >= interval[1] & dat[, variable] <= interval[2]
  } else {
    dat[, variable] > interval[1] & dat[, variable] < interval[2]
  }
  
  if (outside) {
    filter <- !filter
  }
  
  subset(dat, filter)
  }
}
```

```{r}
fun5(exdat, variable = 'b', interval = c(3, 7))
```


<div style="border:2px; border-style:solid; padding: 1em; border-color:#009B8F">
<p style = "color:#009B8F; font-weight:bold;">Note:</p>
Typically we would not just want to print a message but either create a 
`warning()` or an error message (using `stop()`) which would immediately stop
the execution of the function:
```{r}
fun5b <- function(dat, variable, interval, incl_boundaries = TRUE, outside = FALSE) {
  
  if (!is.numeric(dat[, variable])) {
    stop('The variable you selected is not numeric!', call. = FALSE)
  }
  
  filter <- if (incl_boundaries != outside) {
    dat[, variable] >= interval[1] & dat[, variable] <= interval[2]
  } else {
    dat[, variable] > interval[1] & dat[, variable] < interval[2]
  }
  
  if (outside) {
    filter <- !filter
  }
  
  subset(dat, filter)
}
```

```{r, error = TRUE}
fun5b(exdat, variable = 'b', interval = c(3, 7))
```
When we use `stop()` we do not need the `else` part of the `if()` statement
because the function will be stopped immediately and the rest of the syntax
will not be evaluated.
</div>


