---
title: 'Summarizing Data'
output: 
  html_document:
    code_folding: show
    df_print: paged
    toc: true
    toc_float:
      collapsed: false
    number_sections: false
    theme: spacelab
    highlight: tango
    includes:
      after_body: ../footerNE.html
    css: ../style.css
---

```{r, echo = FALSE, purl = FALSE}
library(knitr)

knit_hooks$set(purl = hook_purl)

options(purl = FALSE)

opts_chunk$set(purl = FALSE)
```

```{r, include = FALSE}
opts_hooks$set(eval = function(opt) {
  if (any(opt$exercise))
    opt$eval <- opt$include <- FALSE
  
  opt
})


static <- TRUE

options(width = 100)


hint <- function(text) {
  id <- paste0('hint_', sample.int(1e10, size = 1))
    cat(
    paste0('\n',
      '<button type="button" class="btn btn-info btn-sm" data-toggle="collapse" ',
      'data-target="#', id, '">Hint</button>',
      '<div id = "', id, '" class="collapse" style="border:1px; ',
      'border-style:solid; padding: 1em; border-color:#1F78B4">',
      text, '</div>')
  )
}
```


## Dataset 
For this practical, we will use the **lung** dataset from the **survival** package.

You can either load the survival package to get access to this dataset
```{r}
library(survival)
```
or make a copy of this data
```{r}
lung <- survival::lung
```


## Getting to know the data
### Data dimensions {.tabset .tabset-fade .tabset-pills}
#### Taks
<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">
Find out the class and dimension of the **lung** data.
```{r, echo = FALSE, results = 'asis'}
hint('Use the functions `class()`, `dim()`, `nrow()` and `ncol()`.')
```
</div>

#### Solution
```{r}
dim(lung)
nrow(lung)
ncol(lung)
```


### Data Structure {.tabset .tabset-fade .tabset-pills}
#### Task 1
<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">
Now investigate the structure of the data.

* Which variables does the data contain?
* What types of variables are these?
* Are there missing values
```{r, echo = FALSE, results = 'asis'}
hint('Use the functions `str()`, `head()`, `names()` and `summary()`.')
```
</div>

#### Solution 1
```{r}
head(lung)
names(lung)
str(lung)
```

#### Task 2
<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">

All variables are coded as `numeric`.

* Do you think some of them should be factores?
* Explore how many distrinct values the variables have that may not actually be
  continuous.
```{r, echo = FALSE, results = 'asis'}
hint('The functions `unique()`, `table()` and `length()` are useful here.')
```
</div>

#### Solution 2

* Looking at the output of `str(lung)` we can see that `status`, `sex` and 
  `ph.ecog` may only have values 0, 1 and 2
* Also `inst`, `ph.karno`, `pat.karno`, and `wt.loss` could be categorical.

To confirm that `status`, `sex` and `ph.ecog` only have very few levels, `unique()`
or `table()` can be used:
```{r}
unique(lung$status)
table(lung$sex, exclude = NULL)
table(lung$ph.ecog, exclude = NULL)
```

To prevent lengthy output for the other variables we could first check how many
different vaues there are:
```{r}
length(unique(lung$inst))
length(unique(lung$ph.karno))
length(unique(lung$pat.karno))
length(unique(lung$wt.loss))
```


## Re-coding
### Turning continuous values into factors {.tabset .tabset-pills .tabset-fade}
#### Task
<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">

* Re-code `status`, `sex` and `ph.ecog` as factors.
* Use meaningful factor labels for `status` (1 = censored, 2 = dead) and
  `sex` (1 = male, 2 = female).
* Check that the resulting variables are indeed of class `factor` and that they
  have the correct levels.

```{r, echo = FALSE, results = 'asis'}
hint('Use the function `factor()`.')
hint('To make sure that the correct categories get the correct labels, specify 
     not only the argument `labels`, but also `levels`.')
hint('To check the class, use the function `class()`.
      To check levels, you can either use `levels()` or use a `table()`.')
```
</div>

#### Solution
```{r}
# For ph.ecog, just convert to a factor
lung$ph.ecog <- factor(lung$ph.ecog)

# For the other two, use levels and labels
lung$sex <- factor(lung$sex, levels = c(1, 2), labels = c('male', 'female'))
lung$status <- factor(lung$status, levels = c(1, 2), labels = c('censored', 'dead'))

# Confirm the class, either with
class(lung$ph.ecog)
class(lung$sex)
class(lung$status)
# or just use (str(lung))
 
levels(lung$ph.ecog)
table(lung$ph.ecog)
table(lung$sex)
table(lung$status)
```

## Descriptive Statistics
### Summary {.tabset .tabset-pills .tabset-fade}
#### Task 1
<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">
Get the summary of the **lung** data.
</div>

#### Solution 1
```{r}
summary(lung)
```

#### Task 2
<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">
* Calculate each of the values of this summary for the variables `ph.karno` 
and `ph.ecog` "by hand" (i.e., using other functions).
* Also calculate the inter-quartile range for `ph.karno`.

```{r, echo = FALSE, results = 'asis'}
hint('You can use the functions `min()`, `max()`, `quantile()` for the 
     continuous variable (but there are other options), 
     and `table()` for the categorical variable.')
hint('There are missin values in both variables: 
     remember to set `na.rm = TRUE` or `exclude = NULL`.')
```
</div>

#### Solution 2
```{r}
min(lung$ph.karno, na.rm = TRUE)
max(lung$ph.karno, na.rm = TRUE)
# alternatively: range(lung$ph.karno, na.rm = TRUE)

quantile(lung$ph.karno, probs = c(0.25, 0.5, 0.75), na.rm = TRUE)
# alternatively: median(lung$ph.karno, na.rm = TRUE)

IQR(lung$ph.karno, na.rm = TRUE)

table(lung$ph.ecog, exclude = NULL)
```

### Variance and Standard Deviation {.tabset .tabset-pills .tabset-fade}
#### Task
<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">
* Calculate the variance of `meal.cal`.
* Calculate the standard devition of `meal.cal`.
* Check that the standard deviation is indeed the square root of the variance.
* Check that the variance is indeed the square of the standard deviation.
```{r, echo = FALSE, results = 'asis'}
hint('Use the functions `sd()` and `var()`.')
hint('To calculate the square root, use the function `sqrt()`.')
hint('To calculate the square, use `^2`.')
```
</div>

#### Solution
```{r}
cal_sd <- sd(lung$meal.cal, na.rm = TRUE); cal_sd
cal_var <- var(lung$meal.cal, na.rm = TRUE); cal_var

sqrt(cal_var)
cal_sd^2
```


### Tables {.tabset .tabset-pills .tabset-fade}
#### Task 1
<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">
* Get a table of `sex` and `status`.
* Also calculate the probabilities for each combination:
  * relative to the total number of subjects
  * relative to `sex`
```{r, echo = FALSE, results = 'asis'}
hint('To get the probabilities, use the function `prop.table()`.')
hint('Use the argument `margin` to specify if the probabilities are relative to
the whole sample size or only one of the two variables.')
```
</div>

#### Solution 1
```{r}
# 2x2 table of status and sex:
tab <- table(lung$status, lung$sex)

# probabilities, relative to the total number of subjects:
prop.table(tab)

# probabilities, relative to sex:
prop.table(tab, margin = 2)
# Make sure choose the correct margin. When sex is in the rows, you need margin = 1.
# Check that the table shows the correct numbers by roughly adding the proportions
# per sex in your head.
```

#### Task 2
<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">
* Get a table of `sex` and `ph.ecog` per `status`.
* Experiment with the effect the order of the variables in `table()` has.
* Convert this 3-dimensional table into a flat table.
```{r, echo = FALSE, results = 'asis'}
hint('To get a flat table, use `ftable()`.')
```
</div>

#### Solution 2
```{r}
tab <- table(lung$sex, lung$ph.ecog, lung$status, exclude = NULL)
tab

# The first variable specifies the rows, the second the columns, and the 
# third variables the third dimension:
table(lung$ph.ecog, lung$status, lung$sex, exclude = NULL)

# Convert to a flat table:
ftable(tab, exclude = NULL)
```

