---
title: 'Extra Programming Examples'
output: 
  html_document:
    code_folding: show
    df_print: paged
    toc: true
    toc_float:
      collapsed: false
    number_sections: false
    theme: spacelab
    highlight: tango
    includes:
      after_body: ../footerEA.html
    css: ../style.css
---

```{r, echo = FALSE, purl = FALSE}
knitr::knit_hooks$set(purl = knitr::hook_purl)

options(purl = FALSE)

knitr::opts_chunk$set(purl = FALSE)
```

```{r, include = FALSE}
knitr::opts_hooks$set(eval = function(opt) {
  if (any(opt$exercise))
    opt$eval <- opt$include <- FALSE
  
  opt
})


static <- TRUE

options(width = 100)

```


```{r packages, include = FALSE}
library(kableExtra)
library(knitr)
```


```{r load_data, context="data", include=FALSE}
library(survival)
```


## Preface {data-progressive=FALSE}
### R packages
In this practical, a number of R packages are used.
The packages used (with versions that were used to generate the solutions) are:

* `r R.version.string`
* `survival` (version: `r packageVersion("mice")`)



### Datasets 

For this practical, we will use the **heart** and **retinopathy** data sets from the `survival` package. More details about the data sets can be found in:

https://stat.ethz.ch/R-manual/R-devel/library/survival/html/heart.html

https://stat.ethz.ch/R-manual/R-devel/library/survival/html/retinopathy.html

Always check your data.


## Combinations of Functions and Loops

### Common R Objects {.tabset .tabset-fade .tabset-pills}

```{r, eval = static, echo = FALSE}
asis_output("#### Task 1\\n")
```

```{r, eval = static, echo = FALSE}
asis_output('<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">')
```

Examine the following code in R.
```{r, eval = FALSE, echo = TRUE}
h <- function(k){
       if (k <= 20){
        3 * k
      } else {
      2 * k
    }
}
```
Investigate what this code takes as an input (e.g. scalar, vector, matrix, data.frame, list) and what the output will be in every case.


```{r, eval = static, echo = FALSE}
asis_output("</div>")

asis_output("#### Solution 1\\n")
```

```{r fun1-solution, solution = TRUE}
h <- function(k){
       if (k <= 20){
        3 * k
      } else {
      2 * k
    }
}
scalar_test <- 10
vec_test <- c(1:10)
mat_test <- matrix(1:6, 3, 3)
df_test <- data.frame(v1 = 1:10, v2 = sample(0:1, 10, replace = TRUE))
list_test <- list(l1 = 1:30, l2 = c("m", "f"), l3 = c(TRUE, TRUE, FALSE, FALSE, FALSE))
# h(scalar_test)
# h(vec_test)
# h(mat_test)
# h(df_test)
# h(list_test)
```



```{r, eval = static, echo = FALSE}
asis_output("#### Task 2\\n")
```

```{r, eval = static, echo = FALSE}
asis_output('<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">')
```

Examine the following code in R.
```{r, eval = FALSE, echo = TRUE}
h <- function(k){
  for(i in 1:length(k)){
    k[i]<- if (k[i] <= 20){
      3 * k[i]
    } else {
      2 * k[i] 
    }
  }
  k
}
```
Investigate what this code takes as an input (e.g. scalar, vector, matrix, data.frame, list) and what the output will be in every case.


```{r, eval = static, echo = FALSE}
asis_output("</div>")

asis_output("#### Solution 2\\n")
```

```{r fun2-solution, solution = TRUE}
h <- function(k){
  for(i in 1:length(k)){
    k[i]<- if (k[i] <= 20){
      3 * k[i]
    } else {
      2 * k[i] 
    }
  }
  k
}
scalar_test <- 10
vec_test <- c(1:10)
mat_test <- matrix(1:6, 3, 3)
df_test <- data.frame(v1 = 1:10, v2 = sample(0:1, 10, replace = TRUE))
list_test <- list(l1 = 1:30, l2 = c("m", "f"), l3 = c(TRUE, TRUE, FALSE, FALSE, FALSE))
# h(scalar_test)
# h(vec_test)
# h(mat_test)
# h(df_test)
# h(list_test)
```





```{r, eval = static, echo = FALSE}
asis_output("#### Task 3\\n")
```

```{r, eval = static, echo = FALSE}
asis_output('<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">')
```

Define a function (with the name `mysummary`) that takes one parameter (let’s call it `x`) that investigates if it is a data frame, a factor, or a numeric variable and returns the message "This is a data.frame" if `x` is a data frame, "This is a factor" is `x` is a factor and "This is numeric" if `x` is numeric.  


```{r fun3_hint, eval = static, results = 'asis', echo = FALSE}
cat('<button type="button" class="btn btn-info btn-sm" data-toggle="collapse" data-target="#fun3">Hint</button>
<div id = "fun3" class="collapse" style="border:1px; border-style:solid; padding: 1em; border-color:#1F78B4">
Use an if statement. When you want to display results from the function you can use the print(...) function.
</div>')
```


```{r, eval = static, echo = FALSE}
asis_output("</div>")

asis_output("#### Solution 3\\n")
```

```{r fun3-solution, solution = TRUE}
mysummary <- function(x){
  if(is.data.frame(x)){
    print('This is a data.frame')
  }else if(is.factor(x)){
    print('This is a factor')
  }else if(is.numeric(x)){
    print('This is numeric')
  }
}
df_test <- data.frame(v1 = 1:10, v2 = sample(0:1, 10, replace = TRUE))
vec1_test <- as.factor(c("y", "n", "n", "n"))
vec2_test <- 1:10
mysummary(df_test)
mysummary(vec1_test)
mysummary(vec2_test)
```




### Data Exploration {.tabset .tabset-fade .tabset-pills}

```{r, eval = static, echo = FALSE}
asis_output("#### Task 1\\n")
```

```{r, eval = static, echo = FALSE}
asis_output('<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">')
```

Extend the previous function `mysummary` to take again one parameter (let’s call it `x`) but to return also some descriptive statistics. We therefore modify the function in such a way that a frequency table is given for factors and the mean and sd are given for numeric data.

```{r fun4_hint, eval = static, results = 'asis', echo = FALSE}
cat('<button type="button" class="btn btn-info btn-sm" data-toggle="collapse" data-target="#fun4">Hint</button>
<div id = "fun4" class="collapse" style="border:1px; border-style:solid; padding: 1em; border-color:#1F78B4">
Use the functions table(...), mean(...) and sd(...).
</div>')
```


```{r, eval = static, echo = FALSE}
asis_output("</div>")

asis_output("#### Solution 1\\n")
```

```{r fun4-solution, solution = TRUE}
mysummary <- function(x){
  if(is.data.frame(x)){
    print('A data.frame')
    print('Not yet implemented')
  }else if(is.factor(x)){
    print('This is afactor')
    print(table(x))
  }else if(is.numeric(x)){
    print('This variabele is numeric')
    print('mean')
    print(mean(x))
    print('sd')
    print(sd(x))
  }
}
df_test <- data.frame(v1 = 1:10, v2 = sample(0:1, 10, replace = TRUE))
vec1_test <- as.factor(c("y", "n", "n", "n"))
vec2_test <- 1:10
mysummary(df_test)
mysummary(vec1_test)
mysummary(vec2_test)
```



```{r, eval = static, echo = FALSE}
asis_output("#### Task 2\\n")
```

```{r, eval = static, echo = FALSE}
asis_output('<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">')
```

The previous function `mysummary` still does not do anything useful when we apply it on the whole data.frame. Extend the function to take again one parameter (let’s call it `x`) but to return also some descriptive statistics for the data frame. 

```{r fun5_hint, eval = static, results = 'asis', echo = FALSE}
cat('<button type="button" class="btn btn-info btn-sm" data-toggle="collapse" data-target="#fun5">Hint</button>
<div id = "fun5" class="collapse" style="border:1px; border-style:solid; padding: 1em; border-color:#1F78B4">
This can be solved in the following way: whenever the function is called using a data.frame as argument, we use a for loop to loop over its columns. Within this loop we let the function call itself but now using the column as argument. 
</div>')
```


```{r, eval = static, echo = FALSE}
asis_output("</div>")

asis_output("#### Solution 2\\n")
```

```{r fun5-solution, solution = TRUE}
mysummary <- function(x){
  if(is.data.frame(x)){
    print('A data.frame')
    for(i in 1:ncol(x)){
      print(names(x)[i])
      mysummary(x[,i])
    }     
  }else if(is.factor(x)){
    print('A factor')
    print(table(x))
  }else if(is.numeric(x)){
    print('This variabele is numeric')
    print('mean')
    print(mean(x))
    print('sd')
    print(sd(x))
  }
}
df1_test <- data.frame(v1 = 1:10, v2 = sample(0:1, 10, replace = TRUE))
df2_test <- data.frame(v1 = 1:10, v2 = sample(c("no", "yes"), 10, replace = TRUE))
mysummary(df1_test)
mysummary(df2_test)
```


### Data Manipulation {.tabset .tabset-fade .tabset-pills}

```{r, eval = static, echo = FALSE}
asis_output("#### Task\\n")
```

```{r, eval = static, echo = FALSE}
asis_output('<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">')
```


Create a function with the name `std_num` that takes as input a data frame (let's call it `DF`), standardizes all the numerical variables and returns the new data frame. Apply this function to the **heart** data set.


```{r fun6_hint, eval = static, results = 'asis', echo = FALSE}
cat('<button type="button" class="btn btn-info btn-sm" data-toggle="collapse" data-target="#fun6">Hint</button>
<div id = "fun6" class="collapse" style="border:1px; border-style:solid; padding: 1em; border-color:#1F78B4">
Use an if statement and a for loop. Note that you will have to transform some variables to factors first.
</div>')
```


```{r, eval = static, echo = FALSE}
asis_output("</div>")

asis_output("#### Solution\\n")
```

```{r fun6-solution, solution = TRUE}
heart$event <- as.factor(heart$event)
heart$surgery <- as.factor(heart$surgery)
heart$transplant <- as.factor(heart$transplant)
heart$id <- as.factor(heart$id)

std_num <- function(DF){
  for (j in 1:dim(DF)[2]){
    if (is.numeric(DF[,j])){
      DF[colnames(DF)[j]] <- (DF[colnames(DF)[j]] - mean(DF[[colnames(DF)[j]]]))/sd(DF[[colnames(DF)[j]]])
    }
  }
  DF
}
std_num(heart)
```



### Data Visualization {.tabset .tabset-fade .tabset-pills}

```{r, eval = static, echo = FALSE}
asis_output("#### Task 1\\n")
```

```{r, eval = static, echo = FALSE}
asis_output('<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">')
```


Create a for loop that goes through the columns of the **retinopathy** data set and plots each column. If the column is a numerical variable then use a density plot, if the column is a categorical variable then use a barchart. Create a plot with multiple panel


```{r fun7_hint, eval = static, results = 'asis', echo = FALSE}
cat('<button type="button" class="btn btn-info btn-sm" data-toggle="collapse" data-target="#fun7">Hint</button>
<div id = "fun7" class="collapse" style="border:1px; border-style:solid; padding: 1em; border-color:#1F78B4">
Use an if statement. Use the function par(mfrow = ...) to create multiple plots. Note that you will have to transform some variables to factors first.
</div>')
```


```{r, eval = static, echo = FALSE}
asis_output("</div>")

asis_output("#### Solution 1\\n")
```

```{r fun7-solution, solution = TRUE}
retinopathy$trt <- as.factor(retinopathy$trt)
retinopathy$status <- as.factor(retinopathy$status)

par(mfrow = c(3, 3))
for (j in 1:dim(retinopathy)[2]){
  if (is.numeric(retinopathy[,j])){
    plot(density(retinopathy[,j]), col = rgb(0,0,1,0.5))
    polygon(density(retinopathy[,j]), col = rgb(0,0,1,0.5), border = "blue")
  } else if (is.factor(retinopathy[,j])){
    plot(retinopathy[,j])
  }
}
```




```{r, eval = static, echo = FALSE}
asis_output("#### Task 2\\n")
```

```{r, eval = static, echo = FALSE}
asis_output('<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">')
```


Create a function that applies the previous code in any data set. The name of the function will be `plot_summary` and it will take as input a data frame (let's call it `dt`) and the dimension of the panel that includes the plots. In particular, `dim_row` represents the rows and `dim_col` the columns. The function will return a plot for each column. If the column is a numerical variable then use a density plot, if the column is categorical variable then use a barchart. Apply this function to the **retinopathy** data set.


```{r fun8_hint, eval = static, results = 'asis', echo = FALSE}
cat('<button type="button" class="btn btn-info btn-sm" data-toggle="collapse" data-target="#fun8">Hint</button>
<div id = "fun8" class="collapse" style="border:1px; border-style:solid; padding: 1em; border-color:#1F78B4">
Use an if statement and a for loop. Use the function par(mfrow = ...) to create multiple plots. Note that you will have to transform some variables to factors first.
</div>')
```


```{r, eval = static, echo = FALSE}
asis_output("</div>")

asis_output("#### Solution 2\\n")
```

```{r fun8-solution, solution = TRUE}
retinopathy$trt <- as.factor(retinopathy$trt)
retinopathy$status <- as.factor(retinopathy$status)

plot_summary <- function(dt, dim_row, dim_col){
  par(mfrow = c(dim_row, dim_col))
  for (j in 1:dim(dt)[2]){
    if (is.numeric(dt[,j])){
      plot(density(dt[,j]), col = rgb(0,0,1,0.5))
      polygon(density(dt[,j]), col = rgb(0,0,1,0.5), border = "blue")
    } else if (is.factor(dt[,j])){
      plot(dt[,j])
    }
  }
}
plot_summary(retinopathy, 3, 3)
```



```{r, eval = static, echo = FALSE}
asis_output("#### Task 3\\n")
```

```{r, eval = static, echo = FALSE}
asis_output('<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">')
```


Extend the previous function `plot_summary` to include also plot titles indicating the variable. Apply this function to the **retinopathy** data set.


```{r fun9_hint, eval = static, results = 'asis', echo = FALSE}
cat('<button type="button" class="btn btn-info btn-sm" data-toggle="collapse" data-target="#fun9">Hint</button>
<div id = "fun9" class="collapse" style="border:1px; border-style:solid; padding: 1em; border-color:#1F78B4">
Use the paste0(...) function.
</div>')
```


```{r, eval = static, echo = FALSE}
asis_output("</div>")

asis_output("#### Solution 3\\n")
```

```{r fun9-solution, solution = TRUE}
retinopathy$trt <- as.factor(retinopathy$trt)
retinopathy$status <- as.factor(retinopathy$status)

plot_summary <- function(dt, dim_row, dim_col){
  par(mfrow = c(dim_row, dim_col))
  for (j in 1:dim(dt)[2]){
    if (is.numeric(dt[,j])){
      plot(density(dt[,j]), col = rgb(0,0,1,0.5), main = paste0(colnames(dt)[j]))
      polygon(density(dt[,j]), col = rgb(0,0,1,0.5), border = "blue")
    } else if (is.factor(dt[,j])){
      plot(dt[,j], main = paste0(colnames(dt)[j]))
    }
  }
}
plot_summary(retinopathy, 3, 3)
```




### Subsetting {.tabset .tabset-fade .tabset-pills}

```{r, eval = static, echo = FALSE}
asis_output("#### Task 1\\n")
```

```{r, eval = static, echo = FALSE}
asis_output('<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">')
```


For each data set (**heart** and **retinopathy**), select only the observations that report an event.


```{r fun10_hint, eval = static, results = 'asis', echo = FALSE}
cat('<button type="button" class="btn btn-info btn-sm" data-toggle="collapse" data-target="#fun10">Hint</button>
<div id = "fun10" class="collapse" style="border:1px; border-style:solid; padding: 1em; border-color:#1F78B4">
Note that the variable name of the events is diffent for the two data sets.
</div>')
```


```{r, eval = static, echo = FALSE}
asis_output("</div>")

asis_output("#### Solution 1\\n")
```

```{r fun10-solution, solution = TRUE}
heart[heart$event == 1, ]
retinopathy[retinopathy$status == 1, ]
```



```{r, eval = static, echo = FALSE}
asis_output("#### Task 2\\n")
```

```{r, eval = static, echo = FALSE}
asis_output('<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">')
```


Create a function with the name `dt_events` that takes as input a list with data sets (let's call it `dt_list`) and a vector indicating the name of the event variable for each data set (let's call it `event_name`). The function will return a list consisting of each data set but including only the observations/rows that had an event. Create a list that consists of the data sets **heart** and **retinopathy** and apply the function.


```{r fun11_hint, eval = static, results = 'asis', echo = FALSE}
cat('<button type="button" class="btn btn-info btn-sm" data-toggle="collapse" data-target="#fun11">Hint</button>
<div id = "fun11" class="collapse" style="border:1px; border-style:solid; padding: 1em; border-color:#1F78B4">
Use a for loop. Note that the variable name of the events is different for the two data sets.
</div>')
```


```{r, eval = static, echo = FALSE}
asis_output("</div>")

asis_output("#### Solution 2\\n")
```

```{r fun11-solution, solution = TRUE}
dt_events <- function(dt_lists, event_name){
  new_dt <- list()
  for (k in 1:length(dt_lists)) {
      dt <- dt_lists[[k]]
      new_dt[[k]] <- dt[dt[[event_name[k]]] == 1, ]
  }
  new_dt
}

dt_events(dt_lists = list(heart, retinopathy), event_name = c("event", "status"))
```


### The Apply family {.tabset .tabset-fade .tabset-pills}

```{r, eval = static, echo = FALSE}
asis_output("#### Task 1\\n")
```

```{r, eval = static, echo = FALSE}
asis_output('<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">')
```


For each data set (**heart** and **retinopathy**), obtain the mean age per event status.


```{r, eval = static, echo = FALSE}
asis_output("</div>")

asis_output("#### Solution 1\\n")
```

```{r fun12-solution, solution = TRUE}
tapply(heart$age, heart$event, mean)
tapply(retinopathy$age, retinopathy$status, mean)
```



```{r, eval = static, echo = FALSE}
asis_output("#### Task 2\\n")
```

```{r, eval = static, echo = FALSE}
asis_output('<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">')
```


Create a function with the name `summary_list` that takes as input a list with data sets (let's call it `dt_list`), a vector with the names of the continuous variables (let's call it `dt_cont`) and a vector with the names of the categorical variables (let's call it `dt_cat`) and returns the mean of the continuous variable per group (categorical variable) for each data set. Apply the function to the **heart** and **retinopathy** data sets. In particular, use the continuous variables `year` (from the **heart** data set) and `risk` (from the **retinopathy** data set) and the categorical variables `surgery` (from the **heart** data set) and `type` (from the **retinopathy** data set).

```{r fun13_hint, eval = static, results = 'asis', echo = FALSE}
cat('<button type="button" class="btn btn-info btn-sm" data-toggle="collapse" data-target="#fun13">Hint</button>
<div id = "fun13" class="collapse" style="border:1px; border-style:solid; padding: 1em; border-color:#1F78B4">
Use a for loop. 
</div>')
```


```{r, eval = static, echo = FALSE}
asis_output("</div>")

asis_output("#### Solution 2\\n")
```

```{r fun13-solution, solution = TRUE}
summary_list <- function(dt_list, dt_cont, dt_cat){
  for (i in 1:length(dt_list)){
    dt <- dt_list[[i]]
    print(tapply(dt[[dt_cont[i]]], dt[[dt_cat[i]]], mean))
  }
}

summary_list(dt_list = list(heart, retinopathy), dt_cont = c("year", "risk"), dt_cat = c("surgery", "type"))
```



```{r, eval = static, echo = FALSE}
asis_output("#### Task 3\\n")
```

```{r, eval = static, echo = FALSE}
asis_output('<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">')
```


Extend the previous function to include the mean as input argument (let's call it `calc`). Therefore, the user can decide which function to use.


```{r, eval = static, echo = FALSE}
asis_output("</div>")

asis_output("#### Solution 3\\n")
```

```{r fun14-solution, solution = TRUE}
summary_list <- function(dt_list, dt_cont, dt_cat, calc){
  for (i in 1:length(dt_list)){
    dt <- dt_list[[i]]
    print(tapply(dt[[dt_cont[i]]], dt[[dt_cat[i]]], calc))
  }
}

summary_list(dt_list = list(heart, retinopathy), dt_cont = c("year", "risk"), dt_cat = c("surgery", "type"), calc = median)
```
