---
title: 'Merging Datasets'
output: 
  html_document:
    code_folding: show
    df_print: paged
    toc: true
    toc_float:
      collapsed: false
    number_sections: false
    theme: spacelab
    highlight: tango
    includes:
      after_body: ../footerNE.html
    css: ../style.css
---

```{r setup, include = FALSE}
library(knitr)
library(kableExtra)

opts_chunk$set(echo = TRUE)

hint <- function(text) {
  id <- paste0('hint_', sample.int(1e10, size = 1))
    cat(
    paste0('\n',
      '<button type="button" class="btn btn-info btn-sm" data-toggle="collapse" ',
      'data-target="#', id, '">Hint</button>',
      '<div id = "', id, '" class="collapse" style="border:1px; ',
      'border-style:solid; padding: 1em; border-color:#1F78B4">',
      text, '</div>')
  )
}
```

## Data

In this practical, we will work with example datasets that are
created by the following syntax:
```{r}
basedat <- data.frame(id = 1:5,
                      age = c(37.9, 22.6, 26.9, 37.3, 49.2),
                      sex = factor(c('male', 'female', 'female', 'male', 'male')),
                      center = c('EMC', 'EMC', 'LUMC', 'VUMC', 'LUMC')
)
FUdat <- data.frame(id = c(1, 1, 1, 2, 2, 3, 4, 4, 4, 4, 5, 5, 5),
                    time = c(3.6, 5.3, 7.8, 4.3, 8.2, 5.3, 3.5, 
                             4.9, 6.4, 8.1, 2.2, 6.4, 7.0),
                    weight = c(136.7, 131.5, 135.0, 113.7, 128.9, 97.1, 
                               168.1, 147.7, 153.1, 149.6, 152.6, 137.7, 133.1))

labdat <- data.frame(id = 1:12,
                     ID = c(1, 1, 2, 2, 2, 4, 4, 4, 4, 5, 5, 6),
                     labtime = c(3.1, 6.0, 4.3, 5.8, 8.2, 3.5, 
                                 4.9, 6.4, 8.1, 2.2, 6.4, 4.7),
                     chol = c(347, 309, 215, 436, 360, 335, 1002, 808, 615, 
                              932, 413, 337),
                     center = c(rep('EMC', 5), rep('VUMC', 4), rep('LUMC', 3))
)
```

The resulting datasets are:

**basedat**
```{r, echo = FALSE}
knitr::kable(basedat) %>%
  kable_styling(full_width = FALSE)
```

**FUdat**
```{r, echo = FALSE}
knitr::kable(FUdat) %>%
  kable_styling(full_width = FALSE)
```

**labdat**
```{r, echo = FALSE}
knitr::kable(labdat) %>%
  kable_styling(full_width = FALSE)
```

## Merging `basedat` and `FUdat` {.tabset .tabset-pills .tabset-fade}
### Task 1
<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">
Run the syntax given above to create the datasets `basedat` which contains 
"baseline" information on each patient and `FUdat`, which contains repeated
measurements of a variable `weight`.

Merge `basedat` and `FUdat` by the variable `id`.
```{r, echo = FALSE, results='asis'}
hint('Use the function `merge()`.')
```

</div>

### Solution 1
```{r}
dat1 <- merge(basedat, FUdat)
dat1
```

### Task 2
<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">
Check for two of the subjects that the merged data are correct.
```{r, echo = FALSE, results = 'asis'}
hint("The function `subset()` could be used here.")
```
</div>

### Solution 2
```{r}
# for subject 1
subset(basedat, id == 1)
subset(FUdat, id == 1)
subset(dat1, id == 1)

# for subject 4, using a different method of subsetting (just as a reminder how that worked)
basedat[basedat$id == 4, ]
FUdat[FUdat$id == 4, ]
dat1[dat1$id == 4, ]
```


### Task 3
<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">
Check that the resulting `dat1` has the correct dimensions.
```{r, echo = FALSE, results = 'asis'}
hint("The function `dim()` could be used here.")
```
</div>


### Solution 3
```{r}
dim(dat1)
dim(FUdat)
dim(basedat)
```
`FUdat` and `basedat` contain data on the exact same set of subjects
and only the `id` variable is overlapping between the two datasets.
Therefore, the resulting `dat1` should have the same number of rows as the data
containing the repeated measurements (`FUdat`) and has one column less than 
the sum of the number of columns of both input datasets (6 = 3 + 4 - 1).


## Merging with `labdat` {.tabset .tabset-pills .tabset-fade}
We now want to merge the previously merged data `dat1` with `labdat`.

### Task 1
Note that in `labdat` we have the variables

* `id`: sample number
* `ID`: patient identifier


<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">
Merge `dat` and `labdat` by patient identifier.
```{r, echo = FALSE, results = 'asis'}
hint("Use the arguments `by.x` and `by.y`.")
```
</div>


### Solution 1
```{r}
dat2 <- merge(dat1, labdat, by.x = 'id', by.y = 'ID')
dat2
```

### Task 2
Did you notice that in the resulting `dat2` we now have the two variables
`center.x` and `center.y`, and that we have a variable `id.y`
(which is the sample number from `labdat`)?

Moreover, there are two time variables: `time` (from `dat1`) and `labtime` 
(from `labdat`).

<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">
Re-do the merge of `dat1` and `labdat` taking into account the following points

* merge by patient identifier and time of measurement
* make sure there is only one "center" variable
* give the sample numer (`id.y`) a more informative suffix
```{r, echo = FALSE, results = 'asis'}
hint("Use the arguments `by.x` and `by.y`.")
```
```{r, echo = FALSE, results = 'asis'}
hint("Use the argument `suffixes`.")
```
</div>

### Solution 2
```{r}
dat3 <- merge(dat1, labdat, 
              by.x = c('id', 'center', 'time'),
              by.y = c('ID', 'center', 'labtime'),
              suffixes = c('_base_FU', '_lab'))
```

**Note:**<br>
Here it does not actually matter how we specify the first element of the argument
`suffixes` because it is not used.

### Task 3
You can notice that the merged `dat3` now only contains very few rows.

<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">
* Can you explain what happened?
* Re-do the last merge so that the resulting dataset has the maximal number of rows. 
```{r, echo = FALSE, results = 'asis'}
hint("You need to use the argument `all`.")
```

</div>

### Solution 3
By default, the argument `all` is set to `FALSE`. This means that only those 
rows are kept that have identical entries for all merging variables
in both datasets.
Only a few values of `time` and `labtime` are identical.

When we set `all = TRUE` all rows will be kept.
Where `time` and `labtime` do not match, two rows are created and `NA` is filled in.
```{r}
dat4 <- merge(dat1, labdat, 
              by.x = c('id', 'center', 'time'),
              by.y = c('ID', 'center', 'labtime'),
              suffixes = c('_base_FU', '_lab'), all = TRUE)
dat4
```

