---
title: 'Rolling the Dice'
output: 
  html_document:
    code_folding: show
    df_print: paged
    toc: true
    toc_float:
      collapsed: false
    number_sections: false
    theme: spacelab
    highlight: tango
    includes:
      after_body: ../footerNE.html
    css: ../style.css
---


<link href="../fontawesome-free-5.12.0-web/css/all.css" rel="stylesheet"> <!--load all styles -->

```{r setup, include = FALSE}
library(knitr)
library(kableExtra)

opts_chunk$set(echo = TRUE)

hint <- function(text) {
  id <- paste0('hint_', sample.int(1e10, size = 1))
    cat(
    paste0('\n',
      '<button type="button" class="btn btn-info btn-sm" data-toggle="collapse" ',
      'data-target="#', id, '">Hint</button>',
      '<div id = "', id, '" class="collapse" style="border:1px; ',
      'border-style:solid; padding: 1em; border-color:#1F78B4">',
      text, '</div>')
  )
}
```



In this practical we will write a simple function that simulates rolling one or
more dice.



## The function `sample()`
The function `sample` allows us to draw values from a given set of values, 
with or without replacement.

It has arguments

* `x`: a vector of values to draw from
* `size`: the number of values to draw
* `replace`: can we choose the same element of `x` multiple times?
* `prob`: a vector of probability weights (by default all elements of `x` have equal probability to be drawn)

## One die <i class="fas fa-dice-five"></i> {.tabset .tabset-pills .tabset-fade}
### Task 1
<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">

Write the syntax to simulate a single role of a die using the function `sample()`.
```{r, echo = FALSE, results = 'asis'}
hint("You need to specify the arguments `x` and `size`.")
```
</div>

### Solution 1
```{r}
sample(1:6, size = 1)
```

### Task 2
<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">
Write your own function that simulates this one role of a die.
```{r, echo = FALSE, results = 'asis'}
hint('The general form of a function is<br>
     `<function name> <- function ( <arguments> ) {`<br>
      &nbsp;&nbsp;`<function body>`<br>
     `}`')
```
</div>

### Solution 2
```{r}
die <- function() {
  sample(1:6, size = 1)
}

die()
die()
die()
```

## Multiple dice <i class="fas fa-dice"></i> {.tabset .tabset-pills .tabset-fade}
### Task 1 
<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">
Take another look at the description of the function `sample()`. How would 
you need to change the syntax from **Task 1** from above to simulate rolling
two dice?
```{r, echo = FALSE, results = 'asis'}
hint('You need to change what is passed to the argument `size` and you need
to specify the argument `replace`.')
```
</div>

### Solution 1
```{r}
sample(1:6, size = 2, replace = TRUE)
```

### Task 2
<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">

* Write another function that simulates rolling multiple dice.
* Write the function so that the number of dice is flexible.
```{r, echo = FALSE, results = 'asis'}
hint('This new function needs to have an argument that can be passed to `sample`().')
```

</div>

### Solution 2
```{r}
dice <- function(n) {
  sample(1:6, size = n, replace = TRUE)
}

dice(2)
dice(1)
dice(10)
```

### Task 3
<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">
* What happens when you use the function `dice()` without specifying the argument
`n`?
* Do you know or can you guess why that happens?
</div>

### Solution 3
```{r}
dice()
```

By default, the function `dice()` will role 6 dice.

This is due to the behaviour of the function `sample()`. In the help file
for `sample()` we can read in the section **Details**

> For sample the default for size is the number of items inferred from the first
argument, so that `sample(x)` generates a random permutation of the elements of
`x` (or `1:x`).

**What can we learn from this?**

* Read the help files!
* Think about / try what will happen if "wrong" input is given to your function!

### Task 4
<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">
Re-write the function `dice()` so that by default only one die is used.
</div>

### Solution 4
```{r}
dice <- function(n = 1) {
  sample(1:6, size = n, replace = TRUE)
}

dice()
dice()
```

