---
title: 'The Apply Family'
output: 
  html_document:
    code_folding: show
    df_print: paged
    toc: true
    toc_float:
      collapsed: false
    number_sections: false
    theme: spacelab
    highlight: tango
    includes:
      after_body: ../footerEA.html
    css: ../style.css
---

```{r, echo = FALSE, purl = FALSE}
knitr::knit_hooks$set(purl = knitr::hook_purl)

options(purl = FALSE)

knitr::opts_chunk$set(purl = FALSE)
```

```{r, include = FALSE}
knitr::opts_hooks$set(eval = function(opt) {
  if (any(opt$exercise))
    opt$eval <- opt$include <- FALSE
  
  opt
})


static <- TRUE

options(width = 100)

heart <- survival::heart

```


```{r packages, include = FALSE}
library(kableExtra)
library(knitr)
library(memisc)
```


```{r load_data, context="data", include=FALSE}
library(survival)
```


## Preface {data-progressive=FALSE}
### R packages
In this practical, a number of R packages are used.
The packages used (with versions that were used to generate the solutions) are:

* `r R.version.string`
* `survival` (version: `r packageVersion("survival")`)



### Datasets 

For this practical, we will use the **heart** and **retinopathy** data sets from the `survival` package. More details about the data sets can be found in:

https://stat.ethz.ch/R-manual/R-devel/library/survival/html/heart.html

https://stat.ethz.ch/R-manual/R-devel/library/survival/html/retinopathy.html




## The Apply Family 

### apply {.tabset .tabset-fade .tabset-pills}

```{r, eval = static, echo = FALSE}
asis_output("#### Task 1\\n")
```

```{r, eval = static, echo = FALSE}
asis_output('<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">')
```

* Obtain the mean of the columns `start`, `stop`, `event`, `age`, `year`, `surgery` of the **heart** data set. \
* Obtain the mean of the columns `age`, `futime`, `risk` of the **retinopathy** data set. 


```{r, eval = static, echo = FALSE}
asis_output("</div>")

asis_output("#### Solution 1\\n")
```

```{r apply1-solution, solution = TRUE}
apply(heart[, c("start", "stop", "event", "age", "year", "surgery")], 2, mean)
apply(retinopathy[, c("age", "futime", "risk")], 2, mean)
```




```{r, eval = static, echo = FALSE}
asis_output("#### Task 2\\n")
```

```{r, eval = static, echo = FALSE}
asis_output('<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">')
```

Create the matrix `dataset1 <- cbind(A = 1:30, B = sample(1:100, 30))` and find the row sum of `dataset1`.


```{r, eval = static, echo = FALSE}
asis_output("</div>")

asis_output("#### Solution 2\\n")
```

```{r apply2-solution, solution = TRUE}
dataset1 <- cbind(A = 1:30, B = sample(1:100, 30))
apply(dataset1, 1, sum)
```


### lapply {.tabset .tabset-fade .tabset-pills}

```{r, eval = static, echo = FALSE}
asis_output("#### Task 1\\n")
```

```{r, eval = static, echo = FALSE}
asis_output('<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">')
```

Create the following function `DerivativeFunction <- function(x) { log10(x) + 10 }`. Apply the `DerivativeFunction` to `dataset1 <- cbind(A = 1:30, B = sample(1:100, 30))`. The output should be a list.

```{r, eval = static, echo = FALSE}
asis_output("</div>")

asis_output("#### Solution 1\\n")
```

```{r lapply1-solution, solution = TRUE}
DerivativeFunction <- function(x) { log10(x) + 10 }
dataset1 <- cbind(A = 1:30, B = sample(1:100, 30))
lapply(dataset1, DerivativeFunction)
```



```{r, eval = static, echo = FALSE}
asis_output("#### Task 2\\n")
```

```{r, eval = static, echo = FALSE}
asis_output('<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">')
```

* Create a list that consist of the variables `age` and `year` from the **heart** data set and the variable `risk` from the **retinopathy** data set. Give the name `list1` to this list. \
* Obtain the median of each element of the list. The output should be a list.

```{r, eval = static, echo = FALSE}
asis_output("</div>")

asis_output("#### Solution 2\\n")
```

```{r lapply2-solution, solution = TRUE}
list1 <- list(heart$age, heart$year, retinopathy$risk)
lapply(list1, median)
```



### sapply {.tabset .tabset-fade .tabset-pills}

```{r, eval = static, echo = FALSE}
asis_output("#### Task 1\\n")
```

```{r, eval = static, echo = FALSE}
asis_output('<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">')
```

Create the following function `Function2 <- function(x) { exp(x) + 0.1 }`. Apply the `Function2` to `dataset2 <- cbind(A = c(1:10), B = rnorm(10, 0, 1))`. The output should be simplified.

```{r, eval = static, echo = FALSE}
asis_output("</div>")

asis_output("#### Solution 1\\n")
```

```{r sapply1-solution, solution = TRUE}
Function2 <- function(x) { exp(x) + 0.1 }
dataset2 <- cbind(A = c(1:10), B = rnorm(10, 0, 1))
sapply(dataset2, Function2)
```



```{r, eval = static, echo = FALSE}
asis_output("#### Task 2\\n")
```

```{r, eval = static, echo = FALSE}
asis_output('<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">')
```

* Create a list that consist of the variable `transplant` from the **heart** data set and the variable `status` from the **retinopathy** data set. Give the name `list2` to this list. \
* Obtain the percentages of 0 cases of each element of the list in a simplified output (as a vector).
* Obtain the percentages of 1 cases of each element of the list in a simplified output (as a vector).

```{r, eval = static, echo = FALSE}
asis_output("</div>")

asis_output("#### Solution 2\\n")
```

```{r sapply2-solution, solution = TRUE}
list2 <- list(heart$transplant, retinopathy$status)
sapply(list2, function(x) { percent(x) } )
sapply(list2, function(x) { percent(x) } )
```


```{r, eval = static, echo = FALSE}
asis_output("#### Task 3\\n")
```

```{r, eval = static, echo = FALSE}
asis_output('<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">')
```

* Do you remember the practical `Control_Flow_and_Functions`: `Writing your own function (Task 1 and 2)`? Now try to create again the same function (called `summary_df`) but avoid the use of a for loop. Apply the function to the **retinopathy** dat set.

```{r sapply3_hint, eval = static, results = 'asis', echo = FALSE}
cat('<button type="button" class="btn btn-info btn-sm" data-toggle="collapse" data-target="#sapply3">Hint</button>
<div id = "sapply3" class="collapse" style="border:1px; border-style:solid; padding: 1em; border-color:#1F78B4">
Use the functions summary_continuous() and summary_categorical(). \\
`summary_continuous <- function(x) { ` \\
 &nbsp;&nbsp;  `paste0(round(mean(x), 1), " ( ", round(sd(x), 1), ") ") ` \\
`}`

`summary_categorical <- function(x) {` \\
 &nbsp;&nbsp; `tab <- prop.table(table(x)) `\\
 &nbsp;&nbsp; `paste0(round(tab * 100, 1), "% ", names(tab), collapse = ", ")`\\
`}`
</div>')
```


```{r, eval = static, echo = FALSE}
asis_output("</div>")

asis_output("#### Solution 3\\n")
```

```{r sapply3-solution, solution = TRUE}
summary_continuous <- function(x) {
  paste0(round(mean(x), 1), " (", round(sd(x), 1), ")")
}

summary_categorical <- function(x) {
  tab <- prop.table(table(x))
  paste0(round(tab * 100, 1), "% ", names(tab), collapse = ", ")
}


summary_df <- function(dat) {
  vec_categorical <- sapply(dat, is.factor)
  print(sapply(dat[,vec_categorical], summary_categorical))
  vec_continuous <- sapply(dat, is.numeric)
  print(sapply(dat[,vec_continuous], summary_continuous))
}

summary_df(dat = retinopathy)

```


### tapply {.tabset .tabset-fade .tabset-pills}

```{r, eval = static, echo = FALSE}
asis_output("#### Task 1\\n")
```

```{r, eval = static, echo = FALSE}
asis_output('<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">')
```

* Obtain the median `year` per `transplant` group using the **heart** data set. \
* Obtain the median `futime` per `status` group using the **retinopathy** data set. \

```{r, eval = static, echo = FALSE}
asis_output("</div>")

asis_output("#### Solution 1\\n")
```

```{r tapply1-solution, solution = TRUE}
tapply(heart$year, heart$transplant, median)
tapply(retinopathy$futime, retinopathy$status, median)
```



```{r, eval = static, echo = FALSE}
asis_output("#### Task 2\\n")
```

```{r, eval = static, echo = FALSE}
asis_output('<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">')
```

* Apply the function `Fun1 <- function(x) { mean(x)/(length(x) - 2) }` to `year` per `transplant` and `surgery` group using the **heart** data set. \
* Obtain the mean `futime` per `status`, `type` and `trt` group using the **retinopathy** data set. \

```{r, eval = static, echo = FALSE}
asis_output("</div>")

asis_output("#### Solution 2\\n")
```

```{r tapply2-solution, solution = TRUE}
Fun1 <- function(x) { mean(x)/(length(x) - 2) }
tapply(heart$year, list(heart$transplant, heart$surgery), Fun1)
tapply(retinopathy$futime, list(retinopathy$status, retinopathy$type, retinopathy$trt), mean)
```
