---
title: "Statistical Report"
author: "Eleni-Rosalina Andrinopoulou"
date: "`r Sys.setenv(LANG = 'en_US.UTF-8'); format(Sys.Date(), '%d %B %Y')`"
output:
  word_document: default
---

# Statistical Analysis

* We are going to used the pbc data set from the survival library. 

* Continuous data are presented as mean and standard deviation while categorical variables as counts and percentages. Differences between groups were analyzed by the unpaired Student's t tests for continuous variables while.... To investigate the association between serum bilirubin with sex, age and time, a linear regression was performed... A p-value <0.05 was considered statistically significant.

```{r, echo=TRUE}
library(lattice)
library(knitr)
library(survival)
library(effects)
```

```{r, results='asis', echo=T}
R.Version()$version.string
packageVersion("lattice")
packageVersion("knitr")
packageVersion("survival")
```

The t-test comparing the differences in serum bilirubin in males and females gave up a p-value of:

```{r, echo=FALSE}
t.test(pbc$bili ~ pbc$sex)$p.value
```

```{r, echo=FALSE}
fm1 <- lm(bili ~ sex + age + time, data = pbc)
tab <- summary(fm1)$coefficients
```

```{r, echo=FALSE} 
kable(tab, dig = 4)
```

## Predict weight for 2 male patients with min and max BMI at different ages
```{r, echo=TRUE}
plot(effect(fm1))
```

```{r, echo=FALSE, include=TRUE}
plot(pbc$bili, pbc$time, xlab = "time", ylab = "bili")
```

```{r, echo=FALSE}
xyplot(pred ~ age | time, type = "l", lty = c(1, 2, 2), cex = 1.8,  col = 1, lwd = 2, main = "age vs bili per time group", xlab = "age", ylab = "bili", data = newdata)
```


# Conclusions

We may conclude that:

* A strong association was found between time and serum bilirubin
* A weak association was found between sex and serum bilirubin
  * The same holds for age


Strengths and weknesses of our study are:

1. Small sample size
2. No simulation study was performed

