---
title: "Report for the paper ..."
author: "The Author"
date: "28-02-2019"
output:
  word_document: default
  pdf_document: default
  html_document:
    df_print: paged
---

# Statistical Analysis

* We are going to used the pbc data set from the survival library. 

* Continuous data are presented as mean and standard deviation while categorical variables as counts and percentages. Differences between groups were analyzed by the unpaired Student's t tests for continuous variables while.... To investigate the association between serum bilirubin with sex, age and time, a linear regression was performed... A p-value <0.05 was considered statistically significant.

```{r, echo=TRUE}
library(lattice)
library(knitr)
library(survival)
```

```{r, results='asis', echo=T}
R.Version()$version.string
packageVersion("lattice")
packageVersion("knitr")
packageVersion("survival")
```


```{r, echo=FALSE}
fm1 <- lm(bili ~ sex + age + time, data = pbc)
tab <- summary(fm1)$coefficients
```

```{r, echo=FALSE} 
kable(tab, dig = 4)
```

## Predict weight for 2 male patients with min and max BMI at different ages
```{r, echo=TRUE}
newdata <- with(pbc, data.frame(
  sex = rep(1, each = 40),
  age = rep(seq(min(pbc$age), max(pbc$age), length = 20), each = 2),
  time = rep(c(min(pbc$time), max(pbc$time)), 20)
))

X <- model.matrix(~ sex + age + time, data = newdata)

betas <- summary(fm1)$coefficients[, 1]
newdata$pred <- c(X %*% betas)

newdata$time <- factor(newdata$time, labels = c("min time","max time"))
```

```{r, echo=FALSE, include=TRUE}
plot(pbc$bili, pbc$time, xlab = "time", ylab = "bili")
```

```{r, echo=FALSE}
xyplot(pred ~ age | time, type = "l", lty = c(1, 2, 2), cex = 1.8,  col = 1, lwd = 2, main = "age vs bili per time group", xlab = "age", ylab = "bili", data = newdata)
```


# Conclusions

We may conclude that:

* Conclusion 1
* Conclusion 2
  * Note 2b
* Conclusion 3  


Strengths and weknesses of our study are:

1. It is really strong
2. I mean *really*, really strong

(a) And __not__ weak at all
(b) because it is strong
