---
title: 'Linear Regression'
output: 
  html_document:
    code_folding: show
    df_print: paged
    toc: true
    toc_float:
      collapsed: false
    number_sections: false
    theme: spacelab
    highlight: tango
    includes:
      after_body: ../footerNE.html
    css: ../style.css
---

```{r, include = FALSE}
library(knitr)
options(width = 100)

hint <- function(text) {
  id <- paste0('hint_', sample.int(1e10, size = 1))
    cat(
    paste0('\n',
      '<button type="button" class="btn btn-info btn-sm" data-toggle="collapse" ',
      'data-target="#', id, '">Hint</button>',
      '<div id = "', id, '" class="collapse" style="border:1px; ',
      'border-style:solid; padding: 1em; border-color:#1F78B4">',
      text, '</div>')
  )
}
set.seed(2020)
```


## Dataset 
For this practical, we will use the **birthwt** dataset from the package **MASS**.

You can either load the **MASS** package to get access to this dataset
```{r}
library(MASS)
```
or make a copy of this data
```{r}
birthwt <- MASS::birthwt
```


## Data
The **birthwt** data have the following 10 variables:
```{r}
str(birthwt)
```

Some of the variables are actually categorical variables, but we will not yet
re-code them to factors in order to investigate how the variable coding can change
the results.

## Fitting a Linear Regression{.tabset .tabset-fade .tabset-pills}
### Taks1
<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">
Write the model formula with 

* outcome `bwt` (birth weight) and 
* covariates 
  * `age` (mother's age), 
  * `lwt` (mothers weight in pounds), 
  * `race` (mother's race) and 
  * `smoke` (smoking status during pregnancy).
```{r, echo = FALSE, results = 'asis'}
hint('Use the `~` to separate the outcome from the linear predictor and use `+`
     to separate the covariates.')
```
</div>

### Solution 1
```{r}
bwt ~ age + lwt + race + smoke
```

### Task 2
<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">
Fit a linear regression model with the formula that we have just specified.
```{r, echo = FALSE, results = 'asis'}
hint("Use the function `lm()`.")
hint("Besides the model formula, you also need to provide the data via the argument `data`.")
```
</div>

### Solution 2
```{r}
mod1 <- lm(bwt ~ age + lwt + race + smoke, data = birthwt)
```
Note that if we don't assign a name to the model, the results are just printed,
but we cannot use or investigate them any further.

## Model Summary {.tabset .tabset-pills .tabset-fade}
### Task 1
<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">
Get the `summary()` of the model.
```{r, echo = FALSE, results = 'asis'}
hint("Use the function `lm()`.")
hint("Besides the model formula, you also need to provide the data via the argument `data`.")
```
</div>

### Solution 1
```{r}
summary(mod1)
```
```{r, echo = FALSE}
smod1 <- capture.output(summary(mod1))
```

### Task 2
<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">
Investigate the model summary closely.

* What are the estimated effects of the coefficients?
* Confirm that you identified the correct part of the output by extracting only
 the regression coefficients from the model (not from the model summary!). 
```{r, echo = FALSE, results = 'asis'}
hint("To extract the coefficients, use the function `coef()`.")
```
</div>

### Solution 2
The regression coefficients are given in the first column ("Estimate") of the
output under the header "Coefficients".
```{r, echo = FALSE}
cat("[...]\n", paste0(smod1[9:15], collapse = "\n"), "\n[...]")
```

They are the same as:
```{r}
coef(mod1)
```

### Task 3
<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">

* Where in the summary can you find the p-values?
* Which of the covariates have a statistically significant effect (for $\alpha = 5%$)?
</div>

### Solution 3
The p-values are given in the last column ("Pr(>|t|)") of the output under the 
header "Coefficients".
We can see that the intercept and the effects for `race` and `smoke` significantly
differ from 0.

### Task 4
<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">
What does the model summary tell you about the model fit?
```{r, echo = FALSE, results='asis'}
hint("The model summary contains the $R^2$ and adjusted $R^2$.")
hint("The model summary contains info on an omnibus test (F-statistic).")
```
</div>

### Solution 4
In the lowest part of the output of `summary()`:
```{r, echo = FALSE}
cat("[...]\n", smod1[20], "\n[...]")
```
We can read there that the
R-squared is `r format(summary(mod1)$r.squared, digits = 4)` and that the
adjusted R-squared is `r format(summary(mod1)$adj.r.squared, digits = 4)`.

In the last row we see the F-statistic which tests the model against 
the null-model (intercept only). 
```{r, echo = FALSE}
cat("[...]\n", smod1[21])
```
It shows that the model explains significantly
more variation in the outcome than the null-model (p-value: 
`r format.pval(pf(summary(mod1)$fstatistic[1L], 
                summary(mod1)$fstatistic[2L], summary(mod1)$fstatistic[3L],
               lower.tail = FALSE), digits = 4)`)

## Confidence Intervals {.tabset .tabset-pills .tabset-fade}
### Task 1
<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">
Get the 90% confidence intervals of the model.
```{r, echo = FALSE, results = 'asis'}
hint("Use the function `confint()`.")
```
</div>

### Solution 1
```{r}
confint(mod1, level = 0.9)
```

### Task 2
<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">
Combine the coefficients and 95% confidence intervals of the model into one
`matrix`.
```{r, echo = FALSE, resuts = 'asis'}
cbind(estimate = coef(mod1),
      confint(mod1))
```
Note:

Because `coef()` returns a vector, the corresponding column in our marix would not
have a name. We can assign the name when we use `cbind()`.

`confint()` returns a `matrix` which already has column names.
</div>

## Coding of the covariates  {.tabset .tabset-pills .tabset-fade}
Now we will re-code those covariates that are actually categorical variables.

### Task 1
<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">
Re-code `race` and `smoke` as `factor`.
```{r, echo = FALSE, results = 'asis'}
hint('Use the function `factor()`.')
```
</div>

### Solution 1
```{r}
birthwt$race <- factor(birthwt$race)
birthwt$smoke <- factor(birthwt$smoke)
```

### Task 2
<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">

* Fit the same model as before, but with the re-coded data and give it a new name.
* Print the model summary.

How did the results change?
</div>

### Solution 2
```{r}
mod2 <- lm(bwt ~ age + lwt + race + smoke, data = birthwt)
summary(mod2)
```
```{r, echo = FALSE}
smod2 <- capture.output(summary(mod2))
```

When we compare the "Coefficients" part of the summary for `mod2`
```{r, echo = FALSE}
cat(paste0(smod2[9:16], collapse = "\n"))
```
with the "Coefficients" part of the summary for `mod1`
```{r, echo = FALSE}
cat(paste0(smod1[9:15], collapse = "\n"))
```
we notice that in the summary for `mod2` we have coefficients for `race2` and `race3` where 
in the summary for `mod1` we only had one coefficient, and that the name of the coefficient
for `smoke` is now `smoke1`.

<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">
* R automatically converts a `factor` into dummy variables!
* The **first category** is used as **reference** category.
* The other factor levels are appended to the variable name to identify the different dummy variables.
</div>

Note:
Because we now have more parameters, of course also the results for model fit slightly changed.

### Task 3
<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">
What is the difference in the interpretation of the effects for `race` and `smoke` in `mod1` and `mod2`?
</div>

### Solution 3
For `smoke` there is no real difference. `smoke` has only two categories, therefore there is one dummy
variable `smoke1` which is 1 if `smoke == 1` and 0 otherwise. Since this is the same as the original 
variable `smoke`, nothing really changes.

For `race`, the assumption in `mod1` is that the effect of category `3` compared to `1` is 
twice as large as the effect of category `2` compared to `1` because `race` is treated as continuous
and the coefficient is the effect of a "one unit increase" (and `3` vs `1` is a two unit increase).

In `mod2` there are two dummy variables for `race`, which measure the effect of each category compared
to the reference category `1`. The effects of category `3` is not assumed to be double the effect of
category `2`.



