---
title: 'Logistic Regression: Additional Exercise'
output: 
  html_document:
    code_folding: show
    df_print: paged
    toc: true
    toc_float:
      collapsed: false
    number_sections: false
    theme: spacelab
    highlight: tango
    includes:
      after_body: ../footerNE.html
    css: ../style.css
---

```{r, include = FALSE}
library(knitr)
options(width = 100)

hint <- function(text) {
  id <- paste0('hint_', sample.int(1e10, size = 1))
    cat(
    paste0('\n',
      '<button type="button" class="btn btn-info btn-sm" data-toggle="collapse" ',
      'data-target="#', id, '">Hint</button>',
      '<div id = "', id, '" class="collapse" style="border:1px; ',
      'border-style:solid; padding: 1em; border-color:#1F78B4">',
      text, '</div>')
  )
}
set.seed(2020)
```


## Preface
### R packages
In this practical, we use the R package **ggplot2**.
It can be installed by running the following syntax:
```{r, eval = FALSE}
install.packages("ggplot2")
```


### Dataset 
Like in the previous practical on logistic regression, we will use the **pbc** 
dataset from the package **survival**.

You can either load the **survival** package to get access to this dataset
```{r}
library(survival)
```
or make a copy of this data
```{r}
pbc <- survival::pbc
```

The **pbc** data contain (among others) the following variables:
```{r}
str(pbc[, c(5:12, 20)])
```

The relevant variables are

* `age`: patient's age in years
* `sex`: patient's sex
* `ascites`: presence of ascites
* `spiders`: presence of blood vessel malformations in the skin
* `bili`: serum bilirubin
* `chol`: serum cholesterol
* `stage`: histologic stage of the disease


## Exploring problems in logistic regression {.tabset .tabset-pills .tabset-fade}
In the previous practical, we have re-coded some of the categorical variables 
and fitted the logistic regression model. 

If you are using a new R session, you need to first run the following syntax:
```{r}
pbc$stage <- factor(pbc$stage)
pbc$spiders <- factor(pbc$spiders)

mod1 <- glm(spiders ~ age + sex + bili + stage + ascites, data = pbc,
            family = binomial())
```

We would now like to add another continuous variable to `mod1`: the variable `chol`.

### Task 1
<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">

* Update `mod1` to include `chol`.
* Investigate how the results change.
</div>


### Solution 1
```{r}
mod_chol <- update(mod1, formula = . ~ . + chol)
summary(mod_chol)
```

We can compare the coefficients form `mod1` and `mod_chol` and notice that the 
`(Intercept)`, and the coefficients for `stage` are much more extreme than the
ones in `mod1`:

```{r}
round(
  cbind(mod1 = coef(mod1)[names(coef(mod_chol))],
        mod_chol = coef(mod_chol)
  ), 3)
```

<div style="border:2px; border-style:solid; padding: 1em; border-color:#009B8F">
<p style = "color:#009B8F; font-weight:bold;">Note:</p>
If we would just use `cbind()` to combine the coefficients into a matrix we
could not be sure that the coefficients of the same variable are in the same row.
```{r}
a <- c(x = 1, y = 2, z = 3)
b <- c(x = 1, z = 3, y = 2)
cbind(a, b)
```

Moreover, the "missing" effect for `chol` in `mod1` will just be filled in with
the first value of the vector.
```{r}
cbind(coef(mod1), coef(mod_chol))
```

By using `coef(mod1)[names(coef(mod_chol))]` we make sure that the coefficients
from `mod1` are given in the same order as the coefficients from `mod_chol`. Since
there is no `chol` in `coef(mod1)`, `NA` is filled in:
```{r}
coef(mod1)[names(coef(mod_chol))]
```
</div>


### Visualization
<div style="border:2px; border-style:solid; padding: 1em; border-color:#009B8F">
To compare the coefficients from different models it is useful to plot coefficients
with their corresponding confidence intervals.

**Using base R plots:**
```{r, warning = FALSE}
# coefficients & CI from mod1
res1 <- cbind(coef = coef(mod1), confint(mod1))
res1 <- rbind(res1, chol = c(NA, NA, NA)) # add an "empty" row for chol

# coefficients & CI from mod_chol
res3 <- cbind(coef = coef(mod_chol), confint(mod_chol))

# graphical settings:
# * mfrow = c(3, 3) print the plots in a grid with 3 rows and 3 columns
# * mar = c(3, 3, 2, 1): adjust the plot margins (looks nicer)
par(mfrow = c(3, 3), mar = c(3, 3, 2, 1))

for (k in rownames(res1)) {                               # loop over all variables
  plot(x = c(1, 2),                                       # position on the x-axis
       y = c(res1[k, 'coef'], res3[k, 'coef']),           # coefficient value on the y-axis
       main = k,                                          # set variable name as title
       xlim = c(0, 3),                                    # set limit of the x-axis (looks nicer)
       ylim = range(res1[k, ], res3[k, ], na.rm = TRUE),  # set the limits of the y-axis
                                                          # so that also the CIs fit in
       xaxt = 'n',                                        # don't plot the x-axis (we want our own version)
       ylab = 'log OR & 95% CI')                          # specify the y-axis label
  
  # specification of the x-axis
  axis(side = 1,                    # draw axis at the bottom of the plot
       at = c(1, 2),                # locations of the tick-marks on the x-axis
       labels = c('mod1', 'mod_chol'))  # labels of the tick-marks
  
  # add vertical lines representing the confidence intervals
  lines(x = c(1, 1), y = res1[k, c('2.5 %', '97.5 %')])
  lines(x = c(2, 2), y = res3[k, c('2.5 %', '97.5 %')])
}
```


**Using ggplot2:**
```{r, warning = FALSE}
results <- rbind(
  # data.frame containing results from mod1
  data.frame(coef = coef(mod1), 
             confint(mod1), 
             variable = names(coef(mod1)),
             model = 'mod1',
             check.names = FALSE),
  
  # data.frame containing results from mod_chol
  data.frame(coef = coef(mod_chol), 
             confint(mod_chol), 
             variable = names(coef(mod_chol)),
             model = 'mod_chol',
             check.names = FALSE)
)
```
```{r}
library("ggplot2")
ggplot(results,                           # specification of the data
       aes(x = model, y = coef)) +        # specify what goes on which axis
  geom_point() +                          # plot points for the coefficients
  geom_errorbar(aes(ymin = `2.5 %`,       # add confidence intervals
                    ymax = `97.5 %`),
                width = 0.3) +            # width of the horizontal lines
  facet_wrap('variable',                  # make a separate plot per variable
             scales = 'free_y')           # allow the y-axes to vary between variables
```
</div>

### Task 2
Something is clearly going wrong with the variable `stage`.

<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">
Investigate what is going on by looking at `stage` in relation to the outcome `spiders`.
```{r, echo = FALSE, results = 'asis'}
hint("Use a `table()`.")
```
</div>

### Solution 2
```{r}
table(stage = pbc$stage,
      spiders = pbc$spiders, exclude = NULL)
```
There are very few observations with `spiders` for `stage = 1` and `stage = 2`.

### Task 3
We saw that there is almost perfect separation between `spiders` for the patients
with `stage = 1`, but as long as there is a patient, for every combination of
covariate and outcome, we should not get such non-sense results.

<br>
We may have overlooked something when we created the table.

<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">
Check if the table did indeed include those cases that are analysed in `mod_chol`
by checking if the table at least contains the correct number of cases.
```{r, echo = FALSE, results = 'asis'}
hint('You can find relevant info in the output of the summary().')
```
</div>

### Solution 3
```{r}
sum(table(stage = pbc$stage, spiders = pbc$spiders))
```
In the table, we have `r sum(table(stage = pbc$stage, spiders = pbc$spiders))`
cases where both `spiders` and `stage` are observed.

The output of `summary(mod_chol)` showed the following:
```{r, echo = FALSE}
smod_chol <- capture.output(summary(mod_chol))
cat(paste("[...]", paste0(smod_chol[26:31], collapse = '\n'), sep = "\n"))
```
`r length(summary(mod_chol)$na.action)` observations were excluded.

In the full **pbc** data we have `nrow(pbc) = `r nrow(pbc)`` observations.

We should, hence, be looking at only `418 - 134 = `r  418 - 134`` observations!

There are some other variables in our model that have missing values!

<div style="border:2px; border-style:solid; padding: 1em; border-color:#009B8F">
<p style = "color:#009B8F; font-weight:bold;">Note:</p>
The output of `summary(mod_chol)`) also shows that the deviance of the null model
(the model with only intercept) had 283 degrees of freedom, which is N - 1.
We could also derive from there that there were 283 + 1 cases in the analysis.
</div>


### Task 4
We need to exclude all cases that have missing values in any of the variables
involved in `mod_chol` and check the table of `stage` and `spiders` again.

<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">

* Check which variables have missing values.
* Make a subset excluding those cases.
* Create the table of `stage` vs `spiders` using the subset.
</div>


### Solution 4
A convenient way to obtain all variables from a model is to use the function
`all.vars()` on the model formula:
```{r}
all.vars(formula(mod_chol))
```
We can use this to get a summary of exactly those columns of the **pbc** data:
```{r}
summary(pbc[, all.vars(formula(mod_chol))])
```

<div style="border:2px; border-style:solid; padding: 1em; border-color:#009B8F">
<p style = "color:#009B8F; font-weight:bold;">Note:</p>
Alternatively, we could have looked at the sum of missing values in each column:
```{r}
colSums(is.na(pbc[, all.vars(formula(mod_chol))]))
```
</div>
<br>

We can make a subset in which we exclude cases with missing values in variables
other than `stage` and `spiders`.
(We could also exclude the cases with missing values in those two variables,
but they are shown in separate columns/rows in the table anyway.)
```{r}
pbc_sub <- subset(pbc, !is.na(chol) & !is.na(stage))

# or:
# pbc_sub <- pbc[, !is.na(ascites) & !is.na(chol)]

table(pbc_sub$stage, pbc_sub$spiders, exclude = NULL)
```
Indeed, there are no subjects with `spiders` in the `stage = 1` group in the
analysis. This explains the inflated effect estimates and standard errors.


## Changing the reference category {.tabset .tabset-pills .tabset-fade}
Due to the dummy coding in which `stage = 1` was the reference category,
there was no information available to estimate any of the `stage` dummy effects,
because they are relative to `stage = 1`.

### Task 1
<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">
Change the reference category of `stage` to another level and re-fit `mod_chol`.
</div>

### Solution 1
We choose category 4 as reference category.
There are (at least) two options to change the reference category:
```{r, collapse=TRUE}
pbc$stage_rev <- factor(pbc$stage, levels = 4:1, labels = 4:1)
pbc$stage_ref4 <- relevel(pbc$stage, ref = 4)

levels(pbc$stage)
levels(pbc$stage_ref4)
levels(pbc$stage_rev)
```
Even though `stage_ref4` and `stage_rev` have different ordering of their levels
(but thy are not ordered factors!), they give the same result when used as covariates
in a regression model, since all non-reference categories are only compared with the
reference category.


```{r}
mod_chol2 <- glm(spiders ~ age + sex + chol + bili + stage_rev + ascites, 
                 data = pbc, family = binomial())
summary(mod_chol2)
```

We see that the effect estimates and standard errors of levels 2 and 3 are back
to normal values.

```{r}
library(effects)
plot(predictorEffects(mod_chol2))
```

