---
title: 'Logistic Regression & Non-linear Effects'
output: 
  html_document:
    code_folding: show
    df_print: paged
    toc: true
    toc_float:
      collapsed: false
    number_sections: false
    theme: spacelab
    highlight: tango
    includes:
      after_body: ../footerNE.html
    css: ../style.css
---

```{r, include = FALSE}
library(knitr)
options(width = 100)

hint <- function(text) {
  id <- paste0('hint_', sample.int(1e10, size = 1))
    cat(
    paste0('\n',
      '<button type="button" class="btn btn-info btn-sm" data-toggle="collapse" ',
      'data-target="#', id, '">Hint</button>',
      '<div id = "', id, '" class="collapse" style="border:1px; ',
      'border-style:solid; padding: 1em; border-color:#1F78B4">',
      text, '</div>')
  )
}
set.seed(2020)
```


## Preface
### R packages
In this practical, we use the R package **effects**.
It can be installed by running the following syntax:
```{r, eval = FALSE}
install.packages("effects")
```


### Dataset 
For this practical, we will use the **pbc** dataset on Primary Biliary Cholangitis
from the package **survival**.

You can either load the **survival** package to get access to this dataset
```{r}
library(survival)
```
or make a copy of this data
```{r}
pbc <- survival::pbc
```

The **pbc** data contain (among others) the following variables:
```{r}
str(pbc[, c(5:11, 20)])
```

The relevant variables are

* `age`: patient's age in years
* `sex`: patient's sex
* `ascites`: presence of ascites
* `spiders`: presence of blood vessel malformations in the skin
* `bili`: serum bilirubin
* `stage`: histologic stage of the disease



## Fitting a Logistic Regression

We want to fit a logistic regression model for the outcome `spiders` using the 
covariates `age`, `sex`, `bili`, `stage` and `ascites`.

### Variable coding {.tabset .tabset-fade .tabset-pills}

#### Task 1

<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">
Are the variables coded correctly?
If not, re-code them.
```{r, echo = FALSE, results = 'asis'}
hint('Three of the variables of interest are coded as integer.')
```

</div>

#### Solution 1
The variables `ascites`, `spiders` and `stage` are coded as integer. Both
`ascites` and `spiders` do only have two levels, therefore we *could* use them
as they are:
```{r}
table(pbc$spiders, exclude = NULL)
table(pbc$ascites, exclude = NULL)

#  In this case optional:
pbc$spiders <- factor(pbc$spiders)
```

`stage` has four levels and should be re-coded as a factor:
```{r}
table(pbc$stage, exclude = NULL)
pbc$stage <- factor(pbc$stage)
```


#### Task 2
<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">
Fit the logistic model described above and get the model summary.
```{r, echo = FALSE, results = 'asis'}
hint("A logistic regression model is a GLM.")
hint("You need to specify the model `formula`, the `data` and the `family`.")
```
</div>

#### Solution 2
```{r}
mod1 <- glm(spiders ~ age + sex + bili + stage + ascites, data = pbc,
            family = binomial())
summary(mod1)
```


#### Task 3

<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">

Investigate the output of the model summary.

* Were there missing values in the data? What happened with those cases?
* What is the interpretation of the coefficient for `bili` in this model?
* What is the interpretation of the coefficients for `stage`?
</div>



#### Solution 3
**Missing values:**<br>
In the lower part of the output of the model summary we can read:
```{r, echo = FALSE}
smod1 <- capture.output(summary(mod1))
cat("[...]\n",
    paste0(smod1[25:28], sep = "\n"),
    "[...]")
```

There were `r  length(summary(mod1)$na.action)` cases with missing values. These
were excluded from the analysis.


<br><br>
**Interpretation of the effect of `bili`:**
```{r, echo = FALSE}
cat(paste0(smod1[10:11], sep = "\n"),
    "[...]\n",
    paste0(smod1[15], sep = "\n"),
    "[...]"
    )
```

With every 1 unite increase of `bili` the expected log odds of having `spiders`,
i.e., $\log\left(\frac{P(\texttt{spiders = 1})}{1 - P(\texttt{spiders=1})}\right)$, increases by
$\beta_{\texttt{bili}} = `r sprintf("%.3f", coef(mod1)['bili'])`$ (while all other
covariates in the model are kept constant).

This means that the odds ratio of `bili` is
$\exp(\beta_{\texttt{bili}}) = \exp(`r sprintf("%.3f", coef(mod1)['bili'])`) = `r sprintf("%.3f", exp(coef(mod1)['bili']))`$.

<br><br>
**Interpretation of the effect(s) of `stage`:**
```{r, echo = FALSE}
cat(paste0(smod1[10:11], sep = "\n"),
    "[...]\n",
    paste0(smod1[16:18], sep = "\n"),
    "[...]"
    )
```

A patient with `stage = 2` has a $\beta_{\texttt{stage2}} = `r sprintf("%.3f", coef(mod1)['stage2'])`$
higher log odds to have `spiders` than a patient with `stage = 1` (and the same
values for all other covariates).<br>
The odds ratio of `stage = 2` is 
$\exp(\beta_{\texttt{stage2}}) = \exp(`r sprintf("%.3f", coef(mod1)['stage2'])`) = `r sprintf("%.3f", exp(coef(mod1)['stage2']))`$.

Patients with `stage = 3` or `stage = 4` have a 
$\beta_{\texttt{stage3}} = `r sprintf("%.3f", coef(mod1)['stage3'])`$ and
$\beta_{\texttt{stage4}} = `r sprintf("%.3f", coef(mod1)['stage4'])`$
higher log odds to have `spiders` than a patient with `stage = 1` 
(who has the same values for all other covariates).<br>
The corresponding odds ratios of `stage = 3` and `stage = 4` are 
$\exp(\beta_{\texttt{stage3}}) = \exp(`r sprintf("%.3f", coef(mod1)['stage3'])`) = `r sprintf("%.3f", exp(coef(mod1)['stage3']))`$ and
$\exp(\beta_{\texttt{stage4}}) = \exp(`r sprintf("%.3f", coef(mod1)['stage4'])`) = `r sprintf("%.3f", exp(coef(mod1)['stage4']))`$, 
respectively.


### Comparing Models {.tabset .tabset-fade .tabset-pills}
We would like to know if the variable `stage` has a significant/relevant
contribution to the model.

#### Task 1
<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">
How can we determine this? Does the output of the `summary()` give us an answer?
</div>

#### Solution 1
Since `stage` is a factor with >2 levels, multiple dummy variables are used in the
model. To answer the question whether `stage` has a significant/relevant contribution
we need to consider all three dummy variables simultaneously.

Therefore, we need to compare `mod1` with the model in which `stage` was removed.

#### Task 2
<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">
Re-fit `mod1` without `stage` (but give it a different name).
</div>


#### Solution 2
```{r}
mod1b <- glm(spiders ~ age + sex + bili + ascites, data = pbc,
             family = binomial())

# We could also use:
mod1b <- update(mod1, formula = . ~ . - stage)
```

<div style="border:2px; border-style:solid; padding: 1em; border-color:#009B8F">
<p style = "color:#009B8F; font-weight:bold;">Note:</p>
The function `update()` allows us to update models or formulas.

* To update a model, specify the new values for those arguments that change.
* To update a formula, we can use the `.` to say 'as it was' and only add the change.
</div>

#### Task 3
<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">
* Are the models `mod1` and `mod1b` nested or not?
* Which method can we use to  compare them?
* Perform this comparison and interpret the result.
</div>


#### Solution 3
The models are nested. `mod1b` is a specific case of `mod1` for which 
$\beta_{\texttt{stage2}} = \beta_{\texttt{stage3}} = \beta_{\texttt{stage4}} = 0$ 

To compare nested models, we can use a likelihood ratio test:
```{r, eval = FALSE}
anova(mod1, mod1b, test = 'LRT')
```

```{r, echo = FALSE}
cat(paste0(capture.output(anova(mod1, mod1b, test = 'LRT')), collapse = "\n"))
```

The small p-value indicates that the observed difference in residual deviance 
between the two models would be very unlikely to observe if `mod1b` was the
correct model.

Hence, the model with `stage` fits the data better.


## Non-linear effects {.tabset .tabset-pills .tabset-fade}
We now want to investigate if the continuous variables have non-linear effect.

### Task 1
<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">
To do this, use a natural cubic splines with three degrees of freedom. 
```{r, echo = FALSE, results = 'asis'}
hint('Use the function `ns()` from the package `splines`.')
```

</div>

### Solution 1
```{r}
library(splines)
mod2 <- glm(spiders ~ ns(age, df= 3) + sex + ns(bili, df = 3) + stage + ascites,
            data = pbc, family = binomial())

summary(mod2)
```


### Task 2
<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">
What is the interpretation of the regression coefficients for `bili` in `mod2`?
</div>

### Solution 2
The coefficients tell us how much each of the basis functions of the splines 
have been scaled. They do not have a direct meaningful clinical interpretation.

### Visualization
<div style="border:2px; border-style:solid; padding: 1em; border-color:#009B8F">
To visualize the resulting non-linear fit, we can use an effect plot.

The idea is to obtain fitted values for a specific set of cases, which vary in the
value of the covariate of interest, but have constant values for all other 
covariates.

The R package **effects** can help us in creating such plots:
```{r, fig.width = 12, fig.height = 8, message=FALSE}
# if not installed, run:
# install.packages("effects")

library("effects")
plot(predictorEffects(mod2))
```

From the **effects** package documentation:

>  [...] the vertical axis is in linear predictor scale, which is the log-odds or
logit for this logistic regression example, but the vertical axis labels are in mean (probability) scale,
so the tick-marks are not equally spaced

The plot for `age`, for example, shows us the estimated probability 
(and corresponding 95% confidence band) for having `spiders` for any given 
`age` while the other variables have been fixed to the population mean.


For a more information about the **effects** package and its functions, 
please refer to the reference manual and vignettes available at
[https://CRAN.R-project.org/package=effects](https://CRAN.R-project.org/package=effects).
</div>

### Task 3
The effect plots show that the effect of `age` is quite linear, but that the
effect of `bili` is indeed non-linear.

<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">
Confirm that the non-linear effect for `age` is not needed, i.e., does not fit
the data better than the corresponding model with a linear effect of `age`.
```{r, echo = FALSE, results = 'asis'}
hint("Re-fit the model and replace the spline for `age` with just `age` to compare the model fit.")
hint('For the comparison, use the `AIC()` or `BIC()`.')
```
</div>

### Solution 3
```{r}
mod2b <- glm(spiders ~ age + sex + ns(bili, df = 3) + stage + ascites,
             data = pbc, family = binomial())
```

When comparing models with regards to the effect of a variable that is modelled
using a spline, models are usually not nested (except in specific cases). therefore,
we will use here the information criteria to compare `mod2` and `mod2b`:


```{r, eval = FALSE}
BIC(mod2, mod2b)
```
```{r, echo = FALSE}
cat(paste0(capture.output(BIC(mod2, mod2b)), collapse = "\n"))
```

We see that the smaller (less degrees of freedom / parameters) model has the lower BIC.
This confirms that the spline specification for `age` is not necessary.


### Task 4
The non-linear association for `bili` resembles the shape of the logarithmic function.

<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">
Investigate if we can replace the spline for `bili` with the (natural) logarithm.

* using information criteria
* visually
```{r, echo = FALSE, results = 'asis'}
hint('For the visual comparison, the function 
`predictorEffect(predictor = "bili", mod = <name of model>)` may be useful if 
you want to show the effect plot for the variable `bili` only.')
```
</div>

### Solution 4
```{r}
mod2c <- glm(spiders ~ age + sex + log(bili) + stage + ascites,
             data = pbc, family = binomial())
```
```{r, eval = FALSE}
BIC(mod2b, mod2c)
```
```{r, echo = FALSE}
cat(paste0(capture.output(BIC(mod2b, mod2c)), collapse = '\n'))
```

The BIC shows that the model with `log(bili)` has better model fit.

```{r, fig.width = 5, fig.height = 5, out.width = "50%", fig.align='center', collapse=TRUE}
plot(predictorEffect('bili', mod2b))
plot(predictorEffect('bili', mod2c))
```


## Additional Exercises

### Ordinal Covariate {.tabset .tabset-pills .tabset-fade}

The variable `stage` is actually ordinal, but when we re-coded it above, we
did not consider that.

#### Task 1
<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">
Make a an ordered factor from `stage` and call the new variable `stage_ord`.
</div>

#### Solution 1
```{r}
pbc$stage_ord <- as.ordered(pbc$stage)

# or alternatively:
# pbc$stage_ord <- factor(pbc$stage, ordered = TRUE)
```

<div style="border:2px; border-style:solid; padding: 1em; border-color:#009B8F">
<p style = "color:#009B8F; font-weight:bold;">Note:</p>
R automatically orders factors by their level (alphabetically or by number):
```{r}
factor(c('low', 'medium', 'high'), ordered = TRUE)
```

For ordered factors where the levels should not be ordered by alphabet (or number)
you need to specify the `levels` in `factor()`:
```{r}
factor(c('low', 'medium', 'high'), levels = c('low', 'medium', 'high'), ordered = TRUE)
```
</div>

#### Task 2
<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">
* Fit another version of `mod1` using `stage_ord` instead of `stage`.<br>
  Do not copy-paste the syntax from above, but use `update()` to change the model.
* What is the interpretation of the regression coefficients for `stage_ord`?
</div>

#### Solution 2
To update `mod1`, we need to specify that we want to change the `formula`.
We remove the original `stage` and add `stage_ord`:
```{r}
mod1c <- update(mod1, formula = . ~ . -stage + stage_ord)
summary(mod1c)
```
The summary of the coefficients for `stage_ord` looks a bit different than the
corresponding part in `mod1`. We now have:
```{r, echo = FALSE}
smod1c <- capture.output(summary(mod1c))

cat(paste0(smod1c[10], "\n[...]\n",
       paste0(smod1c[17:19], collapse = "\n"),
       "\n[...]", collapse = '\n')
)
```
where in `mod1` we had:
```{r, echo = FALSE}
cat(paste0(smod1[10], "\n[...]\n",
       paste0(smod1[16:18], collapse = "\n"),
       "\n[...]", collapse = '\n')
)
```

When R creates dummy variables, the factor levels are added at the end of the
variable name. For ordered factors, R uses by default **orthogonal polynomials**, 
i.e., in this case a **l**inear, a **q**uadratic and a **c**ubic term.

You can obtain the coding for the (ordered) factor using `contrasts()`: 
```{r}
contrasts(pbc$stage_ord)
```
These values are values from a linear, quadratic and cubic curve at the numeric
values of the factor levels:
```{r, echo = FALSE, fig.width = 6, fig.height = 4, out.width = "80%", fig.align = 'center'}
x <- seq(1, 4, length = 100)
p4 <- poly(1:4, degree = 3)

library(ggplot2)
plotdat <- reshape2::melt(as.data.frame(cbind(x = x, predict(p4, newdata = x))),
                          id.vars = 'x')

textdat <- reshape2::melt(as.data.frame(cbind(x = 1:4, p4)),
                          id.vars = 'x')
textdat$label <- c('.L', '.Q', '.C')[textdat$variable]

ggplot(plotdat, aes(x = x, y = value, color = variable)) +
  geom_line(size = 1) +
  theme(panel.background = element_rect(fill = grey(0.95), color = grey(0.8)),
        panel.grid = element_blank()) +
  scale_color_viridis_d(guide = FALSE, end = 0.8, begin = 0, option = 'D') +
  geom_label(data = textdat, aes(label = label), label.size = 0, size = 4) +
  xlab('numeric factor level') +
  ylab('orthogonal polynomial') +
  scale_y_continuous(breaks = seq(-1, 1, 0.25))


```

The coefficients in the `summary()` output are thus the effect estimates for
these linear, quadratic and cubic curves and **cannot be interpreted as difference
from a reference category** like it was the case with a dummy coded factor.

