---
title: 'Model Summary Function'
output: 
  html_document:
    toc: true
    toc_float:
      collapsed: false
    number_sections: false
    theme: spacelab
    highlight: tango
    includes:
      after_body: ../footerNE.html
    css: ../style.css
---


```{r setup, include = FALSE}
library(knitr)
library(kableExtra)

opts_chunk$set(echo = TRUE)

hint <- function(text) {
  id <- paste0('hint_', sample.int(1e10, size = 1))
    cat(
    paste0('\n',
      '<button type="button" class="btn btn-info btn-sm" data-toggle="collapse" ',
      'data-target="#', id, '">Hint</button>',
      '<div id = "', id, '" class="collapse" style="border:1px; ',
      'border-style:solid; padding: 1em; border-color:#1F78B4">',
      text, '</div>')
  )
}

set.seed(2020)
```


The aim of this practical is to write a function that returns a nice looking
summary of the results from a linear or logistic regression model.

## Creating Data {.tabset .tabset-pills .tabset-fade}
### Task
<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">
Create a `data.frame` with example data that we can use to fit example models
to test our function.
```{r, echo = FALSE, results = 'asis'}
hint('To create data, you could draw random numbers from a normal distribution, 
using `rnorm()`, from a uniform distribution, using `runif()` or from a 
binomial distribution, using `rbinom()`. The functions `sample()` and `factor()`
could also be useful.')
```
</div>


### Solution
```{r}
N <- 1000
testdat <- data.frame(y = rnorm(N),
                      ybin = factor(rbinom(N, size = 1, prob = 0.7)),
                      x1 = runif(N, min = 0, max = 4),
                      x2 = factor(sample(paste0('cat', 1:4), size = N, replace = TRUE))
)

testdat$x3 <- 0.5 * testdat$y + 0.5 * as.numeric(testdat$ybin) + rnorm(N, 0, 1)

summary(testdat)
```
<div style="border:2px; border-style:solid; padding: 1em; border-color:#009B8F">
<p style = "color:#009B8F; font-weight:bold;">Note:</p>
We create one variable that is (strongly) associated with the outcome variables
to make sure we have some significant p-values.
</div>


## Example Models {.tabset .tabset-pills .tabset-fade}
### Task
<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">
Using your test data, fit a linear and a logistic regression model.
</div>


### Solution
```{r}
test_lm <- lm(y ~ ., data = testdat)
test_logit <- glm(ybin ~ ., data = testdat, family = binomial())
```

## Building the Summary {.tabset .tabset-pills .tabset-fade}
### Task 1
Our summary should contain the estimates of the regression coefficients, 
the 95% confidence interval, and the p-value.

<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">
Explore the `summary()` of the linear regression model (not the printed output
but the object that is returned) to find out how you can access the p-values
from a `lm`.
</div>

### Solution 1
```{r}
summary_lm <- summary(test_lm)
class(summary_lm)
names(summary_lm)
is.list(summary_lm)

summary_lm$coefficients
class(summary_lm$coefficients)
```


### Task 2
<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">
Combine the regression coefficients, the 95% confidence interval and the 
p-values from the linear regression model into one `data.frame`.

Make sure the columns are named properly (for example: `estimate`, `2.5 %`, `97.5 %`, `p-value`).

```{r, echo = FALSE, results = 'asis'}
hint('Try the argument `check.names = FALSE`.')
```
</div>

### Solution 2
```{r}
dat <- data.frame(estimate = coef(test_lm),
                  confint(test_lm),
                  'p-value' = summary(test_lm)$coefficients[, 'Pr(>|t|)'],
                  check.names = FALSE)
dat
```

<div style="border:2px; border-style:solid; padding: 1em; border-color:#009B8F">
<p style = "color:#009B8F; font-weight:bold;">Note:</p>
We need to wrap `p-value` in quotes because it does contain a "-" and is not
a syntactically valid name. 

`estimate` is a syntactically valid name.
</div>

### Task 3
<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">
* Write a function that has a linear regression model as input and returns a 
  summary like the one we have constructed in the previous solution.
* Test the function with the example linear regression.
</div>


### Solution 3
```{r}
fun1 <- function(model) {
  dat <- data.frame(estimate = coef(test_lm),
                  confint(test_lm),
                  'p-value' = summary(test_lm)$coefficients[, 'Pr(>|t|)'],
                  check.names = FALSE)
  
  dat
}

fun1(test_lm)
```


## Improving the Look {.tabset .tabset-pills .tabset-fade}
### Task 1
<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">
To make the output look nicer, we want to limit the decimal places in the
output to a number specified by the user.
Implement this using an argument `decimals` and decide on a default value for it.
</div>


### Solution 1
```{r}
fun2 <- function(model, decimals = 3) {
  dat <- data.frame(estimate = coef(model),
                  confint(model),
                  'p-value' = summary(model)$coefficients[, 'Pr(>|t|)'],
                  check.names = FALSE)
  
  round(dat, decimals)
}

fun2(test_lm)
```


### Task 2
<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">
Previously, we have seen that the function `format.pval()` will convert values
below a certain value in to "<..." notation. 

Use this in our summary function.
```{r, results = 'asis', echo = FALSE}
hint('`format.pval()` has the additional arguments `digits` and `eps`.')
hint('To use the `$` to select a column with a syntactically invalid name you
need to wrap the name in "`", i.e., &#x60;the name&#x60;.')
hint('Values can also be represented as $10^(-3)$.')
```
</div>

### Solution 2
```{r}
fun3 <- function(model, decimals = 3) {
  dat <- data.frame(estimate = coef(model),
                    confint(model),
                    check.names = FALSE)
  
  dat <- round(dat, decimals)
  
  dat$`p-value` <- format.pval(summary(model)$coefficients[, 'Pr(>|t|)'],
                               eps = 10^(-decimals), digits = decimals)
  
  dat
}

fun3(test_lm)
fun3(test_lm, decimals = 5)
fun3(test_lm, decimals = 1)
```


<div style="border:2px; border-style:solid; padding: 1em; border-color:#009B8F">
<p style = "color:#009B8F; font-weight:bold;">Note:</p>
We do not want to first round the p-value and then apply `format.pval()` because
this might give wrong results in specific cases, for example:
```{r}
pv <- 0.0009
format.pval(pv, digits = 2, eps = 0.001)

round(pv, 3)
format.pval(round(pv, 3), digits = 2, eps = 0.001)
```
</div>

## Logistic Regression {.tabset .tabset-pills .tabset-fade}
### Task 1
<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">
We also would like to use this function for output from logistic regression models.

Does the current version of the function work with logistic regression models?

* If it does: does it give us the summary that we are interested in?
* If it does not: investigate what the problem is.
</div>

### Solution 1
```{r, error = TRUE}
fun3(test_logit)
```

We get an error message "subscript out of bounds". This means we are 
trying to select an element that does not exist.

To find the problem, we check each part of the function body, replacing
`model` by `test_logit`:
```{r, error = TRUE}
decimals <- 3

dat <- data.frame(estimate = coef(test_logit),
                  confint(test_logit),
                  check.names = FALSE)

dat <- round(dat, decimals)

dat$`p-value` <- format.pval(summary(test_logit)$coefficients[, 'Pr(>|t|)'],
                             eps = 10^(-decimals), digits = decimals)
```

The problem is in the p-value column, so we split this syntax up further:
```{r, error = TRUE}
summary(test_logit)$coefficients[, 'Pr(>|t|)']
```
Same error, so probably the column `'Pr(>|t|)'` does not exist.

```{r}
summary(test_logit)$coefficients
```
Indeed, for a logistic regression model, the column is called `Pr(>|z|)`!


### Task 2
<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">
Adjust the function so that it works with both linear and logistic regression models.
</div>

### Solution 2
```{r}
fun4 <- function(model, decimals = 3, modeltype) {
  dat <- data.frame(estimate = coef(model),
                    confint(model),
                    check.names = FALSE)
  
  dat <- round(dat, decimals)
  
  if (modeltype == 'lm') {
    dat$`p-value` <- format.pval(summary(model)$coefficients[, 'Pr(>|t|)'],
                                 eps = 10^(-decimals), digits = decimals)
  } else {
    dat$`p-value` <- format.pval(summary(model)$coefficients[, 'Pr(>|z|)'],
                                 eps = 10^(-decimals), digits = decimals)
  }
  
  dat
}

fun4(test_logit, modeltype = 'logit')
fun4(test_lm, modeltype = 'lm')
```

### Task 3
<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">
Now that the function "works", i.e., runs for both types of models, we can come
back to the other question:

For a logistic model: is this the output that we are interested in?
</div>

### Solution 3
For a logistic regression model, the effects are usually reported as odds ratios.
In our current version, we present the effects on the log odds scale.


### Task 4
<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">
Adjust the function so that for logistic regression models the summary is 
given as odds ratios. 

Do not forget to change the column name(s).
</div>

### Solution 4
```{r}
fun5 <- function(model, decimals = 3, modeltype) {
  dat <- data.frame(estimate = coef(model),
                    confint(model),
                    check.names = FALSE)
  
  if (modeltype == 'lm') {
    dat <- round(dat, decimals)
    
    dat$`p-value` <- format.pval(summary(model)$coefficients[, 'Pr(>|t|)'],
                                 eps = 10^(-decimals), digits = decimals)
  } else {
    dat <- round(exp(dat), decimals)
    
    dat$`p-value` <- format.pval(summary(model)$coefficients[, 'Pr(>|z|)'],
                                 eps = 10^(-decimals), digits = decimals)
    
    names(dat)[1] <- 'OR'
  }
  
  dat
}

fun5(test_logit, modeltype = 'logit')
fun5(test_lm, modeltype = 'lm')
```


### Task 5
<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">
To make the function more convenient to use, we want to automatically identify
the model type, so that the user does not need to provide the `modeltype`.
```{r}
hint('The function `family()` can be used on regression models.')
```

</div>

### Solution 5
A first idea might be to use `class(model)` to identify the type of model.
```{r, collapse=TRUE}
class(test_lm)
class(test_logit)
```


The problem with this would be that a user might also have used `glm()` to fit 
a linear regression model:
```{r}
class(glm(formula = y ~ ., data = testdat, family = gaussian()))
```


Moreover, not all GLMs use a logit link (not even all binary GLMs do!) and
using `exp()` would then not give an odds ratio:
```{r}
class(glm(formula = ybin ~ ., data = testdat, family = binomial(link = 'probit')))
```


Using `family()` we can extract the model family and the link function:
```{r}
family(test_lm)
family(test_logit)
```

We need to inspect this further to figure out how to access that information:
```{r}
fam_lm <- family(test_lm)
str(fam_lm)

fam_lm$family
fam_lm$link


family(test_logit)$family
family(test_logit)$link
```

Using this information we can adapt our function:

```{r}
fun6 <- function(model, decimals = 3) {

  # extract the family and link  
  fam <- family(model)$family
  link <- family(model)$link
  
  modeltype <- if (fam == 'binomial' & link == 'logit') {
    'logit'
  } else {
    if (fam == 'gaussian' & link == 'identity') {
      'lm'
    }
  }
  
  dat <- data.frame(estimate = coef(model),
                    confint(model),
                    check.names = FALSE)
  
  if (modeltype == 'lm') {
    dat <- round(dat, decimals)
    
    dat$`p-value` <- format.pval(summary(model)$coefficients[, 'Pr(>|t|)'],
                                 eps = 10^(-decimals), digits = decimals)
  } else {
    dat <- round(exp(dat), decimals)
    
    dat$`p-value` <- format.pval(summary(model)$coefficients[, 'Pr(>|z|)'],
                                 eps = 10^(-decimals), digits = decimals)
    
    names(dat)[1] <- 'OR'
  }
  
  dat
}
```

### Task 6
<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">
What would happen if we would apply our function to a model that is not
a standard linear regression (with identity link) or logistic regression model?

Try to figure this out by looking at the syntax without trying it out!
</div>

### Solution 6
If we provide a `model` that does not fit either of the criteria
`fam == 'binomial' & link == 'logit'` or `fam == 'gaussian' & link == 'identity'`
the value of `modeltype` will be `NULL`.<br>

Why?<br>
The first condition (`fam == 'binomial' & link == 'logit'`) would be `FALSE`
so we would end up in the `else` part.
There, the condition (`fam == 'gaussian' & link == 'identity'`) would also
be `FALSE`. Since this `if()` does not provide any alternative, it will return
`NULL`.


The question is what happens when we try to evaluate `if (modeltype == 'lm')`
when `modeltype = NULL`.

The comparison of `NULL` with `'lm'` will result in a `logical` vector of length
zero:
```{r}
NULL == 'lm'
```

Using a `logical` vector of length zero as condition to `if()` results in an error:
```{r, error = TRUE}
if (NULL == 'lm') {}
```


```{r, error = TRUE}
test_probit <- glm(ybin ~ ., family = binomial(link = 'probit'), data = testdat)
fun6(test_probit)
```


### Task 7
<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">
Prevent this error from occuring by stopping the evaluation of the function
before it happens and provide an informative error message.
```{r, echo = FALSE, results = 'asis'}
hint("To create an error message, use `stop()`.")
hint("You could use the function `is.null()`.")
```
</div>

### Solution 7
```{r}
fun7 <- function(model, decimals = 3) {

  # extract the family and link  
  fam <- family(model)$family
  link <- family(model)$link
  
  modeltype <- if (fam == 'binomial' & link == 'logit') {
    'logit'
  } else {
    if (fam == 'gaussian' & link == 'identity') {
      'lm'
    }
  }
  
  if (is.null(modeltype)) {
    stop("I do not know how to create the summary for the type of model you have provided.")
  }
  
  dat <- data.frame(estimate = coef(model),
                    confint(model),
                    check.names = FALSE)
  
  if (modeltype == 'lm') {
    dat <- round(dat, decimals)
    
    dat$`p-value` <- format.pval(summary(model)$coefficients[, 'Pr(>|t|)'],
                                 eps = 10^(-decimals), digits = decimals)
  } else {
    dat <- round(exp(dat), decimals)
    
    dat$`p-value` <- format.pval(summary(model)$coefficients[, 'Pr(>|z|)'],
                                 eps = 10^(-decimals), digits = decimals)
    
    names(dat)[1] <- 'OR'
  }
  
  dat
}
```

```{r, error = TRUE}
fun7(test_probit)
```

