---
title: 'Basic Use of R'
output: 
  html_document:
    code_folding: show
    df_print: paged
    toc: true
    toc_float:
      collapsed: false
    number_sections: false
    theme: spacelab
    highlight: tango
    includes:
      after_body: ../footerEA.html
    css: ../style.css
---

```{r, echo = FALSE, purl = FALSE}
knitr::knit_hooks$set(purl = knitr::hook_purl)

options(purl = FALSE)

knitr::opts_chunk$set(purl = FALSE)
```

```{r, include = FALSE}
knitr::opts_hooks$set(eval = function(opt) {
  if (any(opt$exercise))
    opt$eval <- opt$include <- FALSE
  
  opt
})


static <- TRUE

options(width = 100)

```


```{r packages, include = FALSE}
library(kableExtra)
library(knitr)
```


```{r load_data, context="data", include=FALSE}
library(survival)
```


## Preface {data-progressive=FALSE}
### R packages
In this practical, a number of R packages are used.
The packages used (with versions that were used to generate the solutions) are:

* `r R.version.string`
* `survival` (version: `r packageVersion("survival")`)
* `memisc` (version: `r packageVersion("memisc")`)
* `ggplot2` (version: `r packageVersion("ggplot2")`)


### Dataset {.tabset .tabset-fade .tabset-pills}

For this practical, we will use the **heart** and **retinopathy** data sets from the `survival` package. More details about the data sets can be found in:

https://stat.ethz.ch/R-manual/R-devel/library/survival/html/heart.html

https://stat.ethz.ch/R-manual/R-devel/library/survival/html/retinopathy.html

Always check your data.


```{r, eval = static, echo = FALSE}
asis_output("#### Task\\n")
```

```{r, eval = static, echo = FALSE}
asis_output('<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">')
```

Explore the **heart** and **retinopathy** data sets - print the first six and last six rows. 

```{r data_hint, eval = static, results = 'asis', echo = FALSE}
cat('<button type="button" class="btn btn-info btn-sm" data-toggle="collapse" data-target="#data">Hint</button>
<div id = "data" class="collapse" style="border:1px; border-style:solid; padding: 1em; border-color:#1F78B4">
Use the functions head(...) and tail(...) to investigate the data set. Replace the dots with the name of the data set.
</div>')
```


```{r, eval = static, echo = FALSE}
asis_output("</div>")

asis_output("#### Solution\\n")
```

```{r data-solution, solution = TRUE}
library(survival)
head(heart)
tail(heart)
head(retinopathy)
tail(retinopathy)
```

## Common R Objects

It is important to distinguish the different object in R.

### Vectors {.tabset .tabset-fade .tabset-pills}

Let's create some vectors.

```{r, eval = static, echo = FALSE}
asis_output("#### Task 1\\n")
```

```{r, eval = static, echo = FALSE}
asis_output('<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">')
```

View the vectors `event` and `age` from the **heart** data set.

```{r vec_hint, eval = static, results = 'asis', echo = FALSE}
cat('<button type="button" class="btn btn-info btn-sm" data-toggle="collapse" data-target="#vec">Hint</button>
<div id = "vec" class="collapse" style="border:1px; border-style:solid; padding: 1em; border-color:#1F78B4">
Use the dollar sign to select the variables.
</div>')
```


```{r, eval = static, echo = FALSE}
asis_output("</div>")

asis_output("#### Solution 1\\n")
```

```{r vec-solution, solution = TRUE}
heart$event
heart$age
```



```{r, eval = static, echo = FALSE}
asis_output("#### Task 2\\n")
```

```{r, eval = static, echo = FALSE}
asis_output('<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">')
```

View the vectors `eye` and `risk` from the **retinopathy** data set.

```{r vec2_hint, eval = static, results = 'asis', echo = FALSE}
cat('<button type="button" class="btn btn-info btn-sm" data-toggle="collapse" data-target="#vec2">Hint</button>
<div id = "vec2" class="collapse" style="border:1px; border-style:solid; padding: 1em; border-color:#1F78B4">
Use the dollar sign to select the variables.
</div>')
```


```{r, eval = static, echo = FALSE}
asis_output("</div>")

asis_output("#### Solution 2\\n")
```

```{r vec2-solution, solution = TRUE}
retinopathy$eye
retinopathy$risk
```





```{r, eval = static, echo = FALSE}
asis_output("#### Task 3\\n")
```

```{r, eval = static, echo = FALSE}
asis_output('<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">')
```

Create a numerical vector that consists of the values: 34, 24, 19, 23, 16. Give the name `numbers` to this new vector.

```{r vec3_hint, eval = static, results = 'asis', echo = FALSE}
cat('<button type="button" class="btn btn-info btn-sm" data-toggle="collapse" data-target="#vec3">Hint</button>
<div id = "vec3" class="collapse" style="border:1px; border-style:solid; padding: 1em; border-color:#1F78B4">
Use the c(...) function. Replace the dots with the numbers.
</div>')
```


```{r, eval = static, echo = FALSE}
asis_output("</div>")

asis_output("#### Solution 3\\n")
```

```{r vec3-solution, solution = TRUE}
numbers <- c(34, 24, 19, 23, 16)
```






```{r, eval = static, echo = FALSE}
asis_output("#### Task 4\\n")
```

```{r, eval = static, echo = FALSE}
asis_output('<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">')
```

Create a numerical vector that takes the integer values from 1 until 200. Give the name `numbers_2` to this new vector.

```{r vec4_hint, eval = static, results = 'asis', echo = FALSE}
cat('<button type="button" class="btn btn-info btn-sm" data-toggle="collapse" data-target="#vec4">Hint</button>
<div id = "vec4" class="collapse" style="border:1px; border-style:solid; padding: 1em; border-color:#1F78B4">
Use the c(...) function. Replace the dots with the numbers.
</div>')
```


```{r, eval = static, echo = FALSE}
asis_output("</div>")

asis_output("#### Solution 4\\n")
```

```{r vec4-solution, solution = TRUE}
numbers_2 <- c(1:200)
```






```{r, eval = static, echo = FALSE}
asis_output("#### Task 5\\n")
```

```{r, eval = static, echo = FALSE}
asis_output('<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">')
```

Create a categorical vector that consists of the values: yes, yes, no, no, no, yes. Give the name `treatment` to this new vector.

```{r vec5_hint, eval = static, results = 'asis', echo = FALSE}
cat('<button type="button" class="btn btn-info btn-sm" data-toggle="collapse" data-target="#vec5">Hint</button>
<div id = "vec5" class="collapse" style="border:1px; border-style:solid; padding: 1em; border-color:#1F78B4">
Use the c(...) function. Replace the dots with the numbers.
</div>')
```


```{r, eval = static, echo = FALSE}
asis_output("</div>")

asis_output("#### Solution 5\\n")
```

```{r vec5-solution, solution = TRUE}
treatment <- c("yes", "yes", "no", "no", "no", "yes")
```



### Matrices and Data Frames {.tabset .tabset-fade .tabset-pills}

Let's create some matrices and data frames.

```{r, eval = static, echo = FALSE}
asis_output("#### Task 1\\n")
```

```{r, eval = static, echo = FALSE}
asis_output('<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">')
```

Create a matrix using the vectors `id` and `age` from the **heart** data set. This matrix should have 2 columns where each column represents each variable.

```{r mat_hint, eval = static, results = 'asis', echo = FALSE}
cat('<button type="button" class="btn btn-info btn-sm" data-toggle="collapse" data-target="#mat">Hint</button>
<div id = "mat" class="collapse" style="border:1px; border-style:solid; padding: 1em; border-color:#1F78B4">
Use the function matrix(...).
</div>')
```


```{r, eval = static, echo = FALSE}
asis_output("</div>")

asis_output("#### Solution 1\\n")
```

```{r mat-solution-1, solution = TRUE}
matrix(c(heart$id, heart$age), , 2) 
```



```{r, eval = static, echo = FALSE}
asis_output("#### Task 2\\n")
```

```{r, eval = static, echo = FALSE}
asis_output('<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">')
```

Create a data frame using the vectors `id`, `type` and `trt` from the **retinopathy** data set. This data frame should have 3 columns where each column represents each variable.

```{r df_hint, eval = static, results = 'asis', echo = FALSE}
cat('<button type="button" class="btn btn-info btn-sm" data-toggle="collapse" data-target="#df">Hint</button>
<div id = "df" class="collapse" style="border:1px; border-style:solid; padding: 1em; border-color:#1F78B4">
Use the function data.frame(...).
</div>')
```


```{r, eval = static, echo = FALSE}
asis_output("</div>")

asis_output("#### Solution 2\\n")
```

```{r dfc-solution, solution = TRUE}
data.frame(id = retinopathy$id, type = retinopathy$type, trt = retinopathy$trt) 
```



### Lists {.tabset .tabset-fade .tabset-pills}

Let's create some lists.

```{r, eval = static, echo = FALSE}
asis_output("#### Task 1\\n")
```

```{r, eval = static, echo = FALSE}
asis_output('<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">')
```

Create a list using the vectors `stop` from the **heart** data set and `id`, `risk` from the **retinopathy** data set. Give the names `stop_heart`, `id_reti` and `risk_reti`.

```{r list_hint, eval = static, results = 'asis', echo = FALSE}
cat('<button type="button" class="btn btn-info btn-sm" data-toggle="collapse" data-target="#list">Hint</button>
<div id = "list" class="collapse" style="border:1px; border-style:solid; padding: 1em; border-color:#1F78B4">
Use the function list(...).
</div>')
```


```{r, eval = static, echo = FALSE}
asis_output("</div>")

asis_output("#### Solution 1\\n")
```

```{r list-solution, solution = TRUE}
list(heart$stop, retinopathy$id, retinopathy$risk)
```


```{r, eval = static, echo = FALSE}
asis_output("#### Task 2\\n")
```

```{r, eval = static, echo = FALSE}
asis_output('<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">')
```

Create a list using the vectors `numbers`, `numbers_2` and `treatment`. Give the names `numbers`, `many_numbers` and `treatment`.

```{r list2_hint, eval = static, results = 'asis', echo = FALSE}
cat('<button type="button" class="btn btn-info btn-sm" data-toggle="collapse" data-target="#list2">Hint</button>
<div id = "list2" class="collapse" style="border:1px; border-style:solid; padding: 1em; border-color:#1F78B4">
Use the function list(...).
</div>')
```


```{r, eval = static, echo = FALSE}
asis_output("</div>")

asis_output("#### Solution 2\\n")
```

```{r list2-solution, solution = TRUE}
list(numbers = numbers, many_numbers = numbers_2, treatment = treatment)
```



## Importing and Saving data
### Save your work {.tabset .tabset-fade .tabset-pills}

```{r, eval = static, echo = FALSE}
asis_output("#### Task\\n")
```

```{r, eval = static, echo = FALSE}
asis_output('<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">')
```

Save the vectors `numbers`, `numbers_2` and `treatment`. Use the name `new_vectors`.

```{r save_hint, eval = static, results = 'asis', echo = FALSE}
cat('<button type="button" class="btn btn-info btn-sm" data-toggle="collapse" data-target="#save">Hint</button>
<div id = "save" class="collapse" style="border:1px; border-style:solid; padding: 1em; border-color:#1F78B4">
Use the function save().
</div>')
```


```{r, eval = static, echo = FALSE}
asis_output("</div>")

asis_output("#### Solution\\n")
```

```{r save-solution, solution = TRUE}
save(numbers, numbers_2, treatment, file = "new_vectors.RData")
```


### Load your work {.tabset .tabset-fade .tabset-pills}

```{r, eval = static, echo = FALSE}
asis_output("#### Task\\n")
```

```{r, eval = static, echo = FALSE}
asis_output('<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">')
```

Load the file `new_vectors`.

```{r load_hint, eval = static, results = 'asis', echo = FALSE}
cat('<button type="button" class="btn btn-info btn-sm" data-toggle="collapse" data-target="#load">Hint</button>
<div id = "load" class="collapse" style="border:1px; border-style:solid; padding: 1em; border-color:#1F78B4">
Use the function load().
</div>')
```


```{r, eval = static, echo = FALSE}
asis_output("</div>")

asis_output("#### Solution\\n")
```

```{r load-solution, solution = TRUE}
load("new_vectors.RData")
```


### Remove your work {.tabset .tabset-fade .tabset-pills}

```{r, eval = static, echo = FALSE}
asis_output("#### Task\\n")
```

```{r, eval = static, echo = FALSE}
asis_output('<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">')
```

Remove the vectors `numbers`, `numbers_2` and `treatment`.

```{r rm_hint, eval = static, results = 'asis', echo = FALSE}
cat('<button type="button" class="btn btn-info btn-sm" data-toggle="collapse" data-target="#rm">Hint</button>
<div id = "rm" class="collapse" style="border:1px; border-style:solid; padding: 1em; border-color:#1F78B4">
Use the function rm().
</div>')
```


```{r, eval = static, echo = FALSE}
asis_output("</div>")

asis_output("#### Solution\\n")
```

```{r rm-solution, solution = TRUE}
rm(numbers, numbers_2, treatment)
```


## Data Manipulation, Exploration and Visualization

Before starting with any statistical analysis it is important to transform and explore your data set.

### Data Manipulation {.tabset .tabset-fade .tabset-pills}


```{r, eval = static, echo = FALSE}
asis_output("#### Task 1\\n")
```

```{r, eval = static, echo = FALSE}
asis_output('<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">')
```

* As mentioned in the manual of the **heart** data set, `age` is equal to `age` - 48. Let's bring `age` back to the normal scale. Do not overwrite the variable `age`, but create a new variable with the name `age_orig`. \
* Convert the variable `surgery` into a factor with levels 0: `no` and 1: `yes`.

```{r, eval = static, echo = FALSE}
asis_output("</div>")

asis_output("#### Solution 1\\n")
```

```{r man1-solution, solution = TRUE}
heart$age_orig <- heart$age + 48
heart$surgery <- factor(heart$surgery, levels = c(0, 1), labels = c("no", "yes"))
```




```{r, eval = static, echo = FALSE}
asis_output("#### Task 2\\n")
```

```{r, eval = static, echo = FALSE}
asis_output('<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">')
```

Categorize as as `young`: [minimum `age` until mean `age`) and `old`: [mean `age` until maximum `age`). Give this variable the name `ageCat`. Print the first 6 rows of the data set **retinopathy**.

```{r man2_hint, eval = static, results = 'asis', echo = FALSE}
cat('<button type="button" class="btn btn-info btn-sm" data-toggle="collapse" data-target="#man2">Hint</button>
<div id = "man2" class="collapse" style="border:1px; border-style:solid; padding: 1em; border-color:#1F78B4">
Use the function as.numeric(...) to convert a variable to a numeric variable. Use the function factor(...) to create a categorical variable. 
</div>')
```

```{r, eval = static, echo = FALSE}
asis_output("</div>")

asis_output("#### Solution 2\\n")
```

```{r man2-solution, solution = TRUE}
retinopathy$ageCat <- as.numeric(retinopathy$age >= mean(retinopathy$age))
retinopathy$ageCat <- factor(retinopathy$ageCat, levels = c(0, 1), labels = c("young", "old"))
head(retinopathy)
```




```{r, eval = static, echo = FALSE}
asis_output("#### Task 3\\n")
```

```{r, eval = static, echo = FALSE}
asis_output('<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">')
```

Categorize `futime` from data set **retinopathy** as follows: \
* `short`: [minimum `futime` until 2). \
* `medium`: [25 until 45). \
* `long`: [45 until maximum `futime`). Give this variable the name `futimeCat`. \
* Print the first 6 rows of the data.


```{r man3_hint, eval = static, results = 'asis', echo = FALSE}
cat('<button type="button" class="btn btn-info btn-sm" data-toggle="collapse" data-target="#man3">Hint</button>
<div id = "man3" class="collapse" style="border:1px; border-style:solid; padding: 1em; border-color:#1F78B4">
 Use the function factor(...) to create a categorical variable. Use the function cut(..., right = FALSE) to convert a continuous variable to a categorical variable. If needed consult the documentation for the cut function using ?cut.
</div>')
```

```{r, eval = static, echo = FALSE}
asis_output("</div>")

asis_output("#### Solution 3\\n")
```

```{r man3-solution, solution = TRUE}
retinopathy$futimeCat <- cut(retinopathy$futime, c(min(retinopathy$futime), 25, 45, max(retinopathy$futime)), right = FALSE)
retinopathy$futimeCat <- factor(retinopathy$futimeCat, levels = c("[0.3,25)", "[25,45)", "[45,75)"), labels = c("short", "medium", "long"))
head(retinopathy)
```








```{r, eval = static, echo = FALSE}
asis_output("#### Task 4\\n")
```

```{r, eval = static, echo = FALSE}
asis_output('<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">')
```

Create 2 vectors of size 50 as follows: \
* `Sex`: takes 2 values 0 and 1. \
* `Age`: takes values from 20 till 80 (hint: use sample()). \
* Convert “Sex” to a factor with levels 0: “female” and 1: “male”. \
* Define the new variable “AgeCat” as dichotomous with `Age` <= 50 to be 0 and 1 otherwise. \
* Convert `AgeCat` to a factor with levels 0: `young` and 1: `old`. \
* Overwrite `Age` variable by $\frac{Age-mean(Age)}{sd(Age)}$.

```{r man4_hint, eval = static, results = 'asis', echo = FALSE}
cat('<button type="button" class="btn btn-info btn-sm" data-toggle="collapse" data-target="#man4">Hint</button>
<div id = "man4" class="collapse" style="border:1px; border-style:solid; padding: 1em; border-color:#1F78B4">
To sample a numarical and categorical variable use the function sample(...). To convert a numerical variable to a categorical use the function factor(...). To dichotomize a numerical variable use the function as.numeric(...).
</div>')
```


```{r, eval = static, echo = FALSE}
asis_output("</div>")

asis_output("#### Solution 4\\n")
```

```{r man4-solution, solution = TRUE}
Sex <- sample(0:1, 50, replace = T)
Age <- sample(20:80, 50, replace = T)
Sex <- factor(Sex, levels = c(0:1), labels = c("female", "male"))
AgeCat <- as.numeric(Age > 50)
AgeCat <- factor(AgeCat, levels = c(0:1), labels = c("young", "old"))
Age <- (Age - mean(Age))/sd(Age)
```




```{r, eval = static, echo = FALSE}
asis_output("#### Task 5\\n")
```

```{r, eval = static, echo = FALSE}
asis_output('<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">')
```

Create a data.frame with the name `DF` as follows: \
* Include the following vectors: `Sex`, `Age`, `AgeCat`. \
* Use the names: `Gender`, `StandardizedAge`, `DichotomousAge`. \
* What are the dimensions of the data.frame?

```{r man5_hint, eval = static, results = 'asis', echo = FALSE}
cat('<button type="button" class="btn btn-info btn-sm" data-toggle="collapse" data-target="#man5">Hint</button>
<div id = "man5" class="collapse" style="border:1px; border-style:solid; padding: 1em; border-color:#1F78B4">
To obtain the dimensions use the function dim(...).
</div>')
```


```{r, eval = static, echo = FALSE}
asis_output("</div>")

asis_output("#### Solution 5\\n")
```

```{r man5-solution, solution = TRUE}
DF <- data.frame(Sex, Age, AgeCat)
DF <- data.frame("Gender" = Sex, "StandardizedAge" = Age, "DichotomousAge" = AgeCat)
dim(DF)
```


```{r, eval = static, echo = FALSE}
asis_output("#### Task 6\\n")
```

```{r, eval = static, echo = FALSE}
asis_output('<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">')
```

Create 2 vectors of size 150 as follows: \
* `Treatment`: takes 2 values 1 and 2. \
* `Weight`: takes values from 50 till 100. \
* Convert `Treatment` to a factor with levels 1: `no` and 2: `yes`. \
* Overwrite `Weight` variable by `Weight` * 1000. \
* Create a data.frame including `Treatment and `Weight`. \



```{r, eval = static, echo = FALSE}
asis_output("</div>")

asis_output("#### Solution 6\\n")
```

```{r man6-solution, solution = TRUE}
Treatment <- sample(1:2, 150, replace = T)
Weight <- sample(50:100, 150, replace = T)
Treatment <- factor(Treatment, levels = c(1:2), labels = c("no", "yes"))
Weight <- Weight * 1000
data.frame(Treatment, Weight)
```



```{r, eval = static, echo = FALSE}
asis_output("#### Task 7\\n")
```

```{r, eval = static, echo = FALSE}
asis_output('<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">')
```

Create a list called `my_list` with the following: \
* `let`: `a` to `i`. \
* `sex`: factor taking the values `males` and `females` and length 50. \
* `mat`: matrix 
<TABLE>
   <TR>
      <TD>1 \ \  </TD>
      <TD>2</TD>
   </TR>
   <TR>
      <TD>3 \ \  </TD>
      <TD>4</TD>
   </TR>
</TABLE> 

```{r man7_hint, eval = static, results = 'asis', echo = FALSE}
cat('<button type="button" class="btn btn-info btn-sm" data-toggle="collapse" data-target="#man7">Hint</button>
<div id = "man7" class="collapse" style="border:1px; border-style:solid; padding: 1em; border-color:#1F78B4">
To obtain letters use the function letters(...).
</div>')
```


```{r, eval = static, echo = FALSE}
asis_output("</div>")

asis_output("#### Solution 7\\n")
```

```{r man7-solution, solution = TRUE}
let <- letters[1:9]
sex <- sample(1:2, 50, replace = TRUE)
sex <- factor(sex, levels = 1:2, labels = c("males", "females"))
mat <- matrix(1:4 ,2, 2, byrow = TRUE)
my_list <- list(let = let, sex = sex, mat = mat) 
```


### Data Exploration {.tabset .tabset-fade .tabset-pills}

Let's obtain some descriptive statistics.

```{r, eval = static, echo = FALSE}
asis_output("#### Task 1\\n")
```

```{r, eval = static, echo = FALSE}
asis_output('<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">')
```

Obtain the mean and sd for `age` only for transplanted patients using **heart** the data set.

```{r exp1_hint, eval = static, results = 'asis', echo = FALSE}
cat('<button type="button" class="btn btn-info btn-sm" data-toggle="collapse" data-target="#exp1">Hint</button>
<div id = "exp1" class="collapse" style="border:1px; border-style:solid; padding: 1em; border-color:#1F78B4">
Use square brackets to select transplanted patients.
</div>')
```

```{r, eval = static, echo = FALSE}
asis_output("</div>")

asis_output("#### Solution 1\\n")
```

```{r exp1-solution, solution = TRUE}
mean(heart$age[heart$transplant == 1])
sd(heart$age[heart$transplant == 1])
```



```{r, eval = static, echo = FALSE}
asis_output("#### Task 2\\n")
```

```{r, eval = static, echo = FALSE}
asis_output('<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">')
```

Using the **retinopathy** data set:
* Obtain the median and interquartile range for `age`. \ 
* Obtain the percentage per `type`. \
* Check whether there are missing values in the variable `age`.

```{r exp2_hint, eval = static, results = 'asis', echo = FALSE}
cat('<button type="button" class="btn btn-info btn-sm" data-toggle="collapse" data-target="#exp2">Hint</button>
<div id = "exp2" class="collapse" style="border:1px; border-style:solid; padding: 1em; border-color:#1F78B4">
Use the functions median(...) and IQR(...). Load the package memisc in order to obain the percentages. To check whether there are missing values use the combination of the functions sum(is.na(...)).
</div>')
```

```{r, eval = static, echo = FALSE}
asis_output("</div>")

asis_output("#### Solution 2\\n")
```

```{r exp2-solution, solution = TRUE}
median(retinopathy$age)
IQR(retinopathy$age)
library(memisc)
percent(retinopathy$type)
sum(is.na(retinopathy$age)) # any(is.na(retinopathy$age))
```


```{r, eval = static, echo = FALSE}
asis_output("#### Task 3\\n")
```

```{r, eval = static, echo = FALSE}
asis_output('<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">')
```

Using the data frame `DF` from the exercise before: \
* Calculate the mean of the variable `StandardizedAge`. \
* Calculate the standard deviation of the variable `StandardizedAge`. \
* Calculate the frequencies of the variable `Gender`. \
* Calculate the frequencies of the variable `DichotomousAge`. \
* Calculate the frequencies of both variables `Gender` and `DichotomousAge` (crosstab table).

```{r exp3_hint, eval = static, results = 'asis', echo = FALSE}
cat('<button type="button" class="btn btn-info btn-sm" data-toggle="collapse" data-target="#exp3">Hint</button>
<div id = "exp3" class="collapse" style="border:1px; border-style:solid; padding: 1em; border-color:#1F78B4">
To calculate the frequencies use function length() or table().
</div>')
```

```{r, eval = static, echo = FALSE}
asis_output("</div>")

asis_output("#### Solution 3\\n")
```

```{r exp3-solution, solution = TRUE}
mean(DF$StandardizedAge)
sd(DF$StandardizedAge)
length(DF$Gender[DF$Gender == "female"])
length(DF$Gender[DF$Gender == "male"])
table(DF$Gender)
table(DF$Gender, DF$DichotomousAge)
```



### Data Visualization {.tabset .tabset-fade .tabset-pills}

Let's us now plot the data in order to understand them even better.

```{r, eval = static, echo = FALSE}
asis_output("#### Task 1\\n")
```

```{r, eval = static, echo = FALSE}
asis_output('<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">')
```

* Create a scatterplot with the variables `age` and `year` using the **heart** data set. \
* Change the labels of the axis. In particular, give the name `Age` for the x-axis and `Year of acceptance` for the y-axis. \
* Give a different color to the patients that had a transplant. \
* Add a legend.

```{r plot1_hint, eval = static, results = 'asis', echo = FALSE}
cat('<button type="button" class="btn btn-info btn-sm" data-toggle="collapse" data-target="#plot1">Hint</button>
<div id = "plot1" class="collapse" style="border:1px; border-style:solid; padding: 1em; border-color:#1F78B4">
Use the function plot(..., xlab, ylab, col). Use the function legend(...) to add a legend to the plot.
</div>')
```

```{r, eval = static, echo = FALSE}
asis_output("</div>")

asis_output("#### Solution 1\\n")
```

```{r plot1-solution, solution = TRUE}
plot(heart$age, heart$year)
plot(heart$age, heart$year, xlab = "Age", ylab = "Year of acceptance")
plot(heart$age, heart$year, xlab = "Age", ylab = "Year of acceptance", col = as.numeric(heart$transplant))
legend(-40, 6, c("no", "yes"), col = c("black", "red"), pch = 1)
```



```{r, eval = static, echo = FALSE}
asis_output("#### Task 2\\n")
```

```{r, eval = static, echo = FALSE}
asis_output('<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">')
```

* Create a boxplot of `age` per `status` using the **retinopathy** data set. \
* Change the color to blue and green respectively.

```{r plot2_hint, eval = static, results = 'asis', echo = FALSE}
cat('<button type="button" class="btn btn-info btn-sm" data-toggle="collapse" data-target="#plot2">Hint</button>
<div id = "plot2" class="collapse" style="border:1px; border-style:solid; padding: 1em; border-color:#1F78B4">
Use the function boxplot(...). 
</div>')
```

```{r, eval = static, echo = FALSE}
asis_output("</div>")

asis_output("#### Solution 2\\n")
```

```{r plot2-solution, solution = TRUE}
boxplot(retinopathy$age ~ retinopathy$status)
boxplot(retinopathy$age ~ retinopathy$status, col = c("blue", "green"))
```


```{r, eval = static, echo = FALSE}
asis_output("#### Task 3\\n")
```

```{r, eval = static, echo = FALSE}
asis_output('<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">')
```

Using the **retinopathy** data set: \
* Create a smooth plot or `age` and `risk`. \
* Create a density plot or `age` per `type` group.

```{r plot3_hint, eval = static, results = 'asis', echo = FALSE}
cat('<button type="button" class="btn btn-info btn-sm" data-toggle="collapse" data-target="#plot3">Hint</button>
<div id = "plot3" class="collapse" style="border:1px; border-style:solid; padding: 1em; border-color:#1F78B4">
Use can use the ggplot2 library and the geom_smooth(...) and geom_density(...) functions.
</div>')
```

```{r, eval = static, echo = FALSE}
asis_output("</div>")

asis_output("#### Solution 3\\n")
```

```{r plot3-solution, solution = TRUE}
library(ggplot2)
ggplot(retinopathy, aes(age, risk)) +
geom_smooth(colour='black', span = 0.4)
ggplot(retinopathy, aes(age, fill = type)) +
geom_density(alpha = 0.25) 
```


## Indexing and Subsetting

Sometimes we want to obtain a subset of the data sets before investigating the descriptive statistics and performing the statistical analysis.

### Indexing {.tabset .tabset-fade .tabset-pills}

```{r, eval = static, echo = FALSE}
asis_output("#### Task\\n")
```

```{r, eval = static, echo = FALSE}
asis_output('<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">')
```

Using the **heart** data set: \
* Select the first row. \
* Select the first column. \
* Select the column `surgery`.


```{r, eval = static, echo = FALSE}
asis_output("</div>")

asis_output("#### Solution\\n")
```

```{r ind1-solution, solution = TRUE}
heart[1, ]
heart[, 1]
heart["surgery"]
heart[["surgery"]]
heart[, "surgery"]
```


### Subsetting {.tabset .tabset-fade .tabset-pills}

```{r, eval = static, echo = FALSE}
asis_output("#### Task 1\\n")
```

```{r, eval = static, echo = FALSE}
asis_output('<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">')
```

Using the **retinopathy** data set: \
* Select the `futime` for all `adult` patients. \
* Select all the variables for patients that received treatment. \



```{r, eval = static, echo = FALSE}
asis_output("</div>")

asis_output("#### Solution 1\\n")
```

```{r sub1-solution, solution = TRUE}
retinopathy$futime[retinopathy$type == "adult"]
retinopathy[retinopathy$trt == 1, ]
```


```{r, eval = static, echo = FALSE}
asis_output("#### Task 2\\n")
```

```{r, eval = static, echo = FALSE}
asis_output('<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">')
```

Using the **retinopathy** data set: \
* Select the `age` for patients that have `futime` more than 20. \
* Select the `age` for patients that have `futime` more than 20 and are adults. \
* Select patients that have no missing values in age.


```{r, eval = static, echo = FALSE}
asis_output("</div>")

asis_output("#### Solution 2\\n")
```

```{r sub2-solution, solution = TRUE}
retinopathy$age[retinopathy$futime > 20]
retinopathy$age[retinopathy$futime > 20 & retinopathy$type == "adult"]
retinopathy[!is.na(retinopathy$age), ]
```



```{r, eval = static, echo = FALSE}
asis_output("#### Task 3\\n")
```

```{r, eval = static, echo = FALSE}
asis_output('<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">')
```

Using the **retinopathy** data set: \
* Select patients that have no missing values in age.


```{r, eval = static, echo = FALSE}
asis_output("</div>")

asis_output("#### Solution 3\\n")
```

```{r sub3-solution, solution = TRUE}
retinopathy[!is.na(retinopathy$age), ]
```