---
title: 'Common R Objects'
output: 
  html_document:
    code_folding: show
    df_print: paged
    toc: true
    toc_float:
      collapsed: false
    number_sections: false
    theme: spacelab
    highlight: tango
    includes:
      after_body: ../footerEA.html
    css: ../style.css
---

```{r, echo = FALSE, purl = FALSE}
knitr::knit_hooks$set(purl = knitr::hook_purl)

options(purl = FALSE)

knitr::opts_chunk$set(purl = FALSE)
```

```{r, include = FALSE}
knitr::opts_hooks$set(eval = function(opt) {
  if (any(opt$exercise))
    opt$eval <- opt$include <- FALSE
  
  opt
})


static <- TRUE

options(width = 100)

```


```{r packages, include = FALSE}
library(kableExtra)
library(knitr)
```


```{r load_data, context="data", include=FALSE}
library(survival)
```


## Preface {data-progressive=FALSE}
### R packages
In this practical, a number of R packages are used.
The packages used (with versions that were used to generate the solutions) are:

* `r R.version.string`
* `survival` (version: `r packageVersion("survival")`)


### Dataset {.tabset .tabset-fade .tabset-pills}

For this practical, we will use the **heart** and **retinopathy** data sets from the `survival` package. More details about the data sets can be found in:

https://stat.ethz.ch/R-manual/R-devel/library/survival/html/heart.html

https://stat.ethz.ch/R-manual/R-devel/library/survival/html/retinopathy.html

Always check your data.


```{r, eval = static, echo = FALSE}
asis_output("#### Task\\n")
```

```{r, eval = static, echo = FALSE}
asis_output('<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">')
```

Explore the **heart** and **retinopathy** data sets - print the first six and last six rows. 

```{r data_hint, eval = static, results = 'asis', echo = FALSE}
cat('<button type="button" class="btn btn-info btn-sm" data-toggle="collapse" data-target="#data">Hint</button>
<div id = "data" class="collapse" style="border:1px; border-style:solid; padding: 1em; border-color:#1F78B4">
Use the functions head(...) and tail(...) to investigate the data set. Replace the dots with the name of the data set.
</div>')
```


```{r, eval = static, echo = FALSE}
asis_output("</div>")

asis_output("#### Solution\\n")
```

```{r data-solution, solution = TRUE}
library(survival)
head(heart)
tail(heart)
head(retinopathy)
tail(retinopathy)
```

## Common R Objects

It is important to distinguish the different object in R.

### Vectors {.tabset .tabset-fade .tabset-pills}

Let's investigate some vectors.

```{r, eval = static, echo = FALSE}
asis_output("#### Task 1\\n")
```

```{r, eval = static, echo = FALSE}
asis_output('<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">')
```

View the vectors `event` and `age` from the **heart** data set.

```{r vec_hint, eval = static, results = 'asis', echo = FALSE}
cat('<button type="button" class="btn btn-info btn-sm" data-toggle="collapse" data-target="#vec">Hint</button>
<div id = "vec" class="collapse" style="border:1px; border-style:solid; padding: 1em; border-color:#1F78B4">
Use the dollar sign to select the variables.
</div>')
```


```{r, eval = static, echo = FALSE}
asis_output("</div>")

asis_output("#### Solution 1\\n")
```

```{r vec-solution, solution = TRUE}
heart$event
heart$age
```



```{r, eval = static, echo = FALSE}
asis_output("#### Task 2\\n")
```

```{r, eval = static, echo = FALSE}
asis_output('<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">')
```

View the vectors `eye` and `risk` from the **retinopathy** data set.

```{r vec2_hint, eval = static, results = 'asis', echo = FALSE}
cat('<button type="button" class="btn btn-info btn-sm" data-toggle="collapse" data-target="#vec2">Hint</button>
<div id = "vec2" class="collapse" style="border:1px; border-style:solid; padding: 1em; border-color:#1F78B4">
Use the dollar sign to select the variables.
</div>')
```


```{r, eval = static, echo = FALSE}
asis_output("</div>")

asis_output("#### Solution 2\\n")
```

```{r vec2-solution, solution = TRUE}
retinopathy$eye
retinopathy$risk
```





```{r, eval = static, echo = FALSE}
asis_output("#### Task 3\\n")
```

```{r, eval = static, echo = FALSE}
asis_output('<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">')
```

Create a numerical vector that consists of the values: 34, 24, 19, 23, 16. Give the name `numbers` to this new vector.

```{r vec3_hint, eval = static, results = 'asis', echo = FALSE}
cat('<button type="button" class="btn btn-info btn-sm" data-toggle="collapse" data-target="#vec3">Hint</button>
<div id = "vec3" class="collapse" style="border:1px; border-style:solid; padding: 1em; border-color:#1F78B4">
Use the c(...) function. Replace the dots with the numbers.
</div>')
```


```{r, eval = static, echo = FALSE}
asis_output("</div>")

asis_output("#### Solution 3\\n")
```

```{r vec3-solution, solution = TRUE}
numbers <- c(34, 24, 19, 23, 16)
```






```{r, eval = static, echo = FALSE}
asis_output("#### Task 4\\n")
```

```{r, eval = static, echo = FALSE}
asis_output('<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">')
```

Create a numerical vector that takes the integer values from 1 until 200. Give the name `numbers_2` to this new vector.

```{r vec4_hint, eval = static, results = 'asis', echo = FALSE}
cat('<button type="button" class="btn btn-info btn-sm" data-toggle="collapse" data-target="#vec4">Hint</button>
<div id = "vec4" class="collapse" style="border:1px; border-style:solid; padding: 1em; border-color:#1F78B4">
Use the c(...) function. Replace the dots with the numbers.
</div>')
```


```{r, eval = static, echo = FALSE}
asis_output("</div>")

asis_output("#### Solution 4\\n")
```

```{r vec4-solution, solution = TRUE}
numbers_2 <- c(1:200)
```






```{r, eval = static, echo = FALSE}
asis_output("#### Task 5\\n")
```

```{r, eval = static, echo = FALSE}
asis_output('<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">')
```

Create a categorical vector that consists of the values: yes, yes, no, no, no, yes. Give the name `treatment` to this new vector.

```{r vec5_hint, eval = static, results = 'asis', echo = FALSE}
cat('<button type="button" class="btn btn-info btn-sm" data-toggle="collapse" data-target="#vec5">Hint</button>
<div id = "vec5" class="collapse" style="border:1px; border-style:solid; padding: 1em; border-color:#1F78B4">
Use the c(...) function. Replace the dots with the categories.
</div>')
```


```{r, eval = static, echo = FALSE}
asis_output("</div>")

asis_output("#### Solution 5\\n")
```

```{r vec5-solution, solution = TRUE}
treatment <- c("yes", "yes", "no", "no", "no", "yes")
```



### Matrices and Data Frames {.tabset .tabset-fade .tabset-pills}

Let's investigate some matrices and data frames.

```{r, eval = static, echo = FALSE}
asis_output("#### Task 1\\n")
```

```{r, eval = static, echo = FALSE}
asis_output('<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">')
```

Create a matrix using the vectors `id` and `age` from the **heart** data set. This matrix should have 2 columns where each column represents each variable.

```{r mat_hint, eval = static, results = 'asis', echo = FALSE}
cat('<button type="button" class="btn btn-info btn-sm" data-toggle="collapse" data-target="#mat">Hint</button>
<div id = "mat" class="collapse" style="border:1px; border-style:solid; padding: 1em; border-color:#1F78B4">
Use the function matrix(...).
</div>')
```


```{r, eval = static, echo = FALSE}
asis_output("</div>")

asis_output("#### Solution 1\\n")
```

```{r mat-solution-1, solution = TRUE}
matrix(c(heart$id, heart$age), , 2) 
```



```{r, eval = static, echo = FALSE}
asis_output("#### Task 2\\n")
```

```{r, eval = static, echo = FALSE}
asis_output('<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">')
```

Create a data frame using the vectors `id`, `type` and `trt` from the **retinopathy** data set. This data frame should have 3 columns, where each column represents each variable.

```{r df_hint, eval = static, results = 'asis', echo = FALSE}
cat('<button type="button" class="btn btn-info btn-sm" data-toggle="collapse" data-target="#df">Hint</button>
<div id = "df" class="collapse" style="border:1px; border-style:solid; padding: 1em; border-color:#1F78B4">
Use the function data.frame(...).
</div>')
```


```{r, eval = static, echo = FALSE}
asis_output("</div>")

asis_output("#### Solution 2\\n")
```

```{r dfc-solution, solution = TRUE}
data.frame(id = retinopathy$id, type = retinopathy$type, trt = retinopathy$trt) 
```



### Arrays {.tabset .tabset-fade .tabset-pills}

Let's investigate some arrays.

```{r, eval = static, echo = FALSE}
asis_output("#### Task 1\\n")
```

```{r, eval = static, echo = FALSE}
asis_output('<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">')
```

Create an array that consists of 2 matrices. Matrix 1 will consist of the values 1:4 and matrix 2 will consist of the values 5:8. Both matrices will have 2 columns and 2 rows.

```{r ar1_hint, eval = static, results = 'asis', echo = FALSE}
cat('<button type="button" class="btn btn-info btn-sm" data-toggle="collapse" data-target="#ar1">Hint</button>
<div id = "ar1" class="collapse" style="border:1px; border-style:solid; padding: 1em; border-color:#1F78B4">
Use the function array(...).
</div>')
```


```{r, eval = static, echo = FALSE}
asis_output("</div>")

asis_output("#### Solution 1\\n")
```

```{r ar-solution-1, solution = TRUE}
array(data = 1:8, dim = c(2, 2, 2))
```



```{r, eval = static, echo = FALSE}
asis_output("#### Task 2\\n")
```

```{r, eval = static, echo = FALSE}
asis_output('<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">')
```

Give the name `ar1` to the previous array. Furthermore, investigate the argument dimnames and change the names of the rows, columns and matrices.

```{r ar2_hint, eval = static, results = 'asis', echo = FALSE}
cat('<button type="button" class="btn btn-info btn-sm" data-toggle="collapse" data-target="#ar2">Hint</button>
<div id = "ar2" class="collapse" style="border:1px; border-style:solid; padding: 1em; border-color:#1F78B4">
Use the function array(...). Check the help page for the dimnames argument. Note that this must be in a list format.
</div>')
```


```{r, eval = static, echo = FALSE}
asis_output("</div>")

asis_output("#### Solution 2\\n")
```

```{r ar-solution-2, solution = TRUE}
ar1 <- array(data = 1:8, dim = c(2, 2, 2), 
             dimnames = list(c("Row1", "Row2"), c("Col1", "Col2"), c("Mat1", "Mat2")))
```


### Lists {.tabset .tabset-fade .tabset-pills}

Let's investigate some lists.

```{r, eval = static, echo = FALSE}
asis_output("#### Task 1\\n")
```

```{r, eval = static, echo = FALSE}
asis_output('<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">')
```

Create a list using the vectors `stop` from the **heart** data set and `id`, `risk` from the **retinopathy** data set. Give the names `stop_heart`, `id_reti` and `risk_reti`.

```{r list_hint, eval = static, results = 'asis', echo = FALSE}
cat('<button type="button" class="btn btn-info btn-sm" data-toggle="collapse" data-target="#list">Hint</button>
<div id = "list" class="collapse" style="border:1px; border-style:solid; padding: 1em; border-color:#1F78B4">
Use the function list(...).
</div>')
```


```{r, eval = static, echo = FALSE}
asis_output("</div>")

asis_output("#### Solution 1\\n")
```

```{r list-solution, solution = TRUE}
list(stop_heart = heart$stop, id_reti = retinopathy$id, risk_reti = retinopathy$risk)
```


```{r, eval = static, echo = FALSE}
asis_output("#### Task 2\\n")
```

```{r, eval = static, echo = FALSE}
asis_output('<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">')
```

Create a list using the vectors `numbers`, `numbers_2` and `treatment`. These variables can be taken from the exercise called `Vectors`.  Give the names: `numbers`, `many_numbers` and `treatment`.

```{r list2_hint, eval = static, results = 'asis', echo = FALSE}
cat('<button type="button" class="btn btn-info btn-sm" data-toggle="collapse" data-target="#list2">Hint</button>
<div id = "list2" class="collapse" style="border:1px; border-style:solid; padding: 1em; border-color:#1F78B4">
Use the function list(...).
</div>')
```


```{r, eval = static, echo = FALSE}
asis_output("</div>")

asis_output("#### Solution 2\\n")
```

```{r list2-solution, solution = TRUE}
list(numbers = numbers, many_numbers = numbers_2, treatment = treatment)
```


