---
title: 'Importing and Saving Data'
output: 
  html_document:
    code_folding: show
    df_print: paged
    toc: true
    toc_float:
      collapsed: false
    number_sections: false
    theme: spacelab
    highlight: tango
    includes:
      after_body: ../footerEA.html
    css: ../style.css
---

```{r, echo = FALSE, purl = FALSE}
knitr::knit_hooks$set(purl = knitr::hook_purl)

options(purl = FALSE)

knitr::opts_chunk$set(purl = FALSE)
```

```{r, include = FALSE}
knitr::opts_hooks$set(eval = function(opt) {
  if (any(opt$exercise))
    opt$eval <- opt$include <- FALSE
  
  opt
})


static <- TRUE

options(width = 100)

```


```{r packages, include = FALSE}
library(kableExtra)
library(knitr)
```


```{r load_data, context="data", include=FALSE}
library(survival)
```


## Preface {data-progressive=FALSE}
### R packages
In this practical, a number of R packages are used.
The packages used (with versions that were used to generate the solutions) are:

* `r R.version.string`
* `survival` (version: `r packageVersion("survival")`)


### Dataset {.tabset .tabset-fade .tabset-pills}

For this practical, we will use the **heart** and **retinopathy** data sets from the `survival` package. More details about the data sets can be found in:

https://stat.ethz.ch/R-manual/R-devel/library/survival/html/heart.html

https://stat.ethz.ch/R-manual/R-devel/library/survival/html/retinopathy.html



## Importing and Saving Data
### Save your work {.tabset .tabset-fade .tabset-pills}

It is important to save your work.

```{r, eval = static, echo = FALSE}
asis_output("#### Task 1\\n")
```

```{r, eval = static, echo = FALSE}
asis_output('<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">')
```

Save the vectors `numbers <- c(34, 24, 19, 23, 16)`, `numbers_2 <- c(1:200)` and `treatment <- c("yes", "yes", "no", "no", "no", "yes")`. Use the name `new_vectors`.

```{r save_hint, eval = static, results = 'asis', echo = FALSE}
cat('<button type="button" class="btn btn-info btn-sm" data-toggle="collapse" data-target="#save">Hint</button>
<div id = "save" class="collapse" style="border:1px; border-style:solid; padding: 1em; border-color:#1F78B4">
Use the function save(...). Note that you need to set the working directory.
</div>')
```


```{r, eval = static, echo = FALSE}
asis_output("</div>")

asis_output("#### Solution 1\\n")
```

```{r save-solution, solution = TRUE}
numbers <- c(34, 24, 19, 23, 16)
numbers_2 <- c(1:200)
treatment <- c("yes", "yes", "no", "no", "no", "yes")
save(numbers, numbers_2, treatment, file = "new_vectors.RData")
```





```{r, eval = static, echo = FALSE}
asis_output("#### Task 2\\n")
```

```{r, eval = static, echo = FALSE}
asis_output('<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">')
```

Save the vectors `events <- heart$event` and `eyes <- retinopathy$eye`. Use the name `vectors_survival`. 

```{r save2_hint, eval = static, results = 'asis', echo = FALSE}
cat('<button type="button" class="btn btn-info btn-sm" data-toggle="collapse" data-target="#save2">Hint</button>
<div id = "save2 class="collapse" style="border:1px; border-style:solid; padding: 1em; border-color:#1F78B4">
Use the function save(...). Note that you need to set the working directory.
</div>')
```


```{r, eval = static, echo = FALSE}
asis_output("</div>")

asis_output("#### Solution 2\\n")
```

```{r save2-solution, solution = TRUE}
events <- heart$event
eyes <- retinopathy$eye
save(events, eyes, file = "vectors_survival.RData")
```


### Load your work {.tabset .tabset-fade .tabset-pills}

Let's continue working on the data sets by loading our results.

```{r, eval = static, echo = FALSE}
asis_output("#### Task\\n")
```

```{r, eval = static, echo = FALSE}
asis_output('<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">')
```

Load the file `new_vectors`.

```{r load_hint, eval = static, results = 'asis', echo = FALSE}
cat('<button type="button" class="btn btn-info btn-sm" data-toggle="collapse" data-target="#load">Hint</button>
<div id = "load" class="collapse" style="border:1px; border-style:solid; padding: 1em; border-color:#1F78B4">
Use the function load(...).
</div>')
```


```{r, eval = static, echo = FALSE}
asis_output("</div>")

asis_output("#### Solution\\n")
```

```{r load-solution, solution = TRUE}
load("new_vectors.RData")
```


### Remove your work {.tabset .tabset-fade .tabset-pills}

Remove unnecessary objects.

```{r, eval = static, echo = FALSE}
asis_output("#### Task 1\\n")
```

```{r, eval = static, echo = FALSE}
asis_output('<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">')
```

Remove the vectors `numbers`, `numbers_2` and `treatment`.

```{r rm_hint, eval = static, results = 'asis', echo = FALSE}
cat('<button type="button" class="btn btn-info btn-sm" data-toggle="collapse" data-target="#rm">Hint</button>
<div id = "rm" class="collapse" style="border:1px; border-style:solid; padding: 1em; border-color:#1F78B4">
Use the function rm(...).
</div>')
```


```{r, eval = static, echo = FALSE}
asis_output("</div>")

asis_output("#### Solution 1\\n")
```

```{r rm-solution, solution = TRUE}
rm(numbers, numbers_2, treatment)
```



```{r, eval = static, echo = FALSE}
asis_output("#### Task 2\\n")
```

```{r, eval = static, echo = FALSE}
asis_output('<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">')
```

Remove the vectors `events` and `eyes`.

```{r rm2_hint, eval = static, results = 'asis', echo = FALSE}
cat('<button type="button" class="btn btn-info btn-sm" data-toggle="collapse" data-target="#rm2">Hint</button>
<div id = "rm2" class="collapse" style="border:1px; border-style:solid; padding: 1em; border-color:#1F78B4">
Use the function rm(...).
</div>')
```


```{r, eval = static, echo = FALSE}
asis_output("</div>")

asis_output("#### Solution 2\\n")
```

```{r rm2-solution, solution = TRUE}
rm(events, eyes)
```

