---
title: 'Indexing and Subsetting'
output: 
  html_document:
    code_folding: show
    df_print: paged
    toc: true
    toc_float:
      collapsed: false
    number_sections: false
    theme: spacelab
    highlight: tango
    includes:
      after_body: ../footerEA.html
    css: ../style.css
---

```{r, echo = FALSE, purl = FALSE}
knitr::knit_hooks$set(purl = knitr::hook_purl)

options(purl = FALSE)

knitr::opts_chunk$set(purl = FALSE)
```

```{r, include = FALSE}
knitr::opts_hooks$set(eval = function(opt) {
  if (any(opt$exercise))
    opt$eval <- opt$include <- FALSE
  
  opt
})


static <- TRUE

options(width = 100)

```


```{r packages, include = FALSE}
library(kableExtra)
library(knitr)
```


```{r load_data, context="data", include=FALSE}
library(survival)
```


## Preface {data-progressive=FALSE}
### R packages
In this practical, a number of R packages are used.
The packages used (with versions that were used to generate the solutions) are:

* `r R.version.string`
* `survival` (version: `r packageVersion("survival")`)


### Dataset {.tabset .tabset-fade .tabset-pills}

For this practical, we will use the **heart** and **retinopathy** data sets from the `survival` package. More details about the data sets can be found in:

https://stat.ethz.ch/R-manual/R-devel/library/survival/html/heart.html

https://stat.ethz.ch/R-manual/R-devel/library/survival/html/retinopathy.html


## Indexing and Subsetting

Sometimes we want to obtain a subset of the data sets before investigating the descriptive statistics and performing the statistical analysis.

### Indexing {.tabset .tabset-fade .tabset-pills}

```{r, eval = static, echo = FALSE}
asis_output("#### Task\\n")
```

```{r, eval = static, echo = FALSE}
asis_output('<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">')
```

Using the **heart** data set: \

* Select the first row. \
* Select the first column. \
* Select the column `surgery`.


```{r, eval = static, echo = FALSE}
asis_output("</div>")

asis_output("#### Solution\\n")
```

```{r ind1-solution, solution = TRUE}
heart[1, ]
heart[, 1]
heart["surgery"]
heart[["surgery"]]
heart[, "surgery"]
```


### Subsetting {.tabset .tabset-fade .tabset-pills}

```{r, eval = static, echo = FALSE}
asis_output("#### Task 1\\n")
```

```{r, eval = static, echo = FALSE}
asis_output('<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">')
```

Using the **retinopathy** data set: \

* Select the `futime` for all `adult` patients. \
* Select all the variables for patients that received treatment. \



```{r, eval = static, echo = FALSE}
asis_output("</div>")

asis_output("#### Solution 1\\n")
```

```{r sub1-solution, solution = TRUE}
retinopathy$futime[retinopathy$type == "adult"]
retinopathy[retinopathy$trt == 1, ]
```


```{r, eval = static, echo = FALSE}
asis_output("#### Task 2\\n")
```

```{r, eval = static, echo = FALSE}
asis_output('<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">')
```

Using the **retinopathy** data set: \

* Select the `age` for patients that have `futime` more than 20. \
* Select the `age` for patients that have `futime` more than 20 and are adults. \
* Select patients that have no missing values in age.


```{r, eval = static, echo = FALSE}
asis_output("</div>")

asis_output("#### Solution 2\\n")
```

```{r sub2-solution, solution = TRUE}
retinopathy$age[retinopathy$futime > 20]
retinopathy$age[retinopathy$futime > 20 & retinopathy$type == "adult"]
retinopathy[!is.na(retinopathy$age), ]
```

